---
title: "3. Continuous tests for 1 or 2 populations"
author: "jsg"
date: "9/8/2020"
output: github_document
---
Before doing this, review the **Week 3** lecture set slides from 
https://sites.google.com/view/biostats/bio-7800178002/week-3 and
the  **Week_3.R** script in the grad_files_2020 folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats).
You’ll also use the qqplot_example.R and
central_limit_theorem.R script from the 
[code_examples folder](https://github.com/jsgosnell/CUNY-BioStats/tree/master/code_examples).

Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  * **push** updates to github

## Recognizing and assessing normality 

1. Using the qqplot_example.R code, examine the following distributions and, for 
the continuous distributions (marked with a “*”), observe how a normal probability
plot (qqplot) can be used to visually test for approximate normality.

* *Normal (u= 0; σ^2^=1, 10, 100) 
* *Student's t (df = 1, 10, 30, & 100) 
* *Chi-square (df= 1, 2, 5, 30, 50)
* Bernoulli (P=0.1, 0.5, & 0.9) 
* Binomial (P=0.05; N= 2, 5, 25, & 50); (P=0.25; N= 2, 5, 25, & 50); (P=0.50; 
N= 2, 5, 25, & 50); (P=0.75; N= 2, 5, 25, & 50); (P=0.95; N= 2, 5, 25, & 50) 
* Poisson ( u= 2, 5, 10, 30, & 50)

For this question, its easiest to just source the main file and see what happens.
When you source a script, it is run in R without showing any console output
(but graphs and objects are still produced!).  Try *source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/qqplot_example.R*

```{r}
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/qqplot_example.R")
```

*Notice the spread of DATA of every distribution tend towards normality as sample size increases*

2. Review the central_limit_theorem.R code if you need to convince/remind yourself how common normality of means is for even non-normal data. You can source the code
using the same approach noted in Question 1.

*Here we are focused on how the means look as sample size increases*

```{r}
#make sure you have VGAM library installed
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/central_limit_theorem.R")
```


## Just for practice (note some sample sizes may be too small for these to all be good ideas!)

3. Seven observers were shown, for a brief period, a grill with 161 flies impaled and were asked to estimate the number. The results are given by Cochran (1954). Based on five estimates, they were 183.2, 149.0, 154.0, 167.2, 187.2, 158.0, and 143.0. Test the null hypothesis that the mean of the estimates is 161 flies.
* Assuming variance = 275

```{r}
flies <- c(183.2, 149.0, 154.0, 167.2, 187.2, 158.0, 143.0)
library(BSDA)
z.test(x=flies, mu = 161, sigma.x=sqrt(275))
```
*Using a z-test, I found a test statistics of z~=0.33 .This corresponds to a p-value of 0.73. This p value is >.05, so I fail to reject 
the null hypothesis that the mean of the estimates is 161 flies.*

*  Estimating the variance from the data
```{r}
t.test(x=flies,mu = 161)
```
*Using a t-test, which is appropriate when the variance must be estimated
from the sample and the means of the data may be assumed to follow a normal 
distribution,  I found a test statistics of t~6~=0.32.  This corresponds to a 
p-value of 0.76. This p-value is >.05, so I fail to reject 
the null hypothesis that the mean of the estimates is 161 flies.*

* Using rank transform analysis 

```{r}
wilcox.test(flies, mu=161)
```
*Using a Wilcoxon signed rank test, which is appropriate when normality assumptions 
can't be met and the distribution of the data appears to be symmetric, I found a test
statistics of V = 15 .This corresponds to a p-value of 0.94. This p-value is >.05, so I 
fail to reject the null hypothesis that the mean of the estimates is 161 flies.*

* Using binary transform analysis
```{r}
SIGN.test(flies, md=161)
```

*Using a sign test, which is appropriate when the data is continuous and other 
assumptions can't be met, I found a test statistics of s = 3 .This corresponds to a p-value of 1. This p-value is >.05, so I fail to reject 
the null hypothesis that the median (**Note change here**) of the estimates is
161 flies.*

4. Yields of 10 strawberry plants in a uniformity trial are given by Baker and
Baker (1953) as 239, 176, 235, 217, 234, 216, 318, 190, 181, and 225 g. Test the
hypothesis that µ = 205

* Assuming variance = 1500
```{r}
strawberries <- c(239, 176, 235, 217, 234, 216, 318, 190, 181, 225)
z.test(x=strawberries,mu = 205, sigma.x=sqrt(1500))
```
*Using a z-test, I found a test statistics of z=1.48.  This corresponds to a 
p-value of 0.14. This p-value is >.05, so I fail to reject 
the null hypothesis that the population mean is equal to 205.*

*  Estimating the variance from the data

```{r}
t.test(x=strawberries,mu = 205)
```
*Using a t-test, which is appropriate when the variance must be estimated
from the sample and the means of the data may be assumed to follow a normal distribution, I found a test statistics of t~9~=1.42.  This corresponds to a 
p-value of 0.19. This p-value is >.05, so I fail to reject 
the null hypothesis that the population mean is equal to 205.*

* Using rank transform analysis 
```{r}
wilcox.test(strawberries, mu=205)
```

*Using a Wilcoxon signed rank test, which is appropriate when normality assumptions 
can't be met and the distribution of the data appears to be symmetric, I found a test statistics of V=40.5.  This corresponds to a 
p-value of 0.20.  This p-value is >.05, so I fail to reject 
the null hypothesis that the population mean is equal to 205.*

* Using binary transform analysis
```{r}
SIGN.test(strawberries, md=205)
```

*Using a sign test, which is appropriate when the data is continuous and other 
assumptions can't be met,  I found a test statistics of s= 7.  This corresponds 
to a p-value of 0.34. This p-value is >.05,so I fail to reject 
the null hypothesis that the population median (**Note change here**) is equal to 205.*

5. Studies of flying snakes have led researchers to posit the mean undulation rate is 1.4 Hz.  You wish to test this hypothesis using the small sample of undulation rates shown below.  Create a small dataset of the paradise tree snake undulation rates and choose and justify a test you can use to assess the data.  
Undulation rates (in Hz):  0.9, 1.4, 1.2, 1.2, 1.3, 2.0, 1.4, 1.6

```{r}
snakes <- c(0.9, 1.4, 1.2, 1.2, 1.3, 2.0, 1.4, 1.6)
t.test(snakes, mu=1.4)
```
*Using a t-test, which is appropriate when the variance must be estimated
from the sample and the means of the data may be assumed to follow a normal
distribution,  I found a test statistics of t~7~=-.22.  This corresponds to a 
p-value of 0.83. This p-value is >.05, so I fail to reject 
the null hypothesis that the mean undulation rate is 1.4 Hz.*

6.  Using data from Australian athletes (http://www.statsci.org/data/oz/ais.html 
for details), determine if the average male training at the Australian Institute 
of Sport differs in weight from the average Australian male (85.9 kg) using
bootstrapping techniques. Data at 
```{r}
sport <- read.table("http://www.statsci.org/data/oz/ais.txt", header = T)
```

Remember you can source the bootstrapjsg function using

```{r}
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/bootstrapjsg.R")
```

*Answer*

```{r}
sport <- read.table("http://www.statsci.org/data/oz/ais.txt", header = T)
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/bootstrapjsg.R")
bootstrapjsg(data1=sport[sport$Sex == "male", "Wt"], null=85.9)
#to get estimates!
summary(sport[sport$Sex == "male",])
hist(sport[sport$Sex == "male", "Wt"])
```

*Using a bootstrap approach with 10,000 samples (default in the bootstrapjsg function), I found a p-value of 0.0069; this is <.05, so I reject the null 
hypothesis that the 
mean weight of male athletes at the AIS is equal to 85.9 kg (the average weight of
an Australian male). Estimates from the da suggest the mean weight of males at 
the AIS is 82.52 kg, which is less than 85.9 kg.*

7. The following data are human blood clotting times (in minutes) of individuals
given one of two different drugs. 

|Drug B | Drug G
|------ |-------
|8.8 | 9.9
|8.4 | 9.0
|7.9 | 11.1
|8.7 | 9.6
|9.1 | 8.7
|9.6 | 10.4 
|    | 9.5

Test the hypothesis that the mean clotting times
are equal for the two groups
*  Estimating the variance from the data 

```{r}
drug_b <- c( 8.8, 8.4, 7.9, 8.7, 9.1, 9.6)
drug_g <- c(9.9, 9.0, 11.1, 9.6, 8.7, 10.4, 9.5)
t.test(drug_b, drug_g)
```
*Using a un-paired t-test, since the experimental units were not matched and I 
assumed the means of each group would follow a normal distribution of unknown 
variance,  I found a test statistics of t~10.701~=-2.544.  This corresponds to a 
p-value of 0.02. This p-value is <.05, so I reject the null hypothesis that 
the mean clotting times are the same for the two drugs.*

* Using rank transform analysis 
```{r}
wilcox.test(drug_b, drug_g)
```
 
*Using a un-paired rank-based test, which is appropriate when normality assumptions 
can't be met and I assumed the means of each group would follow a similar distribution, 
I found a test statistics of W=7.  This corresponds to a 
p-value of 0.05.  This p-value is >.05, so I fail to reject the null hypothesis that 
the mean clotting times are the same for the two drugs.*

* Using a permutation test

```{r}
require(coin) #requires data_frame
clotting <- data.frame(drug = c(rep("drug_b", length(drug_b)), rep("drug_g", 
                                                                   length(drug_g))),
                       clotting = c(drug_b, drug_g))
clotting$drug <- factor(clotting$drug)
independence_test(clotting ~ drug, clotting)
```
*Using a permutation test, which is not fully appropriate here due to small sample 
sizes (and that also assumes similar distributions for each group), 
I found a test statistics of Z=-2.0726..  This corresponds to a 
p-value of 0.038.  This p-value is >.05, so I fail to reject the null hypothesis that 
the mean clotting times are the same for the two drugs.*

* Using a bootstrap test 

```{r}
bootstrapjsg(drug_b, drug_g)
```
*Using a bootstrap test with 10000 samples, which is not fully appropriate here 
due to small sample sizes, 
I found a p value of 0.0047.  This p-value is <.05, so I reject the null 
hypothesis that 
the mean clotting times are the same for the two drugs.*

8. Data on plant heights (in cm) for plants grown with a new and old formulation 
of fertilizer can be found at
https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/fertilizer.csv . 
Use the data to test the hypothesis that there is no difference in mean plant
heights for the two groups 

*  Estimating the variance from the data 
```{r}
fertilizer <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/fertilizer.csv")
t.test(height ~ fertilizer, fertilizer)
```
*Using a un-paired t-test, since the experimental units were not matched and I 
assumed the means of each group would follow a normal distribution of unknown 
variance,  I found a test statistics of t~15.559~=3.01.  This corresponds to a 
p-value of 0.008. This p-value is <.05, so I  to reject the null hypothesis that 
the impact on plant growth doesn not differ between the two fertilizer formulations.*

* Using rank transform analysis 
```{r}
wilcox.test(height ~ fertilizer, fertilizer)
```
*Using a un-paired rank-based test, which is appropriate when normality assumptions 
can't be met and I assumed the means of each group would follow a similar distribution, 
I found a test statistics of W=68.  This corresponds to a 
p-value of 0.012.  This p-value is <.05, so I  to reject the null hypothesis that 
the impact on plant growth doesn not differ between the two fertilizer formulations.*

* Using a permutation test 
```{r}
independence_test(height ~ factor(fertilizer), fertilizer)
```
*Using a permutation test, which is appropriate given the sample sizes and fact 
I assume similar distributions for each group, 
I found a test statistics of Z=2.47.  his corresponds to a 
p-value of 0.01.  This p-value is <.05, so I  to reject the null hypothesis that 
the impact on plant growth doesn not differ between the two fertilizer formulations.*

* Using a bootstrap test 

*This is where you may need the different (wide) format of data! As always, multiple
options exist in R*

```{r}
require(reshape2)
fertilizer$ID <- 1:nrow(fertilizer)
fertilizer_wide <- dcast(fertilizer, ID~fertilizer, value.var = "height")
#subsetting the long-form data
bootstrapjsg(fertilizer[fertilizer$fertilizer =="old", "height"],
              fertilizer[fertilizer$fertilizer =="new", "height"])
#is same as using the wide data
bootstrapjsg(fertilizer_wide$new, fertilizer_wide$old)
```
*Using a bootstrap test with 10000 samples, 
I found a p value of 0.013.  This p-value is <.05, so I  to reject the null hypothesis that 
the impact on plant growth doesn not differ between the two fertilizer formulations.*

Note you may need to cast this dataframe into a dataframe with a column for 
old and new outcomes.  You’ll have to add a unique identifier. Try something like

```{r, eval=FALSE}
fertilizer$ID <- 1:nrow(fertilizer)
fertilizer_wide <- dcast(fertilizer, ID~fertilizer, value.var = "height")
```


9. Metabolic rates were measured in animals before and after administering a drug. Data is
available at 

https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/metabolic_rates.csv

Test the hypothesis that the drug has no effect on metabolic rate with using 

*  a Student's t test
```{r}
metabolic <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/metabolic_rates.csv")
t.test(metabolic$before, metabolic$after, paired = T)
```
*I used a paired t-test because the same animals were measured before and after 
the drug was administered.  I also assumed the difference in means was normally
distributed given the trait and sample size.  The test resulted in a statistic of
t~8~ = -3.39. This corresponds to a p-value of <.001.  Since the p-value is <.05,
I reject the null hypothesis that the drug has no effect on metabolic rate.*

* a rank test
```{r}
wilcox.test(metabolic$before, metabolic$after, paired = T)
```
*I used a paired rank-based test because the same animals were measured before and after 
the drug was administered.  I did not assume the difference in means was normally
distributed but did assume it followed a symmetic distribution.  The test resulted in a statistic of
V = 3.   This corresponds to a p-value of .02.  Since the p-value is <.05,
I reject the null hypothesis that the drug has no effect on metabolic rate.*

* a binary test
```{r}
SIGN.test(metabolic$before, metabolic$after)
```

*I used a sign test (always paired!) because the same animals were measured before and after 
the drug was administered.  I did not assume the difference in means was normally
distributed or that the differences followed a symmetic distribution.  The test resulted in a statistic of
s = 2.   This corresponds to a p-value of .18.  Since the p-value is >.05,
I fail to reject the null hypothesis that the drug has no effect on metabolic rate.*

* bootstrapping
```{r}
bootstrapjsg(metabolic$before - metabolic$after)
```
*Since the same animals were measured before and after the drug was administered,
I used a bootstrap (10,000 samples) focused on the difference in rates.  This 
resulted in a p-value of <.001. Since the p-value is <.05,
I reject the null hypothesis that the drug has no effect on metabolic rate.*

10. Endurance of the above animals was also measured. Data is available at

https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/endurance.csv

Test the hypothesis that the drug has no effect on using 
*  a Student's t test
```{r}
endurance <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/endurance.csv")
head(endurance)
t.test(endurance ~ time, endurance, paired = T)
#or
t.test(endurance[endurance$time == "before", "endurance"],
       endurance[endurance$time == "after", "endurance"],
       paired = T)
```
*I used a paired t-test because the same animals were measured before and after 
the drug was administered.  I also assumed the difference in means was normally
distributed given the trait and sample size.  The test resulted in a statistic of
t~8~ = -3.39. This corresponds to a p-value of .29.  Since the p-value is >.05,
I fail to reject the null hypothesis that the drug has no effect on metabolic rate.*
* a rank test
```{r}
wilcox.test(endurance ~ time, endurance, paired = T)
```
*I used a paired rank-based test because the same animals were measured before and after 
the drug was administered.  I did not assume the difference in means was normally
distributed but did assume it followed a symmetic distribution.  The test resulted in a statistic of
V = 13.   This corresponds to a p-value of .30.  Since the p-value is >.05,
I fail to reject the null hypothesis that the drug has no effect on metabolic rate.*

* a binary test
```{r}
SIGN.test(endurance[endurance$time == "before", "endurance"],
          endurance[endurance$time == "after", "endurance"])
```
*I used a sign test (always paired!) because the same animals were measured before and after 
the drug was administered.  I did not assume the difference in means was normally
distributed or that the differences followed a symmetric distribution.  The test resulted in a statistic of
s = 7.   This corresponds to a p-value of .17.  Since the p-value is >.05,
I fail to reject the null hypothesis that the drug has no effect on metabolic rate.*

* bootstrapping
```{r}
bootstrapjsg(endurance[endurance$time == "before", "endurance"]-
             endurance[endurance$time == "after", "endurance"])
```
*Since the same animals were measured before and after the drug was administered,
I used a bootstrap (10,000 samples) focused on the difference in rates.  This 
resulted in a p-value of 0.23. Since the p-value is >.05,
I fail to reject the null hypothesis that the drug has no effect on metabolic rate.*

## For the following problems, make sure you can justify the test you choose to use and carry it out

11. Using data from Australian athletes (http://www.statsci.org/data/oz/ais.html 
for details), determine if the average male training at the Australian Institute 
of Sport differs in from the average female training at the site in mercury
concentration (Hg). 

```{r}
t.test(Hg ~ Sex, sport)
```
*I used a t-test (not paired) since mercury concentration can be assumed to follow
a normal distribution at this sample size.  This resulted in a test statistic of 
t~199.99~ = -15.25, with a corresponding p-value of <.001. Since this p-value is 
<.05, I reject the null hypothesis that there is no difference in average mercury 
concentration between males and females.  If I wanted to graph this, I might use*

```{r}
library(ggplot2)
ggplot(sport, aes_string("Hg"))+
  geom_histogram() +
  facet_wrap(~Sex) +
  ggtitle("	Plasma ferritin concentration \n of Australian athletes") +
  xlab("Ferritin concentration")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32),
        strip.text.x = element_text(size = 22))
```

*which suggests males tend to have a higher Hg concentration.*


12. Laureysens et al. (2004) measured metal content in the wood of 13 poplar 
clones growing in a polluted area, once in August and once in November.
Concentrations of aluminum (in micrograms of Al per gram of wood) are shown 
below. (From Handbook of Biological Statistics). Has metal content changed over
time?

| Clone | August | November
|------| ---|---
|Columbia River | 18.3 |12.7
|Fritiz Pauley | 13.3 |11.1
|Hazendas | 16.5 |15.3
|Primo | 12.6 | 12.7
|Raspalje | 9.5 |10.5
|Hoogvorst |13.6 |15.6
|Balsam Spire | 8.1 |11.2
|Gibecq   |8.9 |14.2
|Beaupre | 10.0 |16.3
|Unal |8.3 |15.5
|Trichobel |7.9 | 19.9
|Gaver | 8.1 | 20.4
|Wolterson | 13.4 |36.8

```{r}
poplar <- read.table("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/poplar.csv",
                   stringsAsFactors = T, header = T)
poplar$difference <- poplar$November - poplar$August

#plot to consider distribution
ggplot(poplar, aes_string("difference"))+
  geom_histogram() +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32),
        strip.text.x = element_text(size = 22))
#lots of holes, but fairly symmetric
wilcox.test(poplar$August, poplar$November, paired=TRUE)
#also could do
bootstrapjsg(poplar$difference)
```

*I plotted the paired data to consider its distribution.  Given its shape and a 
lack of knowledge regarding metal concentration distribution, I used a rank-based
test for paired data. This led to a statistic of V = 17 and p-value of 0.04. Since
this is < .05, I reject the null hypothesis that average metal concentration did not differ
between the two measurements.  Similar results were reached used a bootstrapping 
approach.*

13. Data was collected on cholesterol levels following heart attacks.  Data
available at 
http://www.statsci.org/data/general/cholestg.txt 

Note to read a .txt file you use,
read.table("http://www.statsci.org/data/general/cholestg.txt", header = T, 
stringsAsFactors = T)
Check the defaults to understand the argument.  
More info on the data is @
http://www.statsci.org/data/general/cholest.html
Is there evidence cholesterol levels changed between day 2 and day 14?  
**THINK ABOUT RELABELLING DAY AS A FACTOR (WHY?)?**

```{r}
cholesterol <- read.table("http://www.statsci.org/data/general/cholestg.txt", header = T)
cholesterol$day <- as.factor(cholesterol$day)
head(cholesterol)

summary(cholesterol)

ggplot(cholesterol[cholesterol$day %in% c("2", "14"),], aes_string("cholest")) +
  geom_histogram() +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))+
  facet_wrap(~day)

t.test(cholesterol[cholesterol$day == "2", "cholest"], 
       cholesterol[cholesterol$day == "14", "cholest"],
       paired = T)
```

*The data are paired since the same patient was measured multiple times. For this
problem I changed day to a factor to consider two groups (we'll return to this in class)
using a t-test, since the plots suggested a distribution that should approach 
normality at these sample sizes. I found a test statistic of t~18~ = 3.28 and a 
p-value of <.01. Since the p-value is < .05, I reject the null hypothesis that 
average
cholesterol did not differ between the two sampling dates.
## Using ggplot2
Let’s return to the mammal sleep dataset that we left off with last week. 
Load the dataset
```{r}
sleep <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/sleep.csv", stringsAsFactors = T)
#need to use stringsAsFactors to make characters read in as factors
```

Last time you used the built-in plot functions to do some plots. Let’s replace 
those with ggplot2 and do some more.

14. First plot how TotalSleep is explained by BrainWt (remember the issues with 
the data).  Use ggplot2 to plot the relationship.

```{r}
ggplot(sleep[sleep$BrainWt <1000, ], aes(x=BrainWt, y = TotalSleep)) +
  geom_point(, size = 4) +
  #below here is ylabel, xlabel, and main title
  ylab("Average hours spent \n sleeping daily") +
  xlab("Brain weight (g)") +
  ggtitle("Time spent sleeping \n decreases with brain \n weight") +
  #theme sets sizes, text, etc
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32),
        # change plot background, grid lines, etc (just examples so you can see)
        panel.background = element_rect(fill="white"),
        panel.grid.minor.y = element_line(size=3),
        panel.grid.major = element_line(colour = "black"),
        plot.background = element_rect(fill="gray"),
        legend.background = element_rect(fill="gray"))
```


15. Next color code each plot point by whether or not its a primate.  In order 
to do this you can use the Primate column or (following class code) make a new 
column called Taxa to represent the information (hint:search for “ revalue”). 
Make sure axes are well-labeled.

```{r}
library(plyr)
sleep$Taxa <- revalue(sleep$Primate, c(Y = "Primate", N = "Non-primate"))
sleep$Taxa <- relevel(sleep$Taxa, "Primate")

ggplot(sleep[sleep$BrainWt <1000, ], aes(x=BrainWt, y = TotalSleep)) +
  geom_point(aes(colour=Taxa), size = 4) +
  #below here is ylabel, xlabel, and main title
  ylab("Average hours spent \n sleeping daily") +
  xlab("Brain weight (g)") +
  ggtitle("Time spent sleeping \n decreases with brain \n weight") +
  #scale commands help with legends
  scale_colour_manual(name="Type of mammal",values = c("#FFA373","#50486D")) +
  #theme sets sizes, text, etc
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32),
        # change plot background, grid lines, etc (just examples so you can see)
        panel.background = element_rect(fill="white"),
        panel.grid.minor.y = element_line(size=3),
        panel.grid.major = element_line(colour = "black"),
        plot.background = element_rect(fill="gray"),
        legend.background = element_rect(fill="gray"))
```

16. Let’s work with histograms.
* What type of variation do we see in total time spent sleeping? Create a 
histogram to explore this issue.

```{r}
ggplot(sleep
       , aes(x=TotalSleep)) +
  geom_histogram() +
  xlab("Total sleep (hours per day")+ggtitle("Variation in sleep levels")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```

* Facet the graph you created based on whether or not the animal is a primate 
(Primate column).

```{r}
ggplot(sleep
       , aes(x=TotalSleep)) +
  geom_histogram() +
  xlab("Total sleep (hours per day")+ggtitle("Variation in sleep levels")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))+ 
  facet_wrap(~Taxa)
```

* Now only graph the data for primates.

```{r}
ggplot(sleep[sleep$Taxa == "Primate",]
       , aes(x=TotalSleep)) +
  geom_histogram() +
  xlab("Total sleep (hours per day")+ggtitle("Variation in sleep levels")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```


17. Develop a properly-labelled bar graph with error bars to explore how total 
sleep changes with 
* Primate (relabel Primate/Non-Primate as Yes/No; note 
there are multiple ways to do this!) – use a 95% confidence interval for the bar

```{r}
#use summarySE function from Rmisc package
sleep$Primate <- revalue(sleep$Primate, c(Y = "Yes", N = "No"))
sleep$Primate <- relevel(sleep$Primate, "No")
library(Rmisc)
sleep_by_primate <- summarySE(sleep, measurevar = "TotalSleep", groupvars = "Primate", na.rm = T)
#look at it
sleep_by_primate
require(ggplot2)
ggplot(sleep_by_primate
       , aes(x=Primate, y=TotalSleep)) +
  geom_col(size = 3) +
  geom_errorbar(aes(ymin=TotalSleep-ci, ymax=TotalSleep+ci), size=1.5) +
  ylab("Total sleep \n (hours per day")+ 
  xlab("Primate?")+ 
  ggtitle("Sleep across different taxa")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32)) 

```

* Predation risk (as a factor!) – use 1 standard error for the bar. Note the difference!

```{r}
sleep$Predation <- as.factor(sleep$Predation)
sleep_by_predation <- summarySE(sleep, measurevar = "TotalSleep", 
                                groupvars = "Predation", na.rm = T)
#look at it
sleep_by_predation
require(ggplot2)
ggplot(sleep_by_predation
       , aes(x=Predation, y=TotalSleep)) +
  geom_col(size = 3) +
  geom_errorbar(aes(ymin=TotalSleep-se, ymax=TotalSleep+se), size=1.5) +
  ylab("Total sleep \n (hours per day)") + 
  ggtitle("Sleep across different \n predation levels")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

```

