---
title: "5. ANOVAs"
author: "jsg"
date: "9/22/2020"
output: github_document
---
Before doing this, review the **Week 5** lecture set slides from 
https://sites.google.com/view/biostats/bio-7800178002/week-5 and
the  **7_ANOVAs.R** and **8_Blocking_and_interactions.R** script in the lecture files folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats).
Make sure you are comfortable with null and alternative hypotheses and appropriate plots
for all examples.


Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  * **push** updates to github
  
## Just for practice. Do this problems using ANOVA, Kruskal-Wallis, and bootstrapping methods.  Make sure you can plot the data and carry out multiple comparison methods as needed. Also be sure to understand the use of coefficients and adjusted R2 values and where to find them. 

1. Use the iris dataset in R to determine if petal length differs among species.  


```{r}
#plot
library(Rmisc)

function_output <- summarySE(iris, measurevar="Petal.Length", groupvars =
                               c("Species"))
library(ggplot2)
ggplot(function_output, aes(x=Species, y=Petal.Length)) +
  geom_col(aes(fill=Species), size = 3) +
  geom_errorbar(aes(ymin=Petal.Length-ci, ymax=Petal.Length+ci), size=1.5) +
  ylab("Petal Length (cm)")+ggtitle("Petal Length of \n various iris species")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

petal <- lm(Petal.Length ~ Species, iris)
plot(petal)
library(car)
Anova(petal, type = "III")
#compare to
summary(petal)
library(multcomp)
comp_cholest <- glht(petal, linfct = mcp(Species = "Tukey"))
summary(comp_cholest)

#kw approach
petal <- kruskal.test(Petal.Length ~ Species, iris)
pairwise.wilcox.test(iris$Sepal.Length, 
                          iris$Species, 
                          p.adjust.method="holm")

#bootstrap
library(WRS2)
t1waybt(Petal.Length~Species, iris)
bootstrap_post_hoc <- mcppb20(Petal.Length~Species, iris)
#use p.adjust to correct for FWER
p.adjust(as.numeric(bootstrap_post_hoc$comp[,6]), "holm")
```
*Answer: We used an ANOVA (a special case of linear models) to investigate how a numerical response variable 
differed among 3 groups.  This was appropriate as evidenced by the residual plots
(there is no pattern in the residuals and they are normally distributed),
but other methods are demonstrated as well.*  

*Using an ANOVA, we found F~2,147~= 1180.2, which led to a p-value of <.001.  Given
this, I reject the null hypothesis there is no difference among mean measurements
for each species.  
Post-hoc testing indicated all species significantly differed from all others 
(all p <.05)
using a Tukey approach to control for family-wise error rate. Kruskal-Wallis and
bootstrapping approaches led to similar conclusions.*

## t-tests are special cases of ANOVAs, which are special cases of linear models!

2. Data on plant heights (in cm) for plants grown with a new and old formulation 
of fertilizer can be found at

https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/fertilizer.csv.

Analyze this data using the t.test function and the lm function to convince yourself the outcome is the same.

```{r}
fertilizer <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/fertilizer.csv",
                       stringsAsFactors = T)
#note use of var.equal!  assumption of ANOVAs
t.test(height ~ fertilizer, fertilizer, var.equal = T)
fert_lm <- lm(height ~ fertilizer, fertilizer)
plot(fert_lm)
summary(fert_lm)
require(car)
Anova(fert_lm, type = "III")
```
*Answer: t-tests and ANOVA (lm) approaches yield the same results. Note for the 
tests to match exactly we have to assume equal variances among groups for the 
t-tests.In both we reject the null hypothesis
of no difference among mean height
of plants based on fertilizer.  Notice the t statistic (2.9884) is the square root
of the F statistic (8.931).  The t distribution corresponds to the F with only 1
df in the numerator (so its not listed!).*


## For the following questions, pick the appropriate method for analyzing the question.  Use a plot of the data and/or model analysis to justify your decision.  Make sure you can carry out multiple comparison methods as needed. Also be sure to understand the use of coefficients and adjusted R^2^ values and where to find them.

3. Data on sugar cane yield for multiple fields is available using 

read.table("https://docs.google.com/spreadsheets/d/e/2PACX-1vRjstKreIM6UknyKFQCtw2_Q6itY9iOAVWO1hUNZkBFL8mwVssvTevqgzV22YDKCUeJq0HBDrsBrf5O/pub?gid=971470377&single=true&output=tsv", 
header = T, stringsAsFactors = T)

More info on the data can be found at 
http://www.statsci.org/data/oz/cane.html.
Is there evidence that location (DistrictPosition column) impacts yield 
(Tonn.Hect column)?  If so, which areas are driving this distance?

```{r}
cane <- read.table("https://docs.google.com/spreadsheets/d/e/2PACX-1vRjstKreIM6UknyKFQCtw2_Q6itY9iOAVWO1hUNZkBFL8mwVssvTevqgzV22YDKCUeJq0HBDrsBrf5O/pub?gid=971470377&single=true&output=tsv", header = T, stringsAsFactors = T)
summary(cane)
cane_summary <- summarySE(cane, measurevar="Tonn.Hect", groupvars =
                               c("DistrictPosition"))

ggplot(cane_summary, aes(x=DistrictPosition, y=Tonn.Hect)) +
  geom_col(size = 3) +
  geom_errorbar(aes(ymin=Tonn.Hect-ci, ymax=Tonn.Hect+ci), size=1.5) +
  ylab("Production (tonnes per hectare)") +
  xlab("District Position") +
  ggtitle("Production differs \n among locations") +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

impact_district <- lm(Tonn.Hect ~ DistrictPosition, cane)
summary(impact_district)
plot(impact_district)#not really normal...lets bootstrap
require(WRS2)
t1waybt(Tonn.Hect ~ DistrictPosition, cane)
mcppb20(Tonn.Hect ~ DistrictPosition, cane)
p <- mcppb20(Tonn.Hect ~ DistrictPosition, cane)
p.adjust(as.numeric(p$comp[,6]), "holm")


#compare to lm apporach
require(car)
Anova(impact_district, type = "III")
require(multcomp)
comp_district <- glht(impact_district, linfct = mcp(DistrictPosition = "Tukey"))
summary(comp_district)

```
*Answer: For this analysis I used a bootstrap approach as the residual plots
suggested a non-normal distribution.  Analysis revealed a test statistics of 
16.52 and p-value of 0, so I reject the null hypothesis of no difference among
Districts.  Post-hoc tests reveal all districts differ from each other except for
south and east, south and central, and east and central (using sequential FDR 
to control for family-wise error rate.)*

*Note that a linear model does lead to slightly different findings regarding which
districts differ from which others.*

4. Data on FEV (forced expiratory volume), a measure of lung function, can be 
found at 
http://www.statsci.org/data/general/fev.txt
More information on the dataset is available at 
http://www.statsci.org/data/general/fev.html.
Is there evidence that FEV depends on gender? If so, which gender has the higher
FEV score?  How much variance does gender explain? 

```{r}
fev <- read.table("http://www.statsci.org/data/general/fev.txt", header = T,
                  stringsAsFactors = T)
fev_summary <- summarySE(fev, measurevar="FEV", groupvars =
                               c("Sex"))

ggplot(fev_summary, aes(x=Sex, y=FEV)) +
  geom_col(size = 3) +
  geom_errorbar(aes(ymin=FEV-ci, ymax=FEV+ci), size=1.5) +
  ylab("FEV (liters)") +
  xlab("Sex") +
  ggtitle("FEV differs \n among males and females") +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))


fev_gender <- lm(FEV ~ Sex, fev)
plot(fev_gender) #anova is fine
summary(fev_gender)
```

*I used an ANOVA (or linear model, or t-test, here, all the same since 2 groups!)
to consider the impact of sex on FEV.  This was appropriate as evidenced by the
residual plots
(there is no pattern in the residuals and they are normally distributed).  Results indicate there is a difference among sexes (F~1,652~ = 29.607, p<.001). There is 
no need for post-hoc tests here since there are only 2 groups being considered.  
Coefficients related to the groups (note female is replaced by intercept here, 
and the SexMale coefficient is relative to that) indicates that males have a 
higher FEV on average.  Graphs also show this relationship.*

5. A professor carried out a long-term study to see how various factors impacted
pulse rate before and after exercise.  Data can be found at 
http://www.statsci.org/data/oz/ms212.txt
With more info at 
http://www.statsci.org/data/oz/ms212.html.
Is there evidence that frequency of exercise (Exercise column) impacts resting
pulse rate (Pulse1 column)?

```{r}
pulse <- read.table("http://www.statsci.org/data/oz/ms212.txt", header = T, 
                    stringsAsFactors = T)
pulse$Exercise <- factor(pulse$Exercise)
library(plyr)
pulse$Exercise <- revalue(pulse$Exercise, c("1" = "high", 
                                            "2" = "moderate", 
                                            "3" = "low"))

exercise_summary <- summarySE(pulse, measurevar="Pulse1", groupvars =
                               c("Exercise"), na.rm = T)

ggplot(exercise_summary, aes(x=Exercise, y=Pulse1)) +
  geom_col(size = 3) +
  geom_errorbar(aes(ymin=Pulse1-ci, ymax=Pulse1+ci), size=1.5) +
  ylab("Resting pulse (beats per minute)") +
  xlab("Exercise level")+
  ggtitle("Resting pulse decreases \n with exercise") +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
initial <- lm(Pulse1 ~ Exercise, pulse)
plot(initial)
summary(initial)
Anova(initial, type = "III") 
```
*I used a linear model to test for difference in mean resting pulse rates based 
on group. 
This was appropriate as evidenced by the
residual plots
(there is no pattern in the residuals and they are normally distributed). 
Analyses suggest there is no difference in resting pulse rate among groups
(F~2,106~ = 2.8, p = .07), so I fail to reject the null hypothesis that a 
difference does not exist in mean resting pulse rate based on groups. The 
alternative hypothesis is that a 
difference does  exist in mean resting pulse rate based on groups.*

6. Continuing with the pulse dataset, is there evidence frequency of exercise
impacts change in pulse rate?  Note we should only consider change for students 
who ran between the first and second pulse measures (Ran column = 1).

```{r}
pulse$change <- pulse$Pulse2 - pulse$Pulse1

change_summary <- summarySE(pulse[pulse$Ran == 1, ], measurevar="change", groupvars =
                               c("Exercise"), na.rm = T)

ggplot(change_summary, aes(x=Exercise, y=change)) +
  geom_col(size = 3) +
  geom_errorbar(aes(ymin=change-ci, ymax=change+ci), size=1.5) +
  ylab("Resting pulse \n (beats per minute)") +
  xlab("Exercise level")+
  ggtitle("Change in pulse does \n not differ among groups") +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

change <- lm(change ~ Exercise, pulse[pulse$Ran == 1, ])
plot(change)
Anova(change, type = "III")
```
*I used a linear model to test for difference in change in pulse rates (before 
and after running) based 
on group. 
This was appropriate as evidenced by the
residual plots
(there is no pattern in the residuals and they are normally distributed). 
Analyses suggest there is no difference in resting pulse rate among groups
(F~2,43~ = .09, p = .90), so I fail to reject the null hypothesis that a 
difference does not exist in change in pulse rates (before 
and after running) based 
on group.  The 
alternative hypothesis is that a 
difference does  exist in change in pulse rates (before 
and after running) based 
on group.*

7. A survey was conducted to see if athletes and non-athletes deal with anger in
the same way.  Data is @

angry <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSaawG37o1ZUEs1B4keIJpZAY2c5tuljf29dWnzqQ0tHNCzfbz85AlWobYzBQ3nPPXJBLP-FWe4BNZB/pub?gid=1784556512&single=true&output=csv", stringsAsFactors = T)

and more information is at 

http://onlinestatbook.com/case_studies/angry_moods.html.

Focus on the following variables:

Sports
1 = athletes, 2 = non-athletes
Gender
1 = males, 2 = females
Expression (AE)
index of general anger expression: 
(Anger-Out) + (Anger-In) - (Control-Out) - (Control-In) + 48

Is there any evidence that gender or athlete status impact how anger is expressed?

```{r}
angry <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSaawG37o1ZUEs1B4keIJpZAY2c5tuljf29dWnzqQ0tHNCzfbz85AlWobYzBQ3nPPXJBLP-FWe4BNZB/pub?gid=1784556512&single=true&output=csv", stringsAsFactors = T)
str(angry)
angry$Gender <- as.factor(angry$Gender)
angry$Gender <- revalue(angry$Gender, c("1" = "athlete", 
                                        "2" = "non-athlete"))
angry$Sports <- as.factor(angry$Sports)
angry$Sports <- revalue(angry$Sports, c("1" = "male",
                                        "2" = "female"))

anger_summary <- summarySE(angry, measurevar="Anger_Expression", groupvars =
                               c("Sports", "Gender"), na.rm = T)
ggplot(anger_summary, aes(x=Gender, y=Anger_Expression, color=Sports, 
                                   shape = Sports)) +
  geom_point(size = 3) +
  geom_line(aes(group=Sports, linetype =Sports), size=2) +
  geom_errorbar(aes(ymin=Anger_Expression-ci, ymax=Anger_Expression+ci), size=1.5) +
  ylab("Anger level")+ 
  xlab("Experience") + 
  scale_shape_discrete(guide=FALSE)+
  scale_linetype_discrete(guide=FALSE)+
  ggtitle("Anger level among groups")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))


angry_gender <- lm(Anger_Expression ~ Sports * Gender, angry)
plot(angry_gender)
Anova(angry_gender, type = "III")

#remove interaction since not significant
angry_gender <- lm(Anger_Expression ~ Sports + Gender, angry)
plot(angry_gender)
Anova(angry_gender, type = "III") #only differs among those who play sports

```


8. Back to the pulse data (#5).  Is there evidence that frequency of exercise
(Exercise column) and gender impact change in pulse rate for students who ran 
(Ran column = 1)?

```{r}
pulse <- read.table("http://www.statsci.org/data/oz/ms212.txt", header = T, 
                    stringsAsFactors = T)
pulse$Exercise <- factor(pulse$Exercise)
library(plyr)
pulse$Exercise <- revalue(pulse$Exercise, c("1" = "high", 
                                            "2" = "moderate", 
                                            "3" = "low"))
pulse$Gender <- factor(pulse$Gender)
pulse$Gender <- revalue (pulse$Gender, c("1" = "male", "2" = "female"))
pulse$change <- pulse$Pulse2 - pulse$Pulse1
change_summary <- summarySE(pulse[pulse$Ran == 1, ], measurevar="change", groupvars =
                               c("Exercise", "Gender"), na.rm = T)

ggplot(change_summary, aes(x=Gender, shape = Exercise, color = Exercise,
                           y=change)) +
  geom_point(size = 3) +
  geom_line(aes(group=Exercise, linetype =Exercise), size=2) +
  geom_errorbar(aes(ymin=change-ci, ymax=change+ci), size=1.5) +
  ylab("Change in pulse \n (beats per minute)") +
  scale_color_discrete(name = "Exercise level")+
  scale_shape_discrete(guide=FALSE)+
  scale_linetype_discrete(guide=FALSE)+
  ggtitle("Change in pulse does \n not differ among groups") +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```
*I first read in and recoded some data for ease and plotting. I then produced a 
plot to consider the null hypotheses that* 

* *exercise level does not influence change in pulse rate*
* *gender does not influence change in pulse rate*
* *gender and exercise level do not interact to influence change in pulse rate*

```{r}
exercise <- lm(change ~ Gender * Exercise, pulse[pulse$Ran == 1, ])
summary(exercise)
Anova(exercise, type = "III")

#rerun without interaction
exercise <- lm(change ~ Gender + Exercise, pulse[pulse$Ran == 1, ])
summary(exercise)
Anova(exercise, type = "III") #no significance
```

*I then analyzed the data using factorial ANOVA.
The outcome is continuous 
and both explanatory variables are categorical. The design is also fully randomized. Residual plots indicated all assumptions were met
(there is no pattern in the residuals and they are normally distributed).
Analysis shows an insignificant interaction (F~2,40~=1.2, p=.21) between 
exercise level and gender, so I removed the interaction term. The reduced model
showed neither gender (F~1,42~=.16, p =.69) or exercise level (F~2,42~=.1, p=.90) 
influenced change in pulse rate, so I failed to reject the related null 
hypotheses.*


9. (Continuing from question 9) I’m curious if the treatment (running vs not) even had an impact, and if so did it depend on gender. How would you test this?  

```{r}
pulse$Ran <- factor(pulse$Ran)
pulse$Ran <- revalue(pulse$Ran, c("1" = "Ran", "2" = "Sat"))
run_summary <- summarySE(pulse, measurevar="change", groupvars =
                               c("Ran", "Gender"), na.rm = T)

ggplot(run_summary, aes(x=Gender, shape = Ran, color = Ran,
                           y=change)) +
  geom_point(size = 3) +
  geom_line(aes(group=Ran, linetype = Ran), size=2) +
  geom_errorbar(aes(ymin=change-ci, ymax=change+ci), size=1.5) +
  ylab("Resting pulse \n (beats per minute)") +
  xlab("Gender")+
  ggtitle("Change in pulse  \n  differed among groups") +
  scale_color_discrete(name = "Ran")+
  scale_shape_discrete(guide=FALSE)+
  scale_linetype_discrete(guide=FALSE)+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

pulse$Ran <- as.factor(pulse$Ran)
levels(pulse$Ran) <- c("ran", "sat")
running <- lm(Pulse2 ~ Gender * Ran, pulse)
Anova(running, type = "III") 

#consider without interaction
running <- lm(Pulse2 ~ Gender + Ran, pulse)
Anova(running, type = "III")  #running mattered
summary(running) #running increased pulse
```
*After plotting the data, I used a linear model to test the null hypotheses that*
* *change in pulse rate did not depend on gender*
* *change in pulse rate did not depend on running treatment*
* *there was no interaction between running treatment and gender on change in pulse
rate*

*Analyses indicated all model assumptions were met(
there is little pattern in the residuals and they are fairly normally distributed,
though you may need to rely more on robustness of test here!
) and there is not enough evidence for an interaction between gender and running
treatment on change in pulse rate (F~1,105~=1.3, p=.25). After removing the interaction,
analysis of a reduced model indicates only running treatment (F~1,106~=212, p<.01)
impacts change in pulse rate. Running did increase pulse rate.*

