---
title: "6. Regression, correlation, and ANCOVAs"
author: "jsg"
date: "9/30/2020"
output: github_document
---
Before doing this, review the **Week 65** lecture set slides from 
https://sites.google.com/view/biostats/bio-7800178002/week-6 and
the  **9_Regression_and_correlation.R** and **10_ANCOVA_and_Regression.R
(up to line 248)** script in the lecture files folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats). 
Make sure you are comfortable with null and alternative hypotheses and appropriate plots
for all examples. Also make sure you understand the difference between
correlation and regression.


Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  * **push** updates to github
  

1.  A professor carried out a long-term study to see how various factors impacted
pulse rate before and after exercise.  Data can be found at 

http://www.statsci.org/data/oz/ms212.txt

With more info at 

http://www.statsci.org/data/oz/ms212.html.

Is there evidence that age, height, or weight impact change in pulse rate for students who ran (Ran column = 1)?

```{r}
pulse <- read.table("http://www.statsci.org/data/oz/ms212.txt", header = T, stringsAsFactors = T)
pulse$change <- pulse$Pulse2 - pulse$Pulse1
#need to make columns entered as numeral change to factor, although it doesn't 
#really matter when only 2 groups (why?)
pulse$Exercise <-as.factor(pulse$Exercise)
pulse$Gender <- as.factor(pulse$Gender)

#age
exercise <- lm(change ~ Age, pulse[pulse$Ran == 1, ])
par(mfrow =c (2,2))
plot(exercise)
require(car)
Anova(exercise, type = "III")
summary(exercise)
```

*First we need to make a column that shows change in pulse rate. We also should 
change Exercise and gender to factors.*  

*For age we note the model meets assumptions. We also find no evidence that 
age impacts change (F~1,44~ = .4950, p = 0.49). We do not reject our null hypothesis.*  

```{r}
#weight
exercise <- lm(change ~ Weight, pulse[pulse$Ran == 1, ])
par(mfrow =c (2,2))
plot(exercise)
Anova(exercise, type = "III")
summary(exercise)
```
*For weight we note the model meets assumptions. We also find no evidence that 
weight impacts change (F~1,44~ = .4027, p = 0.53). We fail to reject our null 
hypothesis that there is no relationship between weight and change in pulse rate.
 We
also note that weight only explains 1% of the variation in change in pulse rate 
(likely due to chance!).*  *  

```{r}
#height
exercise <- lm(change ~ Height, pulse[pulse$Ran == 1, ])
par(mfrow =c (2,2))
plot(exercise)
Anova(exercise, type = "III")
summary(exercise)
```
*For height we note the model meets assumptions. We also find no evidence that 
weight impacts change (F~1,44~ = 1.15, p = 0.29). We fail to reject our null 
hypothesis that there is no relationship between height and change in pulse rate.
 We
also note that age only explains 2.5% of the variation in change in pulse rate 
(likely due to chance!).*    

2. (from OZDASL repository, http://www.statsci.org/data/general/stature.html; reference for more information)

When anthropologists analyze human skeletal remains, an important piece of
information is living stature. Since skeletons are commonly based on
statistical methods that utilize measurements on small bones. The following
data was presented in a paper in the American Journal of Physical
Anthropology to validate one such method.
Data is available @

http://www.statsci.org/data/general/stature.txt

as a tab-delimted file (need to use read.table!)
Is there evidence that metacarpal bone length is a good predictor of
stature? If so, how much variation does it account for in the response
variable?

```{r}
height <- read.table("http://www.statsci.org/data/general/stature.txt", 
                     header = T, stringsAsFactors = T)
head(height)
metacarp_relationship <- lm(Stature ~ MetaCarp, height)
plot(metacarp_relationship)
Anova(metacarp_relationship, type = "III")
summary(metacarp_relationship)
```
*To consider the relationship among these continuous variables, we used linear regression. 
Analysis of model assumptions suggest assumptions are met, although the dataset is 
small.  Analysis suggests there is a significant positive relationship between 
metacarpal length and stature (F~1,7~ = 19.19, p = 0.003). The R^2^ value indicates
that metacarpal length explains 73% of the variation in stature. Coefficients 
indicate that stature increases with increasing metacarpal length.*

3. Data on medals won by various countries in the 1992 and 1994 Olympics is
available in a tab-delimited file at 

http://www.statsci.org/data/oz/medals.txt

More information on the data can be found at: 

http://www.statsci.org/data/oz/medals.html

Is there any relationship between a country’s population and the total
number of medals they win?

```{r}
medals <- read.table(header = T, "http://www.statsci.org/data/oz/medals.txt", 
                     stringsAsFactors = T)
head(medals)
medals$total <- medals$Summer + medals$Winter
population_medals <- lm(total ~ Population, medals)
plot(population_medals)
summary(population_medals)
Anova(population_medals, type = "III")
cor.test(~total + Population, medals, method = "spearman")
```
*There is a high leverage point in the dataset (row 4), but residuals appear to 
be fairly normally distibuted and little structure exists in the graph of Residuals 
vs. Fitted Values.  Analysis using linear regression suggests a significant (
F~1,60~ = 10.45, p = 0.002) positive relationship between population size and 
medal count that explains ~15% of the variation in the response variable.  Rank-
correlation analysis also indicated this relationship exists.*



4. Continuing with the Olympic data, is there a relationship between the
latitude of a country and the number of medals won in summer or winter
Olympics?

```{r}
#still using medals
summer_medals <- lm(Summer ~ Latitude, medals)
plot(summer_medals)
Anova(summer_medals, type = "III")
summary(summer_medals)

winter_medals <- lm(Winter ~ Latitude, medals)
plot(winter_medals)
Anova(winter_medals, type = "III")
summary(winter_medals)
```

*Visual analysis of residuals from both models show some structure in the residual
and deviations from normality, but we continue on with linear regression given the 
small sample size.  Both summer and winter medal counts are positively (surpisingly)
and significantly (both p <.05)
related to latitude, with latitude explaining ~17% of the variation in winter medal
count and ~8% of the data in summer medal count.  

5. Data on FEV (forced expiratory volume), a measure of lung function, can
be found at 

http://www.statsci.org/data/general/fev.txt

More information on the dataset is available at 

http://www.statsci.org/data/general/fev.html.

Is there evidence that FEV depends on age or height? If so, how do these
factors impact FEV, and how much variance does each explain? 

```{r}
fev <- read.table("http://www.statsci.org/data/general/fev.txt", header = T, 
                  stringsAsFactors = T)
head(fev)
fev_height <- lm(FEV ~ Height, fev)
plot(fev_height)
Anova(fev_height, type = "III")
summary(fev_height)
```
Model assumptions appear to be met. Height appears to have a positive relationship
with FEV (F~1,652~ = 1995, p<.001).  

```{r}
fev_age <- lm(FEV ~ Age, fev)
plot(fev_age)
Anova(fev_age, type = "III")
summary(fev_age)
```

Model assumptions appear to be met. Age appears to have a positive relationship
with FEV (F~1,652~ = 872.2, p<.001).  

6. Continuing with the FEV data, produce plots that illustrate how height, age, and gender each impact FEV.  

```{r}
library(ggplot2)
#age plot####
ggplot(fev, aes(x=Age, y=FEV)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm") +
  ylab("FEV (L)")+ggtitle("FEV increases with age")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

#height plot####
ggplot(fev, aes(x=Height, y=FEV)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm") +
  ylab("FEV (L)")+ggtitle("FEV increases with height")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))



#gender plot ####

#bar chart with error bars ####
library(Rmisc)
function_output <- summarySE(fev, measurevar="FEV", groupvars =
                               c("Sex"))

ggplot(function_output, aes(x=Sex, y=FEV)) +
  geom_col(size = 3) +
  ylab("FEV (L)") +
  ggtitle("FEV is higher in males ")+
  geom_errorbar(aes(ymin=FEV-ci, ymax=FEV+ci), size=1.5) +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

```


7. Does the impact of age on FEV differ among genders?

```{r}
fev_age <- lm(FEV ~ Age*Sex, fev)
plot(fev_age)
Anova(fev_age, type = "III")
summary(fev_age)
```

*We can explore this question using an ANCOVA since the response is continuous and
the explanatory variables combine a categorical and continuous variable. Analysis of
residuals indicates the assumptions are met (no pattern, normal distribution). 
There is a significant interaction among age and gender on FEV (F~1,650~=64.535, 
p<.001).  We should explore impacts of age on each gender separately.*

```{r}
fev_age <- lm(FEV ~ Age, fev[fev$Sex == "Male",])
plot(fev_age)
Anova(fev_age, type = "III")
summary(fev_age)
```
*Age has a significant (F~1,334~ = 641, p < 0.01) positive 
(.27 L yr^-1^) impact on FEV in males.*

```{r}
fev_age <- lm(FEV ~ Age, fev[fev$Sex == "Female",])
plot(fev_age)
Anova(fev_age, type = "III")
summary(fev_age)
```
*Age also has a significant (F~1,316~ = 380, p < 0.01) positive 
(.16 L yr^-1^) impact on FEV in females. The interaction is likely
due to the higher rate of increase of FEV with age in males.*

```{r}
ggplot(fev, aes(x=Age, y=FEV, color = Sex, shape = Sex)) +
  geom_point(size = 3) +
  ylab("FEV (L)") +
  ggtitle("FEV increases faster \n with age in males")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32)) +
    geom_smooth(method = "lm", se = F)
```


8. Data on home gas consumption at various temperatures before and after new insulation was installed has been collected @ 

http://www.statsci.org/data/general/insulgas.txt

More information on the data is available @

http://www.statsci.org/data/general/insulgas.html

Is there any relationship between these factors?  How would you test this,
and what type of plot would you produce to accompany your analysis?

```{r}
heat <- read.table("http://www.statsci.org/data/general/insulgas.txt", 
                   header = T, stringsAsFactors = T)
head(heat)
heat_model <- lm(Gas ~ Insulate * Temp, heat)
plot(heat_model)
require(car)
Anova(heat_model, type = "III")

ggplot(heat, aes_string(x="Temp", y="Gas", color = "Insulate")) +
  geom_point(size = 3) +
  ylab(expression(paste("Gas (1000 ",ft^3, ")")))+
  xlab(expression(paste("Temperature (", degree~C, ")")))+
  geom_smooth(method = "lm", se = F) +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))


```
*There is a significant relationship between insulation type (before/after) and
temperature on gas usage. Graphical analysis indicates the old (before) insulation led to higher overall gas usage and gas usage increased faster with colder temperature
compared to the new insulation. Statistical analysis bears this out*

```{r}
heat_model_old <- lm(Gas ~ Temp, heat[heat$Insulate == "Before",])
plot(heat_model_old)
summary(heat_model_old)
Anova(heat_model_old, type = "III")

heat_model_new<- lm(Gas ~ Temp, heat[heat$Insulate == "After",])
plot(heat_model_new)
summary(heat_model_new)
Anova(heat_model_new, type = "III")
```
*There is a significant relationship between gas usage and temperature for old 
and new insulation homes. However, the old insulation led to using 400 ft^3^ more gas per week to heat the 
house with every degree
drop in temperature, while the new insulation leads to a increase of only 
250 ft^3^ more gas per week with each degree drop.*  


