---
title: "2024 Spring Midterm (29 points total) solutions"
format:
  html:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
    theme: readable
---

## Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and those you produced in class). However, you may
not get external help ( from other students, directed internet searches,
etc.). Please update the statement below to acknowledge these
instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students. Infractions will
also result in me being unable to obtain a letter of recommendation from
the department for professional school applications.

Good luck!

## Background

Billion Oyster Project (BOP) is a local group in New York focused on
restoring oysters. They are considering new restoration locations, one
in Manhattan and one in Queens. To help determine which site is better,
they put out oyster cages (small mesh cages holding live oysters) at
each site. Each cage held 50 oysters. Oysters at each site were measured
after 3 months. Now they need help analyzing the data.

## #1

At the Queens site, 15 oysters were dead after 3 months. At the
Manhattan site, 21 were dead. Is there a difference in mortality among
the sites?

Investigate the question. Make sure you include

-   null hypothesis

    -   *H~0~: \# there is no difference in mortality rate/proportion
        among sites; mortality rate is not contingent on site*

-   alternative hypothesis

    -   *H~A~: \# there is a difference in mortality rate/proportion
        among sites; mortality rate is contingent on site*

-   explanation for test you will use

    -   I will attempt to use a $\chi^2$ test given we have discrete
        data from 2 populations.

-   results from statistical test

```{r}
chisq.test(matrix(c(15,21,50-15,50-21), nrow=2, byrow = T))
```

Note

```{r}
chisq.test(matrix(c(15,21,50-15,50-21), nrow=2, byrow = F))
```

-   clear explanation of how results relate to your stated hypotheses
    -   The test led to a value of $\chi^2 = 1.0851$, with a
        corresponding p-value of 0.2976. Thus I fail to reject the null
        hypothesis.

## #2

Make the data

```{r}
set.seed(19)
length <- data.frame(site = c(rep("Queens", 35), rep("Manhattan", 29), rep("Gov_Island", 50)), length=c(rnorm(35,40,5), rnorm(29,39,5), rnorm(50,35,4)))
length$site <- factor(length$site)
length_model <- lm(length~site, length)
library(car)
Anova(length_model, type="III")
library(multcomp)
summary(glht(length_model, linfct = mcp(site="Tukey")))
write.csv(length, "oyster_measures.csv", row.names = F)

```

The oysters that were alive at the end of the sampling period were
measured (length, mm). Data is available using

```{r}
oyster_measures <-read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQAWMON6TYoGDl2VYhK9ONDJW3eul7n0-hw-62F4FQ56bnKuKm8N32jtOMktuLHoS-Lbvl6kmgeUQ6x/pub?gid=0&single=true&output=csv", stringsAsFactors = T)
```

The dataset also includes a sample from Governor's Island, a current
restoration site. How do the sites compare?

Investigate the question. Make sure you include

-   null hypothesis

    -   *H~0~: \# there is no difference in mean length among the 3
        sites;*
        $\mu_{length, Queens}=\mu_{length,Manhattan}=\mu_{length,\textrm{Governor's Island}}$

-   alternative hypothesis

    -   *H~A~: \# there is a difference in mean length among the 3
        sites; means are not all equal; at least one population mean is
        different from the others*

-   explanation for test you will use

    -   the outcome is a continuous variable, and we are considering
        differences among 3 groups. Given the large sample size (\>25)
        and focus on length, I will use an ANOVA and check assumptions.

-   results from statistical test

```{r}
length_model_om <- lm(length~site, oyster_measures)
library(car)
Anova(length_model_om, type="III")
plot(length_model_om)
library(multcomp)
summary(glht(length_model_om, linfct = mcp(site="Tukey")))

```

-   clear explanation of how results relate to your stated hypotheses
    -   All assumptions were met, so I used the ANOVA. The overall test
        led to a p-value of \<0.001 (F~2,111~=18.986), thus I reject the
        null hypothesis. Follow-up post-hoc tests (using the Tukey
        approach to account for FWER) indicated Governor's Island
        differed significantly from the other sites.

## #3

Graph the data from question 2.

Several options

```{r}
library(ggplot2)
library(Rmisc)
oyster_measure_summary <- summarySE(oyster_measures, measurevar = "length", groupvars = "site")

cld_output <- fortify(cld(glht(length_model_om, linfct = mcp(site="Tukey"))))
cld_output$site <- cld_output$lhs

oyster_measure_summary <- merge(oyster_measure_summary, cld_output)


ggplot(oyster_measure_summary, aes(x=site, y=length))+
  geom_col(aes(fill=site)) +
  geom_errorbar(aes(ymin=length-ci, ymax=length+ci)) +
  labs(title="Oyster lengths at 3 restoration sites",
       y= "Oyster length (mm)",
       x= "Site")+
  geom_text(aes(label=letters,y=length+3*ci))

library(plyr)
oyster_measure_summary$Site <- revalue(oyster_measure_summary$site,
                                       c("Gov_Island"="Governor's Island"))
ggplot(oyster_measure_summary, aes(x=Site, y=length))+
  geom_col(aes(fill=Site)) +
  geom_errorbar(aes(ymin=length-ci, ymax=length+ci)) +
  labs(title="Oyster lengths at 3 restoration sites",
       y= "Oyster length (mm)",
       x= "Site")+
  geom_text(aes(label=letters,y=length+3*ci))

ggplot(oyster_measures, aes(x=site, y=length, color=site))+
  geom_jitter() +
  labs(title="Oyster lengths at 3 restoration sites",
       y= "Oyster length (mm)",
       x= "Site")

ggplot(oyster_measures, aes(x=site, y=length, color=site))+
  geom_boxplot() +
  labs(title="Oyster lengths at 3 restoration sites",
       y= "Oyster length (mm)",
       x= "Site")

```

## #4

Extra funding was available for the Queen site to determine the
prevalence of bacterial infection in the oysters. Of the 35 tested
(alive at the end of the 3 months) oysters, 13 were found to have
bacterial infection. Historically, the rate has been 40%. Is there
evidence the infection rate has changed?

Investigate the question. Make sure you include

-   null hypothesis

    -   *H~0~: \# of oysters with bacterial infection is 40%*

-   alternative hypothesis

    -   *H~0~: \# of oysters with bacterial infection is not 40%*

-   explanation for test you will use

    -   *I will use a binomial test to compare the calculated
        transformation rate to a a given value.*

-   results from statistical test

    ```{r}
    binom.test(13,35, .4)
    ```

-   clear explanation of how results relate to your stated hypotheses

    -   *I found a p-value of .87 (\>.05), so I fail to reject the null
        hypothesis. The number of people transforming has not changed.
        **BONUS/not required** It appears to be increasing.*

## #5

The BOP office is confused about your bacteria analysis. They found a
signal of `r 13/35` in the data and took that as p. Why is your value
different? Make sure you explain what a p-value is!

*A p-value is the probability you would see the signal you saw or
something more extreme if the null hypothesis was true and you repeated
the test many times. They are confusing the rate of infection with this
value.*

## #6

The following year, the BOP puts out multiple cages across the Queens
site. They are surprised, given your initial analysis, when all the
cages don't have `r 13/35*100` % showing bacterial infection. Why is
that not true? And is there another summary/output you could provide to
help them think about what they should expect for most cages?

*The p-value we found denotes how likely we are to see `r 13/35`*`*100`%
of oysters showing a bacterial infection if we sampled multiple times
given that we expected 40% of the oysters to have infections. A high (or
low) p-value does not mean we see the value again; due to sampling error
we should expect to get a range of values. We could use a confidence
interval to give a range that will contain the true mean (which should
influence what we see!) for some given % of the time. For example, a 95%
confidence interval

```{r}
library(binom)
binom.confint(13,35)
```

(using the Agresti-Coull method) suggests we 23.1% to 53.7% of oysters
may show infections.
