---
title: "11. Gamma and Bayes intro"
author: "jsg"
date: "11/4/2020"
output: github_document
---
Before doing this, review the **Week 11** lecture set slides from 
https://sites.google.com/view/biostats/bio-7800178002/week-11 and
the  **112_more_model_extensions.R** and 
*15_a_bayes_standalone.R** script in the lecture files folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats). 
Make sure you are comfortable with null and alternative hypotheses for all
examples.

Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  * **push** updates to github
  

1.  Make sure you can describe the main differences between Frequentist, Likelihood, and Bayesian approaches.

2.  Review the video we watched in class to make sure you understand the Bayesian 
connection. You can also read a related post @ https://brilliant.org/wiki/monty-hall-problem/.
* https://www.youtube.com/watch?v=mhlc7peGlGg

3.  I've shared a script in R that lets you test the Monty Hall idea (like in 
the video!).  It's the chivers_monty_hall_script from the 
[code_examples folder](https://github.com/jsgosnell/CUNY-BioStats/tree/master/code_examples)code_examples  
on github.  For this question, its easiest to just source the main file and see what happens.
When you source a script, it is run in R without showing any console output
(but graphs and objects are still produced!).  Try 
*source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/chivers_monty_hall_script.R")*
, then test out the idea here using the following functions which calculate outcomes
under each strategy.
* monty(strat="stay", print_games=F)
* monty(strat="switch", print_games=F)
* monty(strat="random", print_games=F)

```{r}
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/chivers_monty_hall_script.R")
monty(strat="stay", print_games=F)
monty(strat="switch", print_games=F)
monty(strat="random", print_games=F)
```

4. Setup the Monty Hall problem as probabilities and convince yourself how it works.
You may want to remember to think about prior and new information (likelihoods).

5. Run the frog analysis (14/18 frogs are right-pawed) assuming an “uninformed” 
prior (is this really possible?) and priors that predict frogs are likely to be 
left- or right-handed (look under Bayesian analysis in script for functions such
as triplot and qbeta).  Vary both the relationship among the shape variables 
and the magnitude (weighting) to understand how the prior impacts your posterior.

```{r}
library(LearnBayes)
#even, uniform (uninformed) prior
triplot(prior = c(1,1), data = c(14,4), where = "topleft")
#prior assumes left handed
triplot(prior = c(5,20), data = c(14,4), where = "topleft")
#prior assumes right handed 
triplot(prior = c(20,5), data = c(14,4), where = "topleft")
#less sure right handed
triplot(prior = c(4,2), data = c(14,4), where = "topleft")
```
*This is the "big picture" of Bayesian analysis. We combined our prior beliefs and
data to update our beliefs (the posterior). We sample/describe the posterior for
our answer. Everything else is "how do we do that?".*

6. Data on plant heights (in cm) for plants grown with a new and old formulation 
of fertilizer can be found at

https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/fertilizer.csv

Use the data to test the hypothesis that there is no difference in mean plant
heights for the two groups A) Using frequentist methods B) Using Bayesian
approaches.

```{r}
#fertilizer####
fertilizer <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/fertilizer.csv")

t.test(height ~ fertilizer, fertilizer)
library(BayesFactor)
ttestBF(formula = height ~ fertilizer, data = fertilizer)
```


7.Develop a Bayesian model to determine if sepal width (from the iris dataset in
R) differs among populations.

* compare models that parameterize each population as different vs one that only 
examines difference between I. setosa and other species.
  + making a new dummy variable is one way to do this!
