---
title: "3. Continuous tests for 1 or 2 populations"
author: "jsg"
date: "9/8/2020"
output: github_document
---
Before doing this, review the **Week 3** lecture set slides from 
https://sites.google.com/view/biostats/bio-7800178002/week-3 and
the  **Week_3.R** script in the grad_files_2020 folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats).
You’ll also use the qqplot_example.R and
central_limit_theorem.R script from the 
[code_examples folder](https://github.com/jsgosnell/CUNY-BioStats/tree/master/code_examples).

Remember you should
* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  **push** updates to github

## Recognizing and assessing normality 

1. Using the qqplot_example.R code, examine the following distributions and, for 
the continuous distributions (marked with a “*”), observe how a normal probability
plot (qqplot) can be used to visually test for approximate normality.

* *Normal (u= 0; σ2= 10, 100) 
* *Student's t (df = 1, 10, 30, & 100) 
* *Chi-square (df= 1, 2, 5, 30, 50)
* Bernoulli (P=0.1, 0.5, & 0.9) 
* Binomial (P=0.05; N= 2, 5, 25, & 50); (P=0.25; N= 2, 5, 25, & 50); (P=0.50; 
N= 2, 5, 25, & 50); (P=0.75; N= 2, 5, 25, & 50); (P=0.95; N= 2, 5, 25, & 50) 
* Poisson ( u= 2, 5, 10, 30, & 50)

2. Review the central_limit_theorem.R code if you need to convince/remind yourself how common normality of means is for even non-normal data.

## Just for practice (note some sample sizes may be too small for these to all be good ideas!)

3. Seven observers were shown, for a brief period, a grill with 161 flies impaled and were asked to estimate the number. The results are given by Cochran (1954). Based on five estimates, they were 183.2, 149.0, 154.0, 167.2, 187.2, 158.0, and 143.0. Test the null hypothesis that the mean of the population is 161 flies
* Assuming variance = 275
*  Estimating the variance from the data 
* Using rank transform analysis 
* Using binary transform analysis

4. Yields of 10 strawberry plants in a uniformity trial are given by Baker and Baker (1953) as 239, 176, 235, 217, 234, 216, 318, 190, 181, and 225 g. Test the hypothesis that µ = 205
* Assuming variance = 1500
*  Estimating the variance from the data 
* Using rank transform analysis 
* Using binary transform analysis

5. Studies of flying snakes have led researchers to posit the mean undulation rate is 1.4 Hz.  You wish to test this hypothesis using the small sample of undulation rates shown below.  Create a small dataset of the paradise tree snake undulation rates and choose and justify a test you can use to assess the data.  
Undulation rates (in Hz):  0.9, 1.4, 1.2, 1.2, 1.3, 2.0, 1.4, 1.6

6.  Using data from Australian athletes (http://www.statsci.org/data/oz/ais.html 
for details), determine if the average male training at the Australian Institute 
of Sport differs in weight from the average Australian male (85.9 kg) using
bootstrapping techniques. Data at 
```{r}
sport <- read.table("http://www.statsci.org/data/oz/ais.txt", header = T)
```

7. The following data are human blood clotting times (in minutes) of individuals
given one of two different drugs. Test the hypothesis that the mean clotting times
are equal for the two groups
*  Estimating the variance from the data 
* Using rank transform analysis 
* Using a permutation test 
* Using a bootstrap test 

|Drug B | Drug G
|------ |-------
|8.8 | 9.9
|8.4 | 9.0
|7.9 | 11.1
|8.7 | 9.6
|9.1 | 8.7
|9.6 | 10.4 
|    | 9.5

8. Data on plant heights (in cm) for plants grown with a new and old formulation 
of fertilizer can be found at
https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/fertilizer.csv . 
Use the data to test the hypothesis that there is no difference in mean plant
heights for the two groups 

*  Estimating the variance from the data 
* Using rank transform analysis 
* Using a permutation test 
* Using a bootstrap test 

Note you'll need to cast this dataframe into a dataframe with a column for 
old and new outcomes.  You’ll have to add a unique identifier. Try something like

```{r, eval=FALSE}
fertilizer$ID <- 1:nrow(fertilizer)
fertilizer_wide <- dcast(fertilizer, ID~fertilizer, value.var = "height")
```


9. Metabolic rates were measured before and after administering a drug. Data is
available at 

https://raw.github.com/jsgosnell/CUNY-BioStats/blob/master/datasets/metabolic_rates.csv

Test the hypothesis that the drug has no effect on metabolic rate with using 

*  a Student's t test
* a rank test
* a binary test
* bootstrapping


Note you'll need to melt this dataframe into one with a single “time” 
column (should have old or new as a value) and a “rate” column.

10. Endurance of the above animals was also measured. Data is available at

https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/endurance.csv

Test the hypothesis that the drug has no effect on using 
*  a Student's t test
* a rank test
* a binary test
* bootstrapping

## For the following problems, make sure you can justify the test you choose to use and carry it out

11. Using data from Australian athletes (http://www.statsci.org/data/oz/ais.html 
for details), determine if the average male training at the Australian Institute 
of Sport differs in from the average female training at the site in mercury
concentration (Hg). 

12. Laureysens et al. (2004) measured metal content in the wood of 13 poplar 
clones growing in a polluted area, once in August and once in November.
Concentrations of aluminum (in micrograms of Al per gram of wood) are shown 
below. (From Handbook of Biological Statistics). Has metal content changed over
time?

| Clone | August | November
|------| ---|---
|Columbia River | 18.3 |12.7
|Fritiz Pauley | 13.3 |11.1
|Hazendas | 16.5 |15.3
|Primo | 12.6 | 12.7
|Raspalje | 9.5 |10.5
|Hoogvorst |13.6 |15.6
|Balsam Spire | 8.1 |11.2
|Gibecq   |8.9 |14.2
|Beaupre | 10.0 |16.3
|Unal |8.3 |15.5
|Trichobel |7.9 | 19.9
|Gaver | 8.1 | 20.4
|Wolterson | 13.4 |36.8

13. Data was collected on cholesterol levels following heart attacks.  Data
available at 
http://www.statsci.org/data/general/cholestg.txt 

Note to read a .txt file you use,
read.table("http://www.statsci.org/data/general/cholestg.txt", header = T)
Check the defaults to understand the argument.  
More info on the data is @
http://www.statsci.org/data/general/cholest.html
Is there evidence cholesterol levels changed between day 2 and day 14?  
**THINK ABOUT RELABELLING DAY AS A FACTOR (WHY?)?**

## Using ggplot2
Let’s return to the mammal sleep dataset that we left off with last week. 
Load the dataset
```{r}
sleep <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/sleep.csv")

```

Last time you used the built-in plot functions to do some plots. Let’s replace 
those with ggplot2 and do some more.

14. First plot how TotalSleep is explained by BrainWt (remember the issues with 
the data).  Use ggplot2 to plot the relationship.

15. Next color code each plot point by whether or not its a primate.  In order 
to do this you can use the Primate column or (following class code) make a new 
column called Taxa to represent the information (hint:search for “ revalue”). 
Make sure axes are well-labeled.

16. Let’s work with histograms.
* What type of variation do we see in total time spent sleeping? Create a 
histogram to explore this issue.
* Facet the graph you created based on whether or not the animal is a primate 
(Primate column).
* Now only graph the data for primates.

17. Develop a properly-labelled bar graph with error bars to explore how
* total 
sleep changes with Primate (relabeled as yes/no as Primate/Non-Primate; note 
there are multiple ways to do this!) – use a 95% confidence interval for the bar
* Predation risk (as a factor!) – use 1 standard error for the bar. Note the difference!


