---
title: "2023 Fall 78001-2 Final: Snakes on an exam!"
subtitle:  "Thanks for a good semester"
author: "jsg answers"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
    toc_depth: 6
bibliography: references.bib
---

# Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and those you produced in class) - you can even look
for help with R-specific issues! However, you may not get external help
(from other students, directed internet searches re, etc.) regarding the
data. Please update the statement below to acknowledge these
instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students.

Good luck!

# Snakes

Understanding how environmental factors impact species can be important
to conservation and management (and relates directly to exploring how
factors impact outcomes - analysis!). A colleague is focused on how
various factors impact snake behavior, habitat use, and species
interactions for a species that lives in semi-arid deserts that are occasionally
impacted by heavy (monsoon) rains. They've collected a lot of data but aren't sure how to
analyze it.

## 1

From [@blais2023]; I downloaded data and

-   averaged environmental variable start and end time

    -   ambient temperature (TA,°C)

    -   relative humidity (RH)

    -   barometic pressure (BP, millibars)

    -   wind (wind speed, m/s)

-   only included a subset for student use

For one study, the researcher used visual surveys (eg, walking through
the area) to count snakes in a local reserve. For each survey, they
documented

-   were snakes observed (column in dataset: pres , 1=yes, 0= no)

-   season (column in dataset: season, pre-monsoon, monsoon,
    post-monsoon)

-   ambient temperature (column in dataset: TA,°C)

-   relative humidity (column in dataset: RH)

-   barometic pressure (column in dataset: BP, millibars)

-   wind speed(column in dataset: wind , m/s)

-   stream flow (column in dataset: streamflow..m3., max stream flow
    (m3/s) during survey window)

They tried to use a linear model to determine which factors were
important to determing snake presence. They haven't had stats in a
while, but they found some code online. They started by building a model
with all factors, checking the p-values, and checking their assumptions.
However, this led to some issues.

-   season had 3 levels - only 2 are listed

    ```{r}
snakes <- read.csv ("https://docs.google.com/spreadsheets/d/e/2PACX-1vR1L-2LEdYMV-HNo3b6aexfT2DyNfC7PhujoNQ33Y6gIBFIPrrWtvdKIUz146or6q43QI05gv0KR9Fa/pub?gid=1499758404&single=true&output=csv", 
                        stringsAsFactors = T, na.strings = c("NA", "#DIV/0!"))
    ```


    ```{r}
    str(snakes)
    panel.cor <- function(x, y, digits=2, prefix="", cex.cor) 
{
  usr <- par("usr"); on.exit(par(usr)) 
  par(usr = c(0, 1, 0, 1)) 
  r <- abs(cor(x, y)) 
  txt <- format(c(r, 0.123456789), digits=digits)[1] 
  txt <- paste(prefix, txt, sep="") 
  if(missing(cex.cor)) cex <- 0.8/strwidth(txt) 
  
  test <- cor.test(x,y) 
  # borrowed from printCoefmat
  Signif <- symnum(test$p.value, corr = FALSE, na = FALSE, 
                   cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                   symbols = c("***", "**", "*", ".", " ")) 
  
  text(0.5, 0.5, txt, cex = cex * r) 
  text(.8, .8, Signif, cex=cex, col=2) 
}
pairs(snakes, lower.panel=panel.smooth, upper.panel=panel.cor)
library(psych)
pairs.panels(snakes, stars=T)
    ```

    ```{r}
    snake_presence_lm <- lm(pres ~ season+TA+RH+BP+wind+streamflow..m3.,
                          data= na.omit(snakes))
    summary(snake_presence_lm)
    
    ```

-   Some of these plots looked odd (note the na.omit(snakes) portion is
    just removing all rows with an NA - use it as needed)

    ```{r}
    plot(snake_presence_lm)
    
    ```

Do you have any advice to offer on how they may want to analyze the
data? Make sure you justify your approach and explain any needed steps; also list
hypotheses if needed.


### Answer

```{r}
library(psych)
pairs.panels(snakes, stars=T)
```


```{r}
library(MASS)
snake_presence <- glm(pres ~ season+TA+RH+BP+wind+streamflow..m3.,
                      data= na.omit(snakes),
                      family = "binomial")
summary(snake_presence)
library(car)
Anova(snake_presence, type = "III")
```

Dispersion estimate looks ok. Could stop here (one option) or move forward 
with model selection. Top-down using AIC shown here

```{r}
stepAIC(snake_presence)
```

Note "perfect prediction" issues, so we get large standard errors.  



## 2

For the survey, they also noted they weren't sure how to specify season. 
Although they originally specified it as a factor with 3 levels based on date 
(pre-monsoon, monsoon, post-monsoon),they could also just specify it as wet or dry.
They know the 2 measures are related based on the following table:

```{r}
seasons <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vR1L-2LEdYMV-HNo3b6aexfT2DyNfC7PhujoNQ33Y6gIBFIPrrWtvdKIUz146or6q43QI05gv0KR9Fa/pub?gid=1603572731&single=true&output=csv",
                    stringsAsFactors = T, na.strings = c("NA", "#DIV/0!"))
```

```{r}
table(seasons$season.2, seasons$season)
```
Are there any issues with including both variables (labeled season and season.2
in the dataset) in the model? If so, how would you choose which one (if either)
to include?


### Answer
```{r}
chisq.test(table(seasons$season.2, seasons$season))
```





```{r}
AIC(glm(pres~season.2, family="binomial", data=seasons))
AIC(glm(pres~season, family="binomial", data=seasons))
AIC(glm(pres~season+season.2, family="binomial", data=seasons))

AIC(glm(pres~season.2, family="binomial", data=seasons),glm(pres~season, family="binomial", data=seasons), glm(pres~season+season.2, family="binomial", data=seasons))


AIC(lm(pres~season.2,data=seasons))
AIC(lm(pres~season,  data=seasons))
AIC(lm(pres~season+season.2,  data=seasons))


```

```{r}
seasons$pres2 <- seasons$pres
levels(seasons$pres2)<- c("no", "yes")
seasons_full<- lm(pres ~ season + season.2, data= na.omit(seasons)) 
seasons_final <- step(seasons_full)
AIC(seasons_final, lm(pres~season,  data=seasons))
```


## 3

As part of their surveys, they also collected morphometric data including

-   snout-vent length (SVL, mm)
-   mass (mass, g)
-   age (age, factor labelled as neonate, immature, or adult)

A common measure of body condition in their field is log(SVL)/log(M). They are 
curious if this condition index (focus on this for now) differs among age.  
How  would you recommend they
analyze the data? Make sure you justify your approach and explain any needed steps; also list
hypotheses if needed.

```{r}
size <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vR1L-2LEdYMV-HNo3b6aexfT2DyNfC7PhujoNQ33Y6gIBFIPrrWtvdKIUz146or6q43QI05gv0KR9Fa/pub?gid=276877765&single=true&output=csv",
                 stringsAsFactors = T)
```

-   snout-vent length (SVL, mm)
-   mass (mass, g)
-   age (age, labeled as neonate ())

### Answer

```{r}
condition <- lm(logM.logSVL~age, size)
plot(condition)
Anova(condition, type="III")
library(multcomp)
summary(glht(condition, linfct = mcp(age = "Tukey")))
```

## 4

They are curious why  condition index is presented in such a manner (why the logs?). Can you
offer any insight or recommend any other way you could analyze the data?  Note this
is a shorter question (fewer points), so I'm just looking for some general ideas 
they could use to get started.

## 5

Plot the data on condition index and age.

### Answer

```{r}
library(ggplot2)
ggplot(size, aes(x=age, y=logM.logSVL))+
  geom_jitter()
```

## 6

Given that age classes may respond diffently to stressors, the researcher
also wanted to know if snake movement (defined as 0=observed snake not moving;
1=observed snake moving) differed among ages classes.  They share this table:  

```{r, include=F}
behavior <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vR1L-2LEdYMV-HNo3b6aexfT2DyNfC7PhujoNQ33Y6gIBFIPrrWtvdKIUz146or6q43QI05gv0KR9Fa/pub?gid=1276145787&single=true&output=csv",
                     stringsAsFactors = T)
behavior_table <- table(behavior$movement, behavior$ageclass)
```

```{r}
behavior_table
```
Do you have any advice to offer on how they may want to analyze the
data? Make sure you justify your approach and explain any needed steps; also list
hypotheses if needed. 

### Answer
```{r}
chisq.test(behavior_table)
```



```{r}
chisq.test(behavior_table)$expected
```

```{r}
fisher.test(behavior_table)
```
note

```{r}
as_glm <- glm(movement~ageclass, behavior, family = "binomial")
summary(as_glm)
Anova(as_glm, type="III")
chisq.test(behavior_table, correct = F)
```


## 7

Based on @cyriac2019

In a fun/odd twist, they were also curious about how snakes interacted with
their predators. Although many other snakes in use camouflage (brown or black 
color), this snake has several colorful versions (red and yellow)  The researchers
built snake models in these colors out of play dough, plus an orange morph, to 
determine how the bright color impacts predation rates (measured as bird strikes
on the models). They put 50 models of each color in 4 different areas and checked the
models for bird strikes after 48 hours.  All models were collected.

Data includes

* Model (representing color of model: Bl(ack), Br(own), N(orange, novel color),
R(ed), Y(ellow))
* Avian.Attack (evidence of bird attack? 0=no, 1 = yes)
* Transect (location of model)

Do you have any advice to offer on how they may want to analyze the
data? Make sure you justify your approach and explain any needed steps; also list
hypotheses if needed. 


```{r}
color <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQHWuORlDpmKhPsHDWSsGrwlzK66EqP2-GGIikhljBgsRF0nNsqbcFV5_c6JdN6sBAF55-XX4oAZJlc/pub?gid=187879458&single=true&output=csv", stringsAsFactors = T)
```


### Answer
```{r}
str(color)
```

```{r}
color_lm <- lm(Avian.Attack ~ Model + Transect, color)
summary(color_lm)
plot(color_lm)
Anova(color_lm, type="III")
```


```{r}
library(lme4)
color_glmm <- glmer(Avian.Attack ~ Model + (1|Transect), family="binomial", color)
dispersion_glmer <- function(modelglmer){
  n <- length(residuals(modelglmer))
  return(sqrt(sum(c(residuals(modelglmer), modelglmer@u)^2)/n))
}
dispersion_glmer(color_glmm)
```


```{r}
summary(color_glmm)
Anova(color_glmm, type="III")
```

```{r}
summary(glht(color_glmm, linfct = mcp(Model = "Tukey")))
```

or

```{r}
color_table <- aggregate(color, Avian.Attack~Model + Transect, sum)
color_table$not_attacked <- 50 - color_table$Avian.Attack
color_table
```

```{r}
color_glmm_p <- glmer(Avian.Attack ~ Model + (1|Transect), family="poisson", color_table)
summary(color_glmm_p)
Anova(color_glmm_p, type="III")
summary(glht(color_glmm_p, linfct = mcp(Model = "Tukey")))
```

```{r}
color_glmm_prop <- glmer(cbind(Avian.Attack, not_attacked) ~ Model + (1|Transect), family="binomial", color_table)
summary(color_glmm_prop)
Anova(color_glmm_prop, type="III")
```


```{r}
attack <- color$Avian.Attack == 1
nonattack <- color$Avian.Attack == 0
attack_presence_lm <- glmer(cbind(attack, nonattack) ~ Model + (1|Transect), data= color, family = 'binomial')

summary(attack_presence_lm)
```

