---
title: "2023 Fall 78001-2 Midterm"
subtitle:  "Happy (late) Halloween"
author: "jsg answers"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
    toc_depth: 6
---

# Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and those you produced in class) - you can even look
for help with R-specific issues! However, you may not get external help
( from other students, directed internet searches re, etc.) regarding
the dataa. Please update the statement below to acknowledge these
instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students.

Good luck!

# Oysters, predators, and climate change

## Background

Oysters are bivalves (molluscs with 2 shells) that grow in cluster we
call reefs as adults.  Oysters are susceptible to mortality due to both
biotic (predator) and abiotic (heat stress) cues. These factors may also
impact growth.

A study was undertaken to consider these issues in South Carolina. The
focal site had been previously surveyed in 1980 through a set of
undergraduate projects.

## #1

Following up on the 78001 exam, conduct the analysis you recommended for
assessing oyster mortality. Remember:

> Oyster traits were assessed using a random sample. A drone was used to
> capture high-resolution aerial images of the site. These images were
> stitched together using photoshop software, and 1000 points were
> randomly placed on the resulting image. The oyster that was closest to
> each dot was then measured for status (dead/alive) and length.

> Previous studies had noted approximately 15% of oysters on the reef
> were dead. Our study found 155 dead oysters.

```{r}
binom.test(155,1000, .13)
```

### interpretation

## #2

## contingency

Unlike the previous study

```{r}
set.seed(21)
x <- rmultinom(n = 1, size = 500, prob = c(.4, .4, .4, .6, .6, .6, .8))
y <- rmultinom(n = 1, size = 500, prob = c( .6, .6, .6, .4, .4, .4,.2))
colSums(x+y)

oyster_table <- matrix(c(x,y), nrow=2, byrow = T)
colnames(oyster_table) <- c("<15 mm","15-30 mm", "30-45 mm",
                            "45-60 mm", "60-90 mm", "90-120 mm", "120+mm")
rownames(oyster_table) <- c("Alive", "Dead")
oyster_table

chisq.test(oyster_table)

library(rcompanion)
bonf_correct <- pairwiseNominalIndependence(oyster_table, compare="col", method = "holm")
bonf_correct
```

## #3

```{r}
set.seed(19)
x <- runif(10, 1,5)

reef_wide <- data.frame(reef = rep(1:10),
                   old_survey_predator_density = x,
                   new_predator_density = x*1.15)

reef_wide$change <- reef_wide$new_predator_density-
  reef_wide$old_survey_predator_density


reef_long <- data.frame(reef = rep(1:10,2),
                        measure = factor(c(rep("old_survey", 10),
                                         rep("new_survey", 10))),
                        predator_density = c(x, x*1.15))
write.csv(reef_wide, "reef_wide.csv", row.names = F)
write.csv(reef_long, "reef_long.csv", row.names = F)
```

```{r}
reef_long <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSTXRWNfzYrCmCnEYXoeGdgD-uhaXph4XFNwzZdMOTuk9sLUtWVVp91YaBcATCfoAQ4rNgfE4nar5Yj/pub?gid=2007431006&single=true&output=csv", stringsAsFactors = T)
reef_wide <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQK9toq6hztnK3E5roPGO66eofkrjhhnIHfM29UDPXuhQ6w4cAIuswfGsuVBTG7B7dcJ6S2utEQsKZw/pub?gid=1962481186&single=true&output=csv", stringsAsFactors = T)
```

```{r}
t.test(reef_wide$old_survey_predator_density, reef_wide$new_predator_density)
t.test(reef_wide$old_survey_predator_density, reef_wide$new_predator_density,paired = T)
```

```{r}
reef_long_approach <- lm(predator_density~factor(reef)+measure, reef_long)
plot(reef_long_approach)
Anova(reef_long_approach, type = "III")
```

```{r}
reef_colleague <- lm(change~reef, reef_wide)
Anova(reef_colleague, type = "III")
summary(reef_colleague)
```

```{r}
ggplot(reef_wide, aes(x=change))+
       geom_histogram()
```

## #3

## factorial

An experiment was carried out

```{r}
set.seed(19)
experiment <- data.frame(temperature = factor(c(rep("ambient",75), rep("elevated_scenario1", 75), rep("elevated_scenario2", 75))),
                         predator_cue = factor(c(rep(c(rep("none",25),
                                               rep("normal",25),
                                               rep("high",25)),3))),
                         growth = c(rnorm(25,2),
                                    rnorm(25, 1.5),
                                    rnorm(25, .7),
                                    rnorm(25,2.7),
                                    rnorm(25,1.0),
                                    rnorm(25, .4),
                                    rnorm(25, 3.1),
                                    rnorm(25, 3.1),
                                    rnorm(25, 3.1)))
experiment[experiment$growth<0, "growth"] <- 0
write.csv(experiment, "experiment.csv", row.names = F)
```


```{r}
experiment <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRXgAuynB8tusTIkEc93S0Az-dNBvWC8hxj3co4N2907P9wqbmnYBV8SK0GVYaY7ILUdDXQQngVs_Hn/pub?gid=157339363&single=true&output=csv", stringsAsFactors = T)

experiment$temperature <- factor(experiment$temperature, levels=c('ambient', 'elevated_scenario1', "elevated_scenario2"))

experiment$predator_cue <- factor(experiment$predator_cue, levels=c('none', 'normal', "high"))



growth_lm <- lm(growth~predator_cue*temperature, experiment)
plot(growth_lm)
library(car)
Anova(growth_lm, type="III")
```

## #4

## plot

```{r}
library(ggplot2)
library(Rmisc)
experiment_summary <- summarySE(experiment, measurevar = "growth", groupvars = c("temperature", "predator_cue"))
ggplot(experiment_summary, aes(x=temperature, 
                           y=growth,
                           fill=predator_cue)) +
  geom_col(color="black", position=position_dodge()) +
  geom_errorbar(aes(ymin=growth, ymax=growth+ci), position = position_dodge()) +
  labs(title="Impact of increasing predator cue depends on temperature",
       x= "Temperature",
       y= "growth (cm)",
       fill = "Predator cue level")
```
