---
title: "2023 Fall 78002 Midterm"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
    toc_depth: 6
---

# Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and those you produced in class) - you can even look
for help with R-specific issues! However, you may not get external help
( from other students, directed internet searches re, etc.) regarding
the dataa. Please update the statement below to acknowledge these
instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students.

Good luck!

# Oysters, predators, and climate change

## Background

Oysters are bivalves (molluscs with 2 shells) that grow in clusters we
call reefs as adults.  Oysters are susceptible to mortality due to both
biotic (predator) and abiotic (heat stress) cues. These factors may also
impact growth.

A colleague who knows you are taking statistics wants your help.  They
took a course a long time ago but some things are hazy. Your colleague's
work focuses on how increased temperatures and predators impact growth
in oysters.

## #1

Following up on the 78001 exam, conduct the analysis you recommended for
assessing oyster mortality. Remember:

> Their first question focuses on work from one reef.  Oysters at the
> reef were assessed using a random sampling protocol. A drone was used
> to capture high-resolution aerial images of the site. These images
> were stitched together using photoshop software, and 1000 points were
> randomly placed on the resulting image. The oyster that was closest to
> each dot was then measured for status (dead/alive) and length.
>
> 155 of the surveyed oysters were dead.  Previous studies had noted
> approximately 15% of oysters on the reef were dead.  

Make sure you include

-   null hypothesis (1 pt)
-   alternative hypothesis (1 pt)
-   explanation for test you will use (1 pt)
-   results from statistical test (1 pt)
-   clear explanation of how results relate to your stated hypotheses (2
    pt)

## #2

Following up on the 78001 exam, conduct the analysis you recommended for
assessing oyster mortality given additional age class data. Remember:

> Given your colleagues ability to measure mortality and length, they
> are curious if there is anything else they can do with the data. 
> Below is a sample from a different reef.  

|       |                |       |       |       |       |        |      |
|-------|----------------|-------|-------|-------|-------|--------|------|
|       | size range(mm) |       |       |       |       |        |      |
|       | \<15           | 15-30 | 30-45 | 45-60 | 60-90 | 90-120 | 120+ |
| Alive | 60             | 54    | 62    | 74    | 69    | 81     | 100  |
| Dead  | 104            | 95    | 81    | 61    | 77    | 52     | 30   |

Make sure you include

-   null hypothesis (1 pt)
-   alternative hypothesis (1 pt)
-   explanation for test you will use (1 pt)
-   results from statistical test (1 pt)
-   clear explanation of how results relate to your stated hypotheses (2
    pt)

## #3

Following up on the 78001 exam, conduct the analysis you recommended for
assessing changes in predator density. As a reminder:

> Besides measuring oysters, your colleague also measured predator
> density in order to determine if it may have changed.  They chose 10
> sites where predator (blue crab) density had been measured for
> undergraduate research projects in the 1980's and carried out new
> surveys at those sites using the same methods.  They felt better about
> working with numerical data, so this time they bring  you some
> preliminary analysis...
>
> They knew sites were paired or something, so they subtracted the old
> from the new density to get the change column. Densities are crabs/m2.

The data they used can be found in 2 formats using the inputs below.

```{r}
reef_long <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSTXRWNfzYrCmCnEYXoeGdgD-uhaXph4XFNwzZdMOTuk9sLUtWVVp91YaBcATCfoAQ4rNgfE4nar5Yj/pub?gid=2007431006&single=true&output=csv", stringsAsFactors = T)
reef_wide <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQK9toq6hztnK3E5roPGO66eofkrjhhnIHfM29UDPXuhQ6w4cAIuswfGsuVBTG7B7dcJ6S2utEQsKZw/pub?gid=1962481186&single=true&output=csv", stringsAsFactors = T)
```

and their initial analysis code (screenshots shown on test) is

```{r}
reef_colleague <- lm(change~reef, reef_wide)
library(car)
Anova(reef_colleague, type = "III")
summary(reef_colleague)
```

Make sure you include

-   null hypothesis (1 pt)
-   alternative hypothesis (1 pt)
-   explanation for test you will use (1 pt)
-   results from statistical test (1 pt)
-   clear explanation of how results relate to your stated hypotheses (2
    pt)

## #4

Following up on the 78001 exam, conduct the analysis you recommended for
assessing changes in oyster growth from the lab experiment. As a
reminder:

> Given the acknowledgement your colleague gave you at a conference
> presentation,  you are asked to help with a related manuscript. 
> Another lab member followed up on the work by carrying out a
> controlled lab experiment. Oysters were placed in individual tanks and
> held at normal summer (calculated using recent data) temperature or at
> temperatures predicted under 2 IPCC --- Intergovernmental Panel on
> Climate Change- scenarios. Oysters were exposed to predator cues by
> dropping in water from tanks with 0, low (.25/m2), or high (2/m2)
> predators. After two months changes in oyster shell length (growth)
> was measured.  Twenty-five oysters were measured for each treatment
> combination.  

Data can be downloaded using

```{r}
experiment <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRXgAuynB8tusTIkEc93S0Az-dNBvWC8hxj3co4N2907P9wqbmnYBV8SK0GVYaY7ILUdDXQQngVs_Hn/pub?gid=157339363&single=true&output=csv", stringsAsFactors = T)
```

-   null hypothesis (1 pt)
-   alternative hypothesis (1 pt)
-   explanation for test you will use (1 pt)
-   results from statistical test (1 pt)
-   clear explanation of how results relate to your stated hypotheses (2
    pt)

## #5

Plot the data from the lab experiment (#4).
