---
title: "13. Multivariate methods answers"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
```


1. A pharmaceutical company has a drug that may help an illness that causes 
fever (temperature in degrees Celsius), blood pressures, and “aches” (scored on an
index).  Data is collected for several patients. To determine if the drug actually 
helps, test for differences in multivariate means for the fever, 
pressure and aches column,
against the grouping variable treatment.

```{r}
illness <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRlcjpU0XHfXF1WId1C5ZYX0YdY53KI9Nv91_tNCMj4z4iTjr-XMW1L_Ln8j3ahk5GUPZy4kGzSlA96/pub?gid=1322236994&single=true&output=csv",
                    stringsAsFactors = T)
```

We will test the multiple outcomes using a MANOVA. This tests the null hypothesis
that there is no difference in the vector of mean parameters for the groups. 

```{r}
m <- manova(cbind(fever,pressure,aches)~treatment, illness)
summary(m)
```
The MANOVA shows a significant difference (Pillai's trace = .55466, p <.001), so
we reject the  null hypothesis.  To follow this up we consider ANOVAs for each 
trait to determine which ones differ among groups

```{r}
summary.aov(m)
```

These indicate that only fever differs among groups. Finally, a post-hoc test is used
to determine which levels differ for fever. Here I will rebuild model using lm for
illustration.

```{r}
fever <- lm(fever~treatment, illness)
library(car)
Anova(fever, type = "III")
library(multcomp)
comp_fever <- glht(fever, linfct = mcp(treatment = "Tukey"))
summary(comp_fever)
```

2. Darlingtonia californica is a partly carnivorous pitcher plant that grows in 
fens and along seeps and
streams in the mountains of Oregon and California. Its pitchers are tubular l
eaves with a round hood and
a mouth at the base of the hood (see figure below). A "fishtail" appendage 
hangs from the mouth.
Wasps and other prey are attracted to nectar secreted by extrafloral nectaries 
along the hood, mouth,
and fishtail. Plants absorb nutrients excreted by a food web of bacteria, 
protozoa, mites, and fly larvae
that break down the prey. 

Measurements of 87 plants from four sites were made by Ellison and Farnsworth 
(2005, The cost of
carnivory for Darlingtonia californica (Sarraceniaceae): evidence from 
relationships among leaf traits.
Am. J. Botany 92: 1085-1093). Their measurements are available using


```{r}
pitcher <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQZf2mS4NmfBUUsn7lY2RTpuVjuWvRYN4MdLNt2XdS4WepolrxvWCKBI5diKBMWPLhdbEGwP-hfWOnz/pub?gid=1427497144&single=true&output=csv",
                    stringsAsFactors = T)
```

I obtained them from the web page
(http://harvardforest.fas.harvard.edu/personnel/web/aellison/publications/primer/primer.html)
of A. M. Ellison for the book by Gotelli and Ellison (2004, A primer of ecological 
statistics. Sinauer,
Sunderland, Mass.). To simplify, outliers have been removed. Most plant traits 
in the file are illustrated in the image below, and trait labels are fairly 
self-explanatory. Keel width measures the span of the
pitcher tube. "Wing" traits refer to the lengths of the fishtail appendage.

<center>


![Photograph of a *Darlingtonia californica* pitcher with morphological 
measurements indicated (lower diameter at ground level not shown). Note the 
translucent hood and the pronounced fishtail appendage attached to the proximal 
side of the mouth](https://bsapubs.onlinelibrary.wiley.com/cms/asset/1258a675-69f2-4bca-80de-d76ed9c43c47/ajb21085-fig-0001-m.jpg){width=30%}

</center>

* Use a MANOVA to consider differences in plant traits (do not follow-up with 
almost 20 ANOVA's! Just consider why PCA might be useful with large datasets!
* Use principal component analysis to investigate variation among
individual plants in their dimensions. Along the way, make sure you
  * construct screeplots
  * determine how many principal components to retain (and why)
  * Use biplots and/or loadings to see if you can understand/interpret the 
  first few principal components
  
3. The data for this exercise are rodent species abundance from 28 sites
in California (Bolger et al. 1997, Response of rodents to habitat fragmentation 
in coastal Southern
California, Ecological Applications 7: 552–563). The file (in contingency table 
format) is on Blackboard
(rodent.csv). This data comes from the (website)[http://www.zoology.unimelb.edu.au/qkstats/data.htm) of Quinn and Keough (2002, Experimental
Design and Data Analysis for Biologists, Cambridge Univ. Press, Cambridge, UK).

