---
title: "2022 Spring 78002 Midterm"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
---

# Instructions! Read this first.

The exam is open note, open course website, open class-related code repositories 
(mine and those you produced in class). However, you may not get external help (
from other students, directed internet searches, etc.).  Please update the statement
below to acknowledge these instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam. I 
understand any academic issues will result in a D or F on the exam or in the class
and be reported to the Dean of Students.

The exam is due by 12 PM on Monday, 4/4/2022.  

As often happens, your excellent statistical advice (provided via the 78001 exam) 
has led to more calls for help. The below questions build on the earlier exam.

# Parthenogenesis

1. The colleague working on parthenogenesis got a grant to get a  larger sample
(you probably noted having only one success made estimating proportions hard!).  
The second sample led to a sample of 500 snakes! While inspecting the samples, 
your colleague realized the sample actually contained 2 cryptic species (species
that look very similar and  may be confused).  From the sample, she obtained 225
specimens of species A, of which 15 were born via parthenogenesis.  Of the 275 
specimens of species B she collected, she found 32 individuals born via 
parthenogenesis.

Help her analyze the new data to investigate if parthenogenesis rates differ 
between the 2 species.  Make sure you include

* null hypothesis
* alternative hypothesis
* explanation for test you will use 
* results from statistical test
* clear explanation of how results relate to your stated hypotheses

# Fear

2. The data on mice heart rates is available via

```{r}
fear <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRE3OEf3EaR0uKgrqB6rZ8DxqpiNa5Z__MoAvN1iabF6Yp2fcLQbXFAXVfilAwY9TTYno786SstYCVp/pub?gid=1908952046&single=true&output=csv", stringsAsFactors = T)
```


Analyze it using an appropriate option to consider differences in heart rates.  
Make sure you include

* null hypothesis
* alternative hypothesis
* explanation for test you will use 
* results from statistical test
* clear explanation of how results relate to your stated hypotheses


# Salamanders

The data on salamander heart beats is available via 

```{r}
sallies <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSJHbPnebpKq58mE4kQWF9mQ_0ak2sTyaUbEvtmgoSkb31DprOmAS9clOoNNHD_3tHITOXKdhMbWfBf/pub?gid=1050993814&single=true&output=csv", stringsAsFactors = T)
```

3. Analyze the data correctly to consider how stage and morphotype impact heart 
rate (hopefully following the instructions you gave the 
authors!). Make sure you include

* null hypothesis
* alternative hypothesis
* explanation for test you will use 
* results from statistical test
* clear explanation of how results relate to your stated hypotheses

4. Create a properly-labelled plot of the data. There are many options here, but
make sure you show trends and patterns that match your analysis.


5. The data on temperature impacts on heart rate (the follow-up study) 
is available @ 


```{r}
temperature <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRhcbtbwuBje-9I7abkP2EXWuddGVo5VH1N2CgNtR8NMj-sjS-9KXjrJuZY9an9Gt4c-JbdCwwP9S0B/pub?gid=1230482827&single=true&output=csv", stringsAsFactors = T)
```


Analyze the data correctly (hopefully following the instructions you gave the 
authors!). Make sure you include

* null hypothesis
* alternative hypothesis
* explanation for test you will use 
* results from statistical test
* clear explanation of how results relate to your stated hypotheses


6. Create a properly-labelled plot of the data. There are many options here, but
make sure you show trends and patterns that match your analysis.



