---
title: "2022 Spring 78002 Midterm answers"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
---

# Instructions! Read this first.

The exam is open note, open course website, open class-related code repositories 
(mine and those you produced in class). However, you may not get external help (
from other students, directed internet searches, etc.).  Please update the statement
below to acknowledge these instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam. I 
understand any academic issues will result in a D or F on the exam or in the class
and be reported to the Dean of Students.

The exam is due by 12 PM on Monday, 4/4/2022.  

As often happens, your excellent statistical advice (provided via the 78001 exam) 
has led to more calls for help. The below questions build on the earlier exam.

# Parthenogenesis

##1. 

The colleague working on parthenogenesis got a grant to get a  larger sample
(you probably noted having only one success made estimating proportions hard!).  
The second sample led to a sample of 500 snakes! While inspecting the samples, 
your colleague realized the sample actually contained 2 cryptic species (species
that look very similar and  may be confused).  From the sample, she obtained 225
specimens of species A, of which 15 were born via parthenogenesis.  Of the 275 
specimens of species B she collected, she found 32 individuals born via 
parthenogenesis.

Help her analyze the new data to investigate if parthenogenesis rates differ 
between the 2 species.  Make sure you include

* null hypothesis
* alternative hypothesis
* explanation for test you will use 

```{r}
parthenogenesis <- matrix(c(15, 225-15, 32, 275-32), ncol = 2, byrow = T)
chisq.test(parthenogenesis)
#note same as
parthenogenesis <- matrix(c(15, 225-15, 32, 275-32), ncol = 2, byrow = F)
chisq.test(parthenogenesis)
```


* results from statistical test
* clear explanation of how results relate to your stated hypotheses

# Fear

Create data, make sure it works, then save for future use and no longer run this.

```{r, eval=F}
library(ggplot2)
library(EnvStats)
# Rigth skewed
a <- rlnormTruncAlt(50, 85, 3, 60, 100)
b <- rlnormTruncAlt(50, 110, 3.2, 80, 130)
fear <- data.frame(treatment = c(rep("control", 50), rep("owl", 50)),
                   heart_rate = c(a,b))
fear <- write.csv(fear, "fear.csv", row.names = F)
```

##2. 

The data on mice heart rates is available via

```{r}
fear <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRE3OEf3EaR0uKgrqB6rZ8DxqpiNa5Z__MoAvN1iabF6Yp2fcLQbXFAXVfilAwY9TTYno786SstYCVp/pub?gid=1908952046&single=true&output=csv", stringsAsFactors = T)
```


Analyze it using an appropriate option to consider differences in heart rates.  
Make sure you include

* null hypothesis
* alternative hypothesis
* explanation for test you will use 
* results from statistical test

Options for analyzing include t-test, bootstrapping, permutation. Sign isn't possible (not paired), and Wilcox makes less sense as not same shape.

```{r}
t.test(heart_rate~treatment,fear)
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/bootstrapjsg.R")
bootstrapjsg(fear[fear$treatment == "control", "heart_rate"], fear[fear$treatment == "owl", "heart_rate"])
library(coin)
independence_test(heart_rate~treatment,fear)
```


* clear explanation of how results relate to your stated hypotheses


# Salamanders

## Data creation

```{r, eval = F}
sallies <- data.frame(morphotype = c(rep(1,60), rep(2,60), rep(3,60)),
                      stage = rep(c(rep("juvenile", 30), rep("adult", 30)),3),
                      tail_length  = c(rnorm(30, 25,2), rnorm(30, 28,2),
                                       rnorm(30,25.2,2.1),rnorm(30,28.1,2.1),
                                       rnorm(30, 28,2.2), rnorm(30,28.2,2.1)))
write.csv(sallies, "sallies.csv", row.names = F)
```

The data on salamander heart beats is available via 

```{r}
sallies <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSJHbPnebpKq58mE4kQWF9mQ_0ak2sTyaUbEvtmgoSkb31DprOmAS9clOoNNHD_3tHITOXKdhMbWfBf/pub?gid=1050993814&single=true&output=csv", stringsAsFactors = T)
```

##3. 

Analyze the data correctly to consider how stage and morphotype impact heart 
rate (hopefully following the instructions you gave the 
authors!). Make sure you include

* null hypothesis
* alternative hypothesis
* explanation for test you will use 

```{r}
sallies_lm_morphotype <- lm(tail_length~morphotype, sallies)
plot(sallies_lm_morphotype)
summary(sallies_lm_morphotype)
library(car)
Anova(sallies_lm_morphotype, type = "III")
sallies_lm_stage <- lm(tail_length~stage, sallies)
plot(sallies_lm_stage)
summary(sallies_lm_stage)
Anova(sallies_lm_stage, type = "III")
sallies$morphotype <- as.factor(sallies$morphotype)
sallies$stage <- relevel(sallies$stage, "juvenile")
sallies_lm_int <- lm(tail_length~stage*morphotype, sallies)
plot(sallies_lm_int)
summary(sallies_lm_int)
Anova(sallies_lm_int, type = "III")
```

break up

* results from statistical test
* clear explanation of how results relate to your stated hypotheses

##4.

Create a properly-labelled plot of the data. There are many options here, but
make sure you show trends and patterns that match your analysis.

```{r}
library(Rmisc)
sallies_summary <- summarySE(sallies, measurevar="tail_length", groupvars =
                               c("morphotype", "stage"), na.rm = T)
ggplot(sallies_summary, aes(x=morphotype, y=tail_length, color=stage, 
                                   shape = stage)) +
  geom_point(size = 3) +
  geom_line(aes(group=stage, linetype =stage), size=2) +
  geom_errorbar(aes(ymin=tail_length-ci, ymax=tail_length+ci), size=1.5) +
  ylab("Tail Length")+ 
  xlab("Morphotype") + 
  scale_shape_discrete(guide=FALSE)+
  scale_linetype_discrete(guide=FALSE)+
  ggtitle("Tail length across stages of various morphotypes")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
ggplot(sallies_summary, aes(x=stage, y=tail_length, color=morphotype, 
                                   shape = morphotype)) +
  geom_point(size = 3) +
  geom_line(aes(group=morphotype, linetype =morphotype), size=2) +
  geom_errorbar(aes(ymin=tail_length-ci, ymax=tail_length+ci), size=1.5) +
  ylab("Tail Length")+ 
  xlab("stage") + 
  scale_shape_discrete(guide=FALSE)+
  scale_linetype_discrete(guide=FALSE)+
  ggtitle("Tail length across stages of various morphotypes")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
#or prettier
```


##5.

## Data creation

```{r, eval=F}
temp <- runif(82,15,27)
heart_rate <- temp*1.5+rnorm(82,0,2)
temperature <- data.frame(temp, heart_rate)
write.csv(temperature, "temperature.csv", row.names = F)
```


The data on temperature impacts on heart rate (the follow-up study) 
is available @ 


```{r}
temperature <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRhcbtbwuBje-9I7abkP2EXWuddGVo5VH1N2CgNtR8NMj-sjS-9KXjrJuZY9an9Gt4c-JbdCwwP9S0B/pub?gid=1230482827&single=true&output=csv", stringsAsFactors = T)
```


Analyze the data correctly (hopefully following the instructions you gave the 
authors!). Make sure you include

* null hypothesis
* alternative hypothesis
* explanation for test you will use 
* results from statistical test

```{r}
relationship <- lm(heart_rate ~ temp, temperature)
plot(relationship)
summary(relationship)
```
* clear explanation of how results relate to your stated hypotheses


##6. 

Create a properly-labelled plot of the data. There are many options here, but
make sure you show trends and patterns that match your analysis.

```{r}
ggplot(relationship, aes(x=temp, y=heart_rate)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm") +
  ylab("Heart rate (bpm)")+ggtitle("Heart rate increases with temperature")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```

