---
title: "2022 Spring 78002 final"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
---

# Instructions! Read this first.

The exam is open note, open course website, open class-related code repositories 
(mine and those you produced in class). However, you may not get external help (
from other students, directed internet searches, etc.).  Please update the statement
below to acknowledge these instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam. I 
understand any academic issues will result in a D or F on the exam or in the class
and be reported to the Dean of Students.

The exam is due by 12 PM on Monday, 5/16/2022.

As often happens, your excellent statistical advice (provided via the 78001 exam) 
has led to more calls for help (or just being given the data to anlayze). 
The below questions build on the earlier exam.

# Fear effects

Following your advice, your colleague asked you to help analye the data. 
As a reminder:

Prey (tadpoles) were housed in tanks (20 per tank).  Prey in each tank were 
exposed to one of two treatments. Predator treatment tanks also housed a 
dragonfly naiad (a major predator of tadpoles)  in a small cylinder with mesh 
(so water was shared but the predator could not consume the tadpoles).  No 
predator treatment tanks housed an empty cylinder. The colleague measured the 
**length of each tadpole** and the **number of dead tadpoles** in each tank.   

##1.

Your colleagues tadpole length data is found below.  Help them analyze the data.

```{r}
fear <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSybsT7VO2rv24Rhi_WD5MBB3ww5RKXnqM2BveSP848vnldhPeFigBWPjxcY_WpvB1f1cd1PKQe1v3v/pub?gid=1960907558&single=true&output=csv",
                 stringsAsFactors = T)
```

Make sure you include

* null hypothesis
* alternative hypothesis
* explanation for test you will use 
* results from statistical test
* clear explanation of how results relate to your stated hypotheses

##2. 

Create a properly-labelled plot of the data. There are many options here, but
make sure you show trends and patterns that match your analysis.

##3. 

Your colleague provided you the data on tadpole mortality too.  

```{r}
fear_mortality <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQHLgiC6o2ezJPbEEcaU4Cb7Mz7F1-NMc8t5gMyEIMbLiPSDzslclIgRD51GefIBEl9gH7Fpylfbws5/pub?gid=2111555168&single=true&output=csv", stringsAsFactors = T)
```
Help them analyze the data. Make sure you include

* null hypothesis
* alternative hypothesis
* explanation for test you will use 
* results from statistical test
* clear explanation of how results relate to your stated hypotheses

##4. 

You are also given data from the followup study where length and width were 
analyzed in response to fear. The setup was the same (4 tanks for each treatment,
20 tadpoles per tank), but this time they recorded length and width of each 
tadpole. However, **they realized at the end of their work they were dealing with
a cryptic species!**, meaning half the tadpoles belong to species A and half to 
species B. They are not sure if this matters.

Data is shared below.  Help them analyze it.

```{r}
fear_lw <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSataGluosvnurZ4qOn_UTyHod85nbPTC-V9g5FkMZjKO6CEeRCn6CTAClrMf3Suxo_3kM8k-N2AQVD/pub?gid=1237391888&single=true&output=csv",
                    stringsAsFactors = T)
```

# Recruits

##5. 

The data on recruits is found below. As a reminder: 

The paper focused on predicting the number of new deer (fawns, density per km^2) produced at a 
nature reserve each season.  Potential covariates include adult abundance the 
previous season (density per km^2), rain the previous year (inches), and 
winter harshness (
ranked 1-5, least to most severe) were monitored over 8 years. 


```{r}
recruits <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vT4tICnyqUNSfWo4P3cMQ2DWxqe602IzdodBu5xoPKq43x5-u6vGpSDB_eXTvjFbHKQWwHHK9tkiJdL/pub?gid=1345393421&single=true&output=csv",
                     stringsAsFactors = T)
```

Analyze the data. Make sure you include

* null hypothesis
* alternative hypothesis
* explanation for test you will use 
* results from statistical test
* clear explanation of how results relate to your stated hypotheses


