---
title: "4. Continuous tests for 1 population"
author: "jsg answers"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
---

Before doing this, review the **Introduction to normality and one-sample tests for continous data** lecture set slides from 
https://sites.google.com/view/biostats/lessons/hypothesis-testing-with-the-binomial-distribution and
the  **4_normality.R** script in the lecture_files folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats).


You’ll also use the qqplot_example.R and
central_limit_theorem.R script from the 
[code_examples folder](https://github.com/jsgosnell/CUNY-BioStats/tree/master/code_examples).

Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  * **push** updates to github

## Recognizing and assessing normality 

1. Using the qqplot_example.R code, examine the following distributions and, for 
the continuous distributions (marked with a “*”), observe how a normal probability
plot (qqplot) can be used to visually test for approximate normality.

* *Normal (u= 0; σ^2^= 1, 10, 100) 
* *Student's t (df = 1, 10, 30, & 100) 
* *Chi-square (df= 1, 2, 5, 30, 50)
* Bernoulli (P=0.1, 0.5, & 0.9) 
* Binomial (P=0.05; N= 2, 5, 25, & 50); (P=0.25; N= 2, 5, 25, & 50); (P=0.50; 
N= 2, 5, 25, & 50); (P=0.75; N= 2, 5, 25, & 50); (P=0.95; N= 2, 5, 25, & 50) 
* Poisson ( u= 2, 5, 10, 30, & 50)

For this question, its easiest to just source the main file and see what happens.
When you source a script, it is run in R without showing any console output
(but graphs and objects are still produced!).  Try *source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/qqplot_example.R")*

```{r}
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/qqplot_example.R")
```

*Notice the spread of DATA of every distribution tend towards normality as sample size increases*

2. Review the central_limit_theorem.R code if you need to convince/remind 
yourself how common normality of means is for even non-normal data. You can source the code
using the same approach noted in Question 1.

*Here we are focused on how the means look as sample size increases*

```{r}
#make sure you have VGAM library installed
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/central_limit_theorem.R")
```


## Just for practice (note some sample sizes may be too small for these to all be good ideas!)

Make sure you are comfortable with null and alternative hypotheses for all examples. 
You should also feel comfortable graphing the data.

3. Seven observers were shown, for a brief period, a grill with 161 flies impaled 
and were asked to estimate the number. The results are given by Cochran (1954). 
Based on five estimates, they were 183.2, 149.0, 154.0, 167.2, 187.2, 158.0, and 143.0. Test the null hypothesis that the mean of the estimates is 161 flies.

* Assuming variance = 275


```{r}
flies <- c(183.2, 149.0, 154.0, 167.2, 187.2, 158.0, 143.0)
library(BSDA)
z.test(x=flies, mu = 161, sigma.x=sqrt(275))
```
*Using a z-test, I found a test statistics of z~=0.33 .This corresponds to a p-value of 0.73. This p value is >.05, so I fail to reject 
the null hypothesis that the mean of the estimates is 161 flies.*

*  Estimating the variance from the data
```{r}
t.test(x=flies,mu = 161)
```
*Using a t-test, which is appropriate when the variance must be estimated
from the sample and the means of the data may be assumed to follow a normal 
distribution,  I found a test statistics of t~6~=0.32.  This corresponds to a 
p-value of 0.76. This p-value is >.05, so I fail to reject 
the null hypothesis that the mean of the estimates is 161 flies.*

* Using rank transform analysis 

```{r}
wilcox.test(flies, mu=161)
```
*Using a Wilcoxon signed rank test, which is appropriate when normality assumptions 
can't be met and the distribution of the data appears to be symmetric, I found a test
statistics of V = 15 .This corresponds to a p-value of 0.94. This p-value is >.05, so I 
fail to reject the null hypothesis that the mean of the estimates is 161 flies.*

* Using binary transform analysis
```{r}
SIGN.test(flies, md=161)
```

*Using a sign test, which is appropriate when the data is continuous and other 
assumptions can't be met, I found a test statistics of s = 3 .This corresponds to a p-value of 1. This p-value is >.05, so I fail to reject 
the null hypothesis that the median (**Note change here**) of the estimates is
161 flies.*

Note there are several ways to load the data!  You can make a list (since the 
list is short):

```{r}
flies <- c(183.2, 149.0, 154.0, 167.2, 187.2, 158.0, 143.0 )
```

or make a dataframe in a spreadsheet software (eg, Excel, Google Sheets) and then upload using a read.csv command. We did this in the introduction to R!


4. Yields of 10 strawberry plants in a uniformity trial are given by Baker and 
Baker (1953) as 239, 176, 235, 217, 234, 216, 318, 190, 181, and 225 g. 
Test the hypothesis that µ = 205
* Assuming variance = 1500
```{r}
strawberries <- c(239, 176, 235, 217, 234, 216, 318, 190, 181, 225)
z.test(x=strawberries,mu = 205, sigma.x=sqrt(1500))
```
*Using a z-test, I found a test statistics of z=1.48.  This corresponds to a 
p-value of 0.14. This p-value is >.05, so I fail to reject 
the null hypothesis that the population mean is equal to 205.*

*  Estimating the variance from the data

```{r}
t.test(x=strawberries,mu = 205)
```
*Using a t-test, which is appropriate when the variance must be estimated
from the sample and the means of the data may be assumed to follow a normal distribution, I found a test statistics of t~9~=1.42.  This corresponds to a 
p-value of 0.19. This p-value is >.05, so I fail to reject 
the null hypothesis that the population mean is equal to 205.*

* Using rank transform analysis 
```{r}
wilcox.test(strawberries, mu=205)
```

*Using a Wilcoxon signed rank test, which is appropriate when normality assumptions 
can't be met and the distribution of the data appears to be symmetric, I found a test statistics of V=40.5.  This corresponds to a 
p-value of 0.20.  This p-value is >.05, so I fail to reject 
the null hypothesis that the population mean is equal to 205.*

* Using binary transform analysis
```{r}
SIGN.test(strawberries, md=205)
```

*Using a sign test, which is appropriate when the data is continuous and other 
assumptions can't be met,  I found a test statistics of s= 7.  This corresponds 
to a p-value of 0.34. This p-value is >.05,so I fail to reject 
the null hypothesis that the population median (**Note change here**) is equal to 205.*

4. Evolutionary geneticists predicts the family sex ratio will be 80% female in 
broods of eagles that successfully fledge >3 young. Nests that fledge 3 or more
chicks are very rare but a sample of 30 chicks are obtained from such nests and
they yield 25 females and 5 males. Test the hypotheses that that: 
* a) the sex ratio is 50% females

```{r}
#a
binom.test(25,30, p=.5)
```
*A binomial test was used as we are comparing an observed proportion against a 
set value.  Given a p-value of <.001, I reject the null hypothesis that the 
proportion of sons is equal to .5.*

* b) the sex ratio is 80% females.

```{r}
binom.test(25,30, .8)
```
*A binomial test was used as we are comparing an observed proportion against a 
set value.  Given a p-value of <.001, I fail to reject the null hypothesis that the 
proportion of sons is equal to .8.*

5. Studies of flying snakes have led researchers to posit the mean undulation 
rate is 1.4 Hz.  You wish to test this hypothesis using the small sample of 
undulation rates shown below.  Create a small dataset of the paradise tree snake 
undulation rates and choose and justify a test you can use to assess the data.

Undulation rates (in Hz):  0.9, 1.4, 1.2, 1.2, 1.3, 2.0, 1.4, 1.6

```{r}
snakes <- c(0.9, 1.4, 1.2, 1.2, 1.3, 2.0, 1.4, 1.6)
t.test(snakes, mu=1.4)
```
*Using a t-test, which is appropriate when the variance must be estimated
from the sample and the means of the data may be assumed to follow a normal
distribution,  I found a test statistics of t~7~=-.22.  This corresponds to a 
p-value of 0.83. This p-value is >.05, so I fail to reject 
the null hypothesis that the mean undulation rate is 1.4 Hz.*

