---
title: "Quiz 6 Spring 2022 answers"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
---

## Making the data (not in student version)

From:  
* https://covid.cdc.gov/covid-data-tracker/#vaccinations_vacc-people-onedose-pop-5yr
* https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf


```{r}
regions <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSHjDv2eSCOtuux9E6JIoEx2F4cOyU6jX6mXUr_FQDtTjpQxroh-IP_3D1z_qcSDr-9lQRQG_CRMZXY/pub?gid=0&single=true&output=csv",
                    stringsAsFactors = T)
rates <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vT6WmtWdZSUo--2i2nMCWUIMx6BjdJJHKH42oJBzdIZFlAxLH1XztcMI8na-4QKU8ySMHf2-41JwR1y/pub?gid=418071752&single=true&output=csv",
                   stringsAsFactors =T)
```
merge

```{r}
names(rates)[names(rates) == "State.Territory.Federal.Entity"] = "State"
covid_regions <- merge(rates, regions, all.x = F, all.y = T)
covid_regions <- covid_regions[,c("State", "Doses.Delivered.per.100K", "Region", "Division")]
write.csv(covid_regions, "covid_regions.csv", row.names = F)
```

## Quiz version

COVID-19 vaccines are a key part of the country's response to the pandemic, but 
they only work if people take them. The CDC keeps detailed records of dosages, 
which I combined with the US Census Bureau's data on regions. Data from:  
* https://covid.cdc.gov/covid-data-tracker/#vaccinations_vacc-people-onedose-pop-5yr
* https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf

Use the data (combined version provided below) to determine if vaccination rates 
differ by region.

```{r}
vaccine <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQsI2fpJSpI4pEgvUba3xfafpubrdWj3BaoCigWr4K0eUOGdwWgPygu2-Y6kwqZ-Abi92t4xFDsjbIY/pub?gid=200028780&single=true&output=csv",
                             stringsAsFactors = T)
```
Investigate the question. Make sure you include

* null hypothesis
  * H~0~: there is no difference among regions in vaccination rates
* alternative hypothesis
  * H~A~: there is a difference among regions in vaccination rates
* explanation for test you will use 
  * I will use an ANOVA as we are considering continuous data from differnet groups.
  I will also check assumptions
* results from statistical test
```{r}
vaccine_lm <- lm(Doses.Delivered.per.100K~Region, vaccine)
par(mfrow=c(2,2))
plot(vaccine_lm)
summary(vaccine_lm)
library(car)
Anova(vaccine_lm, type="III")
```
I noted a significant difference among regions, so I carried out post-hoc 
tests.

```{r}
library(multcomp)
region_compare <-   glht(vaccine_lm, linfct = mcp(Region = "Tukey"))
summary(region_compare)
```

* clear explanation of how results relate to your stated hypotheses
  * There is a difference among regions (F~3,46~=10.059, p <.01).  Posthoc tests
  using the Tukey method to control for FWER indicated the following areas differ
  at the $a = 0.05$ level:
    * NE- MW
    * S - NE
    * W- NE

