---
title: "2022 Spring 2100 Final answers (33 points total)"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
---

# Instructions! Read this first.

The exam is open note, open course website, open class-related code repositories 
(mine and those you produced in class). However, you may not get external help (
from other students, directed internet searches, etc.).  Please update the statement
below to acknowledge these instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam. I 
understand any academic issues will result in a D or F on the exam or in the class
and be reported to the Dean of Students.  Infractions will also result in me 
being unable to obtain a letter of recommendation from the department for 
professional school applications.

Note some questions are based on pretend surveys, etc, for teaching purposes.

Good luck!


## Making the data for # 4 (not in student version)

Real data from:  

* https://covid.cdc.gov/covid-data-tracker/#vaccinations_vacc-people-onedose-pop-5yr
* https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf


```{r}
regions <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSHjDv2eSCOtuux9E6JIoEx2F4cOyU6jX6mXUr_FQDtTjpQxroh-IP_3D1z_qcSDr-9lQRQG_CRMZXY/pub?gid=0&single=true&output=csv",
                    stringsAsFactors = T)
ratestotal <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTSY3PjqxOgUcMtx15Dzh1Z_TelhY6lsRloAyV65vKhqaPEahv5ojY_y7IOstTuUokyHtAYarkb6qqo/pub?gid=993160825&single=true&output=csv",
                   stringsAsFactors =T)
str(regions)
str(ratestotal)
```
separate only some dates for rates

```{r}
rates_focal <- ratestotal[ratestotal$Date %in% c("03/22/2021","03/21/2022"),c("Date","Location","Admin_Per_100K")]
library(reshape2)
rates_wide <- dcast(rates_focal, Location ~ Date, value = "Admin_Per_100k" )
```
merge

```{r}
names(rates_wide)[names(rates_wide) == "Location"] = "State.Code"
covid_regions_wide <- merge(rates_wide, regions, all.x = F, all.y = T)
names(covid_regions_wide)[2:3] <- c("early", "late")
covid_regions_wide$change <- covid_regions_wide$late - covid_regions_wide$early
covid_regions_wide <- covid_regions_wide[,c("State", "early", "late","Region", "change")]
#covid_regions_wide_south <- covid_regions_wide[covid_regions_wide$Region %in% c("South"),]
#covid_regions_wide_ne <- covid_regions_wide[covid_regions_wide$Region %in% c("Northeast"),]
#write.csv(covid_regions_wide_south, "covid_regions_wide_south.csv", row.names = F)
write.csv(covid_regions_wide, "covid_regions_wide.csv", row.names = F)
```

## COVID impacts

### 1. 

Investigators want to know if the rates of vaccination has changed at different
paces across the country. Data on change (column in dataset) in rates (number administered in March 
2022 (late)
minus number administered in March 2021(early) ) are available @

```{r}
covid_rates_over_year <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRLp3znA4ARXCNLlVHj4VSFPHyE7qCTHcs1xWDULfYxUux_6Fia_ugbVnHu8cURowahXWLz5yT7JFf1/pub?gid=1458185696&single=true&output=csv",
                                  stringsAsFactors = T)
```

Investigate the question. Make sure you include (6 pts)

* null hypothesis
  * *H~0~: Regions do not differ in rates of changes in vaccination*
* alternative hypothesis
  * *H~A~: Regions do differ in rates of changes in vaccination*
* explanation for test you will use 
  * I will use an ANOVA (form of linear model) since we have a continous outcome
  variable and a categorical predictor with more than 2 levels (else it would be
  a t-test)
* results from statistical test

```{r}
vaccine_lm <- lm(change~Region, covid_rates_over_year)
par(mfrow=c(2,2))
plot(vaccine_lm)
```

```{r}
summary(vaccine_lm)
library(car)
Anova(vaccine_lm, type="III")
```

```{r}
library(multcomp)
region_compare <-   glht(vaccine_lm, linfct = mcp(Region = "Tukey"))
summary(region_compare)
```
* clear explanation of how results relate to your stated hypotheses
  * Test assumptions appear to be met (no patterns in residuals, which are also
  normally distributed and homogeneous). ANOVA results indicate significant differences
  among regions (F~3,47~=10.843, p<.001), so I reject my null hypothesis. 
  Given these results, I used a Tukey post
  hoc test to control FWER while also seeing how each region compares to others. 
  Results indicate significant differences between the NE-MW, S-NE,  and W-NE 
  (so the NE differs from all other regions!).


### 2.  

There has been concern that the COVID pandemic may differentially influence 
male and female fetuses and thus alter the regional sex ratios
of newborn children (Abdoli 2020).  When comparing biological males and females at 
birth, males have historically represented approximately 50% of births. A hospital 
in New York City recorded 55 male and 45 female births in 2021 (in the middle of
the COVID pandemic).  Use this data to investigate if sex ratios have changed. 

Make sure you include (6 pts)

* null hypothesis
  * *H~0~: the sex ratio in 2021 is 50/50 (the historical average)*
* alternative hypothesis
  * *H~A~: the sex ratio in 2021 is not 50/50 (the historical average)*
* explanation for test you will use 
  * *I will use a binomial test since we have 2 outcomes (male/female) and a 
  historical proportion to compare to.
* results from statistical test
```{r}
binom.test(55,100)
```
* clear explanation of how results relate to your stated hypotheses
  * *Results indicate a p-value of .3682, so I fail to reject the null hypothesis.
  There is not evidence that COVID led to a change in sex ratios at birth.*

## To a happier place - Penguins!

3. 3 species of penguins (Gentoo, Chinstrap, Adelie) were measured for various 
morphological traits. Using the provided data, consider the impact of species(Species)
and body mass (Body_mass_g)
on flipper length (Flipper_length_mm). (6 pts)


```{r}
penguin <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQEyo_2bjGK6yEj5gj3SKP_VsqDxsqer4PRqTqqV_FRw9m0OM3u7SgbaT9OF4whgEb1Nx0z2QNU8ovd/pub?gid=836786692&single=true&output=csv", stringsAsFactors = T)
```

Make sure you include (6 pts)

* null hypothesis
* alternative hypothesis
* explanation for test you will use 
* results from statistical test
```{r}
penguin_lm_int <- lm(Flipper_length_mm~Body_mass_g*Species, penguin)
plot(penguin_lm_int)
summary(penguin_lm_int)
Anova(penguin_lm_int, type = "III")
```

interaction is significant, so complete for each

```{r}
chinstrap <- lm(Flipper_length_mm~Body_mass_g, penguin[penguin$Species == "Chinstrap",])
plot(chinstrap)
summary(chinstrap)
Anova(chinstrap, type = "III")
```

```{r}
Gentoo <- lm(Flipper_length_mm~Body_mass_g, penguin[penguin$Species == "Gentoo",])
plot(Gentoo)
summary(Gentoo)
Anova(Gentoo, type = "III")
```

```{r}
Adelie <- lm(Flipper_length_mm~Body_mass_g, penguin[penguin$Species == "Adelie",])
plot(Adelie)
summary(Adelie)
Anova(Adelie, type = "III")
```

* clear explanation of how results relate to your stated hypotheses





4. Create a properly-labelled plot of the data. There are many options here, but
make sure you show trends and patterns that match your analysis. (5 pts)

```{r}
library(ggplot2)
ggplot(penguin, aes(x=Body_mass_g, y=Flipper_length_mm, color= Species, 
                                   shape = Species)) +
  geom_point(size = 3) +
  geom_line(aes(group=Species, linetype =Species), size=2) +
  geom_smooth(method = "lm")+
  ylab("Flipper Length (mm)")+ 
  xlab("Body mass (g)") + 
  scale_shape_discrete(guide="none")+
  scale_linetype_discrete(guide="none")+
  ggtitle("The relationship between flipper length and body mass differs among 
          penguin species")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```

5. Based on your plot and results (answers to questions 3 and 4), clearly explain 
what your results mean.(4 pts)

*The significant interaction means species differ in the relationship between
body mass and flipper length. However, all are positive and significant.*

6. Upon reviewing the data, the researchers are concerned that the male-female
ratio may be different on various islands. Some of their initial analysis is 
provided below.

```{r}
penguin_not_missing <- penguin[penguin$Sex %in% c("MALE", "FEMALE"),]
penguin_not_missing$Sex <- factor(penguin_not_missing$Sex)
table(penguin_not_missing$Sex,penguin_not_missing$Island)
```
Use this data to investigate if sex ratios differ among islands. (6 pts)

Make sure you include

* null hypothesis
* alternative hypothesis
* explanation for test you will use 
* results from statistical test
* clear explanation of how results relate to your stated hypotheses

```{r}
island_impacts <-matrix(c(80,61,24,83,62,23), nrow = 2, byrow = T)
rownames(island_impacts) <- c("Male", "Female")
colnames(island_impacts) <- c("Biscoe", "Dream", "Torgersen")
chisq.test(island_impacts)
#or
chisq.test(table(penguin_not_missing$Sex,penguin_not_missing$Island))
```


## Butterflies  

7.Data on butterflies from the United Kingdom is available @

```{r}
butterflies <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTGEITZzcvnfDMwdXGScOaAtzMQDtIKaFJXw7220J_efxvvPNrPLIt1qG1UP2bHiNZ2uJDZInOsZJTg/pub?gid=987706272&single=true&output=csv",
                        stringsAsFactors = T)
```

Use this data to investigate if there is a relationship between dry mass 
(estimated_dry_mass) and the maximum wing size (forewing_maximum) observed in 
species. 

Make sure you include (6 pts)

* null hypothesis
* alternative hypothesis
* explanation for test you will use 
* results from statistical test
* clear explanation of how results relate to your stated hypotheses

```{r}
butterfly_lm <- lm(forewing_maximum ~estimated_dry_mass, butterflies)
plot(butterfly_lm)
summary(butterfly_lm)
Anova(butterfly_lm, type = "III")
```


