---
title: "Quiz 8 Spring 2022 answers"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
---


The salamander scientists continued their work.  In another study, the authors
determined the impact of temperature on heart rate. Data is available @

```{r}
temperature <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRhcbtbwuBje-9I7abkP2EXWuddGVo5VH1N2CgNtR8NMj-sjS-9KXjrJuZY9an9Gt4c-JbdCwwP9S0B/pub?gid=1230482827&single=true&output=csv", stringsAsFactors = T)
```

Analyze the data correctly. Make sure you include

* null hypothesis
  * H~0~: there is no relationship between heart rate and tempeature.
* alternative hypothesis
  * H~A~: there is a relationship between heart rate and tempeature.
* explanation for test you will use
    * Since heart rate is a continuous variable and temperature is a 
    continuous explanatory variable, I will plan to use a linear model (regression). However, I will check assumptions to ensure this is
  appropriate.
* results from statistical test

```{r}
relationship <- lm(heart_rate ~ temp, temperature)
plot(relationship)
summary(relationship)
library(car)
Anova(relationship)
```

  * Assumptions appear to be met.  I found temperature has a significant
  (F~1,80~=535, p<.001) impact on heart rate, with heart rate increasing 
  with temperature. Tempeature explains 86% of variation in heart rate.
* clear explanation of how results relate to your stated hypotheses
    * Since p<.05, I reject the null hypothesis.

* a properly-labelled plot of the data. There are many options here, but
make sure you show trends and patterns that match your analysis.

```{r}
library(ggplot2)
ggplot(relationship, aes(x=temp, y=heart_rate)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm") +
  ylab("Heart rate (bpm)")+ggtitle("Heart rate increases with temperature")+
  xlab("Temperature")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))