---
title: "Quiz 3 Spring 2022"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Medical applications of oxygen

A study was done to determine if oxygen was being appropriately supplied to patients
in a medical unit (Barrett et al 2020). They surveyed 636 patients in multiple wards and
found 66 were using oxygen.  Oxygen saturation levels were measured for each patient
using oxygen. Is there evidence patients receiving oxygen have saturation levels
that differ from the standard for a healthy person (95%)? 

Data can be loaded using 

```{r}
oxygen_use <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSC8ft2LReouozG9fDK0Pa-QXESZ82qzADTCCBr_LoQeVRlRDLS4KOhMNYjGnGZFTrq5hkuYcOk6cKm/pub?gid=1295231330&single=true&output=csv",
                       stringsAsFactors = T)
```

Make sure your answers include

* null hypothesis
* alternative hypothesis
* explanation for test you will use
* results from statistical test
* clear explanation of how results relate to your stated hypotheses
* confidence interval for your estimate