---
title: "Quiz 7 Spring 2022 answers"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
---


Three morphotypes of salamanders were tested to see if they had similar tail lengths.  The authors collected tail length (in mm) in the three morphotypes. They also noted the stage (juvenile vs adult) of each individual. Data is available @

```{r}
sallies <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSJHbPnebpKq58mE4kQWF9mQ_0ak2sTyaUbEvtmgoSkb31DprOmAS9clOoNNHD_3tHITOXKdhMbWfBf/pub?gid=1050993814&single=true&output=csv", stringsAsFactors = T)
```

Analyze the data correctly to consider how stage and morphotype impact heart 
rate . Make sure you include

* null hypothesis
  * H~0~: there is no difference among morphotypes in tail length
  * H~0~: there is no difference among stages in tail length
  * H~0~: there is no interaction among morphotype and stage on tail length
* alternative hypothesis
  * H~A~: there is a difference among morphotypes in tail length
  * H~A~: there is a difference among stages in tail length
  * H~A~: there is an interaction among morphotype and stage on tail length
* explanation for test you will use
  * Since tail length is a continous variable and morphotype and stage 
  are both categorical explanatory variables, I will plan to use a linear model (2 way ANOVA). However, I will check assumptions to ensure this is
  appropriate.
  
```{r}
sallies$morphotype <- as.factor(sallies$morphotype)
sallies$stage <- relevel(sallies$stage, "juvenile")
sallies_lm_int <- lm(tail_length~stage*morphotype, sallies)
plot(sallies_lm_int)
summary(sallies_lm_int)
library(car)
Anova(sallies_lm_int, type = "III")
```

* results from statistical test
    * Graphs show assumptions appear to be met.  Since the interaction is 
    significant (F~174,2~=4.62, p = 0.01) I analyzed both data sets 
    individually.
    
```{r}
sallies_lm_1 <- lm(tail_length~stage, sallies[sallies$morphotype == "1",])
plot(sallies_lm_1)
summary(sallies_lm_1)
Anova(sallies_lm_1, type = "III")

sallies_lm_2 <- lm(tail_length~stage, sallies[sallies$morphotype == "2",])
plot(sallies_lm_2)
summary(sallies_lm_2)
Anova(sallies_lm_2, type = "III")

sallies_lm_3 <- lm(tail_length~stage, sallies[sallies$morphotype == "3",])
plot(sallies_lm_3)
summary(sallies_lm_3)
Anova(sallies_lm_3, type = "III")
```
    
Tail length appear to differ among stage for morphotypes 1 (F~1,58~=31.23,
p <.001) and 2 (F~1,58~=41.23,
p <.001) with adults having larger tails.  However, tail length does not 
differ by stage for morphotype 3 (F~1,58~=0.916,
p =.161)).
* clear explanation of how results relate to your stated hypotheses
  * Findings indicate we should reject the null hyothesis of no interactiong among stage and morphotype on tail length.  The impact of stage differs by morphotype.
* a properly-labelled plot of the data. There are many options here, but
make sure you show trends and patterns that match your analysis.

```{r}
library(Rmisc)
sallies_summary <- summarySE(sallies, measurevar="tail_length", groupvars =
                               c("morphotype", "stage"), na.rm = T)
library(ggplot2)
ggplot(sallies_summary, aes(x=morphotype, y=tail_length, color=stage, 
                                   shape = stage)) +
  geom_point(size = 3) +
  geom_line(aes(group=stage, linetype =stage), size=2) +
  geom_errorbar(aes(ymin=tail_length-ci, ymax=tail_length+ci), size=1.5) +
  ylab("Tail Length")+ 
  xlab("Morphotype") + 
  scale_shape_discrete(guide=FALSE)+
  scale_linetype_discrete(guide=FALSE)+
  ggtitle("Tail length across stages of various morphotypes")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

ggplot(sallies_summary, aes(x=stage, y=tail_length, color=morphotype, 
                                   shape = morphotype)) +
  geom_point(size = 3) +
  geom_line(aes(group=morphotype, linetype =morphotype), size=2) +
  geom_errorbar(aes(ymin=tail_length-ci, ymax=tail_length+ci), size=1.5) +
  ylab("Tail Length")+ 
  xlab("stage") + 
  scale_shape_discrete(guide=FALSE)+
  scale_linetype_discrete(guide=FALSE)+
  ggtitle("Tail length across stages of various morphotypes")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

```



