---
title: "2020 Bio 2100 Final: A brief history of preventative dentistry (26 pts total)"
author: "jsg"
date: "12/12/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Instructions! Read this first.

The exam is open note, open course website, open class-related code repositories 
(mine and those you produced in class). However, you may not get external help (
from other students, directed internet searches, etc.).  Please update the statement
below to acknowledge these instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam. I 
understand any academic issues will result in a D or F on the exam or in the class
and be reported to the Dean of Students.

Good luck!

Have you ever wondered why we use fluoride in toothpaste? The data below is not 
exact but follows the general story and gives a good view of how science and 
statistics work together.  Let's pretend you were a fellow scientist/dentist/doctor
who was part of the story.

1. Frederick McKay moved to Colorado Springs, Colorado after completing dental 
school in 1901.  He was surprised to find the locals had extremely stained teeth 
(to the point they permanently looked like they had recently eaten chocolate candy).
In general
8% of people at the time had stained teeth, but in his first month on the job 
he noted six
out of 15 patients had stained teeth. 

He thinks this seems like a lot but asks you to help give his feeling a quantitative
analysis.  How would you evaluate the data? Make sure to include

  - null hypothesis
  - alternative hypothesis
  - explanation for test you will use
  - results from statistical test (including post-hoc tests if needed!)
  - clear explanation of how results relate to your stated hypotheses

```{r}
binom.test(6,15,.08)
binom.test(6,15)
```

2. McKay spent the next several years trying to determine what was causign the 
staining (or mottling, as he called it) and was assisted by multiple others.  Dr.
G.V. Black was a primary collaborator.  Together they investigated a few ideas.  First, 
they compared tooth staining rates in people who had been in the town vs those who
had moved their at multiple points.  They found 

| *Age moved to town*           | Stains present | Stains absent    |
| ---------                     | -------        | -------          |
| born in town                  | 50               | 20              |
| moved to town as young child (<10)  | 48           | 23              |
| moved to town as older child (11-20)     | 7           | 81              |
| moved to town as adult  (20+)      | 3            |  35             |

How would you evaluate the data? Make sure to include

  - null hypothesis
  - alternative hypothesis
  - explanation for test you will use
  - results from statistical test (including post-hoc tests if needed!)
  - clear explanation of how results relate to your stated hypotheses
  
```{r}
stain_group <-matrix(c(50,20,48,23,7,81,3,35), ncol = 2, byrow = T)
colnames(stain_group) <- c("stain", "no stain")
rownames(stain_group) <- c("born","child","teenager","adult")
#check it
stain_group
chisq.test(stain_group)
library(rcompanion)
pairwiseNominalIndependence(stain_group, compare = "row", method = "holm")
```

3.  What does your answer to question #2 imply?

4.  During their work the dentists also noticed that stained teeth behaved 
differently than unstained teeth.  Data they collected on cavities is found at

```{r}
set.seed(4)
cavities <- data.frame(stained = c(rep("yes", 25),rep("no", 50)), 
                       number_of_cavities_per_tooth = c(runif(25), 2 + runif(50)))
write.csv(cavities, "cavities.csv", row.names = F)
```

```{r}
cavities <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRuczoRwyQsMBSH0ONdoUlkzBFLbzwSEGNaj14ZXePJtr59QY7C3N2AYzKhm4Pdo5bKteqRMzwTZT9c/pub?gid=220386619&single=true&output=csv", stringsAsFactors = T)
```
What does the data suggest? Make sure to include

  - null hypothesis
  - alternative hypothesis
  - explanation for test you will use
  - results from statistical test (including post-hoc tests if needed!)
  - clear explanation of how results relate to your stated hypotheses
  
```{r}
t.test(number_of_cavities_per_tooth~stained, cavities)
cavities_lm <- lm(number_of_cavities_per_tooth~stained, cavities)
library(car)
Anova(cavities_lm, type = "III")
```
5. Provide a properly-labelled graph of the data from question 4. Make sure it shows both the signal
and noise in the data!

```{r}
library(ggplot2)
library(Rmisc)
cavities_summary <- summarySE(cavities, 
                              measurevar = "number_of_cavities_per_tooth",
                              groupvars = "stained")
ggplot(cavities_summary, aes(x=stained, y=number_of_cavities_per_tooth))+
  geom_point() +
  geom_errorbar(aes(ymin = number_of_cavities_per_tooth - ci, 
                    ymax = number_of_cavities_per_tooth + ci))

ggplot(cavities, aes(x=stained, y=number_of_cavities_per_tooth))+
  geom_point()
```


6. An unfortunate incident in Idaho shed more light on the causes of staining. 
People noticed a slight increase in tooth staining after a town switched to a new
water source. Consider the average tooth color (0 is white using analysis tools from
the time, 100 is black) of children from families who had children before and 
after the switch.  Data is available @ 

```{r}
set.seed(2)
idaho_cavities <- data.frame(family = letters[1:25],before = 10 + rnorm(25,0,3))
idaho_cavities$after <- idaho_cavities$before + 2 + rnorm(25,0,3)
write.csv(idaho_cavities, "idaho_cavities.csv", row.names = F)
```

```{r}
idaho_cavities <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQB6vCEkF79IKADjr8nugsbrOnHyk15NJbo7_Q4qWRfJ-OrJ8FoFfkkzFQ25mtVwKVZyWF_kh97SJfO/pub?gid=1969060757&single=true&output=csv", stringsAsFactors = F)

```

What does the data suggest? Make sure to include

  - null hypothesis
  - alternative hypothesis
  - explanation for test you will use
  - results from statistical test (including post-hoc tests if needed!)
  - clear explanation of how results relate to your stated hypotheses

```{r}
t.test(idaho_cavities$before, idaho_cavities$after, paired = T)
t.test(idaho_cavities$before, idaho_cavities$after)
```

7.  Analysis of water from other towns where staining was common eventually revealed
high amounts of fluoride. Scientists then collected data on average tooth decay rates
and fluoride levels from multiple towns (this part is made up!).  Data is available @

```{r}
set.seed(4)
fluoride <- data.frame(fluoride_ppm = runif(50))
fluoride$tooth_decay_rates <- 25 + fluoride$fluoride_ppm*15+2*runif(50) 
write.csv(fluoride, "fluoride.csv", row.names = F)
```

```{r}
fluoride <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vS7VvJ5WyzEIMyj-HGArfu_ZIT5cneHI5RiqmDIW_TaDDKgFTtBgcPXCPA00TLO4ax-xJlknn2bWfBi/pub?gid=1554561003&single=true&output=csv", stringsAsFactors = T)
```

What does the data suggest? Make sure to include

  - null hypothesis
  - alternative hypothesis
  - explanation for test you will use
  - results from statistical test (including post-hoc tests if needed!)
  - clear explanation of how results relate to your stated hypotheses

```{r}
fluoride_lm <- lm(tooth_decay_rates ~ fluoride_ppm, fluoride)
summary(fluoride_lm)
```
#Epilogue

In the real world, researchers worked to determine how much fluoride was 
too much in regards to staining (which is a more complicated analysis) before 
eventually considering its usefulness in low (safe) concentrations.  After 
large-scale studies showed results, it became a major component of preventative
dentistry.





