---
title: "2020 Bio 2100 Midterm: Halloween Redux (26 pts total)"
author: "jsg"
date: "10/27/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Instructions! Read this first.

The exam is open note, open course website, open class-related code repositories 
(mine and those you produced in class). However, you may not get external help (
from other students, directed internet searches, etc.).  Please update the statement
below to acknowledge these instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam. I 
understand any academic issues will result in a D or F on the exam or in the class
and be reported to the Dean of Students.

Good luck!

# Mind reading
## *Questions 1 - 3*

Researchers want to determine if mind-reading is possible.  To test this, they
place a purported mind-reader in a soundproof room. Directly outside the room 
is a chair. For 
each data point, a volunteer sits in the chair or doesn't. All the mind-reader has 
to do is "detect" the presence of the volunteer (this should be the first step in 
reading minds!).  Even though the mind-reader can't 
see or hear if the volunteer is present, he correctly determines if a person is 
present or not 12 times out of 20. 

1.  How would you assess this evidence to see if 
it offers evidence supporting the existence of mind-reading (6 pts)? 

Make sure your answers include (if applicable)

* null hypothesis
* alternative hypothesis
* explanation for test you will use
* results from statistical test
* clear explanation of how results relate to your stated hypotheses


2.  Calculate a confidence interval for the mind-reader data.  Share it and clearly
explain what it means (3 pts).  

3. If the mind-reader wanted to redo the experiment, what could
they do to reduce the width of the confidence interval **ASSUMING** the noted signal
in the data remains the same. Make sure you explain and justify your answer (3 pts). 

# Ghosts
## *Questions 4 - 6*

People who hunt for ghosts typically note their "presence" is associated with lower
temperatures.  To note this, they use a "ghost detecting machine" to document the
presence or absence of ghosts in a room.  The machine also documents the room 
temperature.  Data is available using

```{r}
ghosts <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/ghosts.csv", stringsAsFactors = T)
```


4. Provide a graph of the data that clearly displays patterns in the central 
tendency and variance of the data (many options here!) (5 pts).

5.  One issue with the data is that ghosts mainly occur in old (and poorly insulated)
homes.  Data shows the average temperature of homes where ghosts are present is 
on average 56 degrees when ghosts are not detected.  Using this information, how 
would you assess the evidence 
to see if 
it offers evidence supporting the existence of ghosts (6 pts)? 

As a hint, remember you can subset the data using 
`ghosts[ghosts$Present == "Present", "Temperature"]`.

Make sure your answers include (if applicable)

* null hypothesis
* alternative hypothesis
* explanation for test you will use
* results from statistical test
* clear explanation of how results relate to your stated hypotheses
* confidence interval for your estimate

6.  The investigator is confused. He found an average temerature of 55.8, which is
less than 56.  He thinks that should confirm the existence of ghosts. Explain any
issues you see with his rationale in plain English.  Also explain to him what a 
p-value means as part of your answer (3 pts).




