---
title: "2020 Bio 2100 Midterm: Halloween Redux (26 pts total)"
author: "jsg"
date: "10/27/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Instructions! Read this first.

The exam is open note, open course website, open class-related code repositories 
(mine and those you produced in class). However, you may not get external help (
from other students, directed internet searches, etc.).  Please update the statement
below to acknowledge these instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam. I 
understand any academic issues will result in a D or F on the exam or in the class
and be reported to the Dean of Students.

Good luck!

# Mind reading
## *Questions 1 - 3*

Researchers want to determine if mind-reading is possible.  To test this, they
place a purported mind-reader in a soundproof room. Directly outside the room 
is a chair. For 
each data point, a volunteer sits in the chair or doesn't. All the mind-reader has 
to do is "detect" the presence of the volunteer (this should be the first step in 
reading minds!).  Even though the mind-reader can't 
see or hear if the volunteer is present, he correctly determines if a person is 
present or not 12 times out of 20. 

1.  How would you assess this evidence to see if 
it offers evidence supporting the existence of mind-reading (6 pts)? 

Make sure your answers include (if applicable)

* null hypothesis

*(For a sided test) The mind reader can not detect the presence of people, so they will guess 
correctly 50% or less of the time (what we would expect by chance).* 
->H~O~: p <= .5

*(For a non-sided test) The mind reader can not detect the presence of people, so they will guess 
correctly 50% of the time (what we would expect by chance).* ->H~O~: p = .5

* alternative hypothesis

*(For a sided test) The mind reader can detect the presence of people, so they will guess correctly greater than 50% of the time .*->H~A~: p >.5

*(For a non-sided test) The mind reader can  detect the presence of people, so
the proportion of times they will guess correctly is not equal to 50% 
(what we would expect by chance).* ->H~A~: p != .5

* explanation for test you will use

*I will use binomial test since its discrete data (counts) that falls into 2 
groups.*

* results from statistical test

```{r}
binom.test(12,20, alternative="greater", .5)
```

*For the sided version, I found a p-value of .2517.*

```{r}
binom.test(12,20, .5)
```

*For the non-sided version, I found a p-value of .5034*


* clear explanation of how results relate to your stated hypotheses

*In both cases I fail to reject the null hypothesis. The data does not contain
enough evidence to indicate mind readers can detect the presence of people.*


2.  Calculate a confidence interval for the mind-reader data.  Share it and clearly
explain what it means (3 pts).  

```{r}
library(binom)
#non-sided option
binom.confint(12,20) 
```

*Using the Agresti-Coull method, we find a 95% confidence interval of .39 - .78. 
a 95% confidencei interval means 95% of the time we sample the population and 
create this interval it will contain the true parameter. Note that this interval
contains .5, meaning the data does not offer much evidence against the parameter 
under the null hypothesis.*

3. If the mind-reader wanted to redo the experiment, what could
they do to reduce the width of the confidence interval **ASSUMING** the noted signal
in the data remains the same. Make sure you explain and justify your answer (3 pts). 

*Given the formula for confidence intervals, the best approach is to have a larger
sample size.  A larger sample size where the mind reader guessed correctly 60% of 
the time could lead to "significant" outcome.*

# Ghosts
## *Questions 4 - 6*

People who hunt for ghosts typically note their "presence" is associated with lower
temperatures.  To note this, they use a "ghost detecting machine" to document the
presence or absence of ghosts in a room.  The machine also documents the room 
temperature.  Data is available using

```{r}
ghosts <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/ghosts.csv", stringsAsFactors = T)
```

3. Provide a graph of the data that clearly displays patterns in the central 
tendency and variance of the data (many options here!) (5 pts).

*Lots of options here. Some shown below. Key was to include some measure of 
central tendency and variation OR just plot the data!*

```{r}
library(ggplot2)
#dotplot
ggplot(ghosts, aes (x = Present, y = Temperature)) +
  geom_point() +
  ggtitle("Room temperature depends on presence of ghosts") +
  xlab("Ghost Reading") +
  ylab ("Temperature in degrees F")
#with confidence interval
library(Rmisc)
ghosts_summary <- summarySE(ghosts, measurevar = "Temperature", groupvars = "Present")
ggplot(ghosts_summary, aes (x = Present, y = Temperature)) +
  geom_point() +
  geom_errorbar(aes(ymin = Temperature - ci, ymax = Temperature + ci))+
  ggtitle("Room temperature depends on presence of ghosts") +
  xlab("Ghost Reading") +
  ylab ("Temperature in degrees F")
```


4.  One issue with the data is that ghosts mainly occur in old (and poorly insulated)
homes.  Data shows the average temperature of homes where ghosts are present is 
on average 56 degrees when ghosts are not detected.  Using this information, how 
would you assess the evidence 
to see if 
it offers evidence supporting the existence of ghosts (6 pts)? 

As a hint, remember you can subset the data using 
`ghosts[ghosts$Present == "Present", "Temperature"]`.

Make sure your answers include (if applicable)

* null hypothesis

*People had issues with my wording here, so I was very open to interpretation.  
What I was trying to point out is that old  houses tend to be colder in general, 
so our null hypothesis would be the temperature when ghosts are "detected" is
no different than 56 degrees. H~O~: mean temperature = 56*

* alternative hypothesis

The temperature when ghosts are "detected" is
different than 56 degrees. H~A~: mean temperature != 56

* explanation for test you will use

*This is continuous data from one population, so we can use a t.test if we assume
the data (and thus residusals, as we now know) are normally distributed. We could 
also bootstrap the data. Other options (sign tests or rank tests) also work but 
have less power.*

* results from statistical test

```{r}
t.test(ghosts[ghosts$Present == "Present", "Temperature"], mu=56)
#or bootstrap
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/bootstrapjsg.R")
bootstrapjsg(ghosts[ghosts$Present == "Present", "Temperature"], null=56)
```
*Using the t-test I found a test value t~29~=0.18536, which corresponds to a 
p-value of 0.8542.  Bootstrap tests led to similar p-value (.8494)

* clear explanation of how results relate to your stated hypotheses

*Using either method I reject the null hypothesis. There is not enough evidence
to indicate the presence of ghosts lowers temperatures.*

* confidence interval for your estimate

*From the t-test, I get a 95% confidence interval of 54.00882-57.66024; the 
bootstrap interval is 54.06-57.52. Note both include 56.*


5.  The investigator is confused. He found an average temerature of 55.8, which is
less than 56.  He thinks that should confirm the existence of ghosts. Explain any
issues you see with his rationale in plain English.  Also explain to him what a 
p-value means as part of your answer (3 pts).

*The small difference he observed between 55.8 and 56 degrees may be explained by
sampling error.  A p-value indicates hwo likely we were to see the observed signal
or something more extreme if the null hypothesis were true and multiple samples 
were collected. 55.8 is within the region of acceptable variation from 56; it 
does not indicate a devation from the average temperature of 56.*




