---
title: "Quiz 8"
author: "jsg"
date: "11/18/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Vaccine effectiveness

A recent study of the new coronavirus vaccine produced by Johnson and Johnson 
led to a surprising result. Lower dosages of the vaccine led to a higher 
effectiveness. Typically one would assume that higher dosages are better.

Higher dosages, however, may lead to more side effects and also impact availability.  Imagine a study considering dosage level (of a different vaccine) impacts on systolic blood pressure (a potential side effect). Data can be found at

```{r eval=F, include=F}
set.seed(26)
dosage <- data.frame(dose_micrograms = c(rep(0,25),rep(25,25), rep(50,25), 
                                         rep (75, 25), rep (100,25)))
dosage$systolic <- 110+.35*dosage$dose_micrograms + rnorm(125,0,3)
write.csv(dosage, "dosage.csv", row.names = F)
```

```{r}
dosage <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSEsIPp-49ERvzN2EAC37NrSr9GLsWFTVgIIjBxGwloYoxz5mpUTJ2KCH7kXVxBGLtg9eSLs6YIx32E/pub?gid=993951219&single=true&output=csv", stringsAsFactors = T)
```

How would you evaluate the results?

Make sure your answers include

* null hypothesis

*There is no relationship between dosage level and systolic blood pressure.* B=0

* alternative hypothesis

*There is a relationship between dosage level and systolic blood pressure.* B!=0

* graph that clearly displays the data and represents these hypotheses

```{r}
library(ggplot2)
ggplot(dosage, aes(x=dose_micrograms, y=systolic)) +
         geom_point()+
         geom_smooth(method = "lm")+
         xlab("Vaccine dose (micrograms)")+
         ylab("Systolic blood pressure")+
         ggtitle("Systolic blood pressure increases as vaccine dosage increases")
```

* explanation for test you will use

*I would test for a relationship using regression. The dosage levels were preset, 
and both the response and explanatory variables are continuous.*

* results from statistical test (including post-hoc tests if needed and indication
of how much variation the model explains!)

```{r}
impacts <- lm(systolic~dose_micrograms, dosage)
plot(impacts)
library(car)
Anova(impacts, type = "III")
summary(impacts)
```

*I first ensured model assumptions were met. Plots show no patterns in residuals
and a relatively normal distribution, which means they are. Analysis indicates a
significant relationship between dose and systolic blood pressure (F~1,123~=2059,
p<.001), so I reject the null hypothesis. Model analysis indicates dosage level
explains 94% of the variation in blood pressure (this an example!). There is no
need for post-hoc tests in regression!*

* clear explanation of how results relate to your stated hypotheses

*noted above.*