---
title: "Quiz 9"
author: "jsg"
date: "11/18/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Vaccine effectiveness

An earlier study (Quiz 8) focused on a relationship between dosage and systolic blood pressure).  A follow-up wanted to consider if ethnicity impacted this 
relationship.  Data were collected from 4 different groups (labeled A-D here) on 
vaccine dosage level and systolic blood pressure. Data can be found at

```{r}
dosage_group <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSXfNA4nl1SRK6BDxfJqo7izW254hxP8N4tjK9snXZngfgQFJ3aeKlg5EAAWPfrStv_61e3FKsAOjjx/pub?gid=73925069&single=true&output=csv", stringsAsFactors = T)
```

How would you evaluate the results?

Make sure your answers include

* null hypothesis
* alternative hypothesis
* graph that clearly displays the data and represents these hypotheses
* explanation for test you will use
* results from statistical test (including post-hoc tests if needed and indication
of how much variation the model explains!)
* clear explanation of how results relate to your stated hypotheses