---
title: "Quiz 8"
author: "jsg"
date: "11/18/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Vaccine effectiveness

A recent study of the new coronavirus vaccine produced by Johnson and Johnson 
led to a surprising result. Lower dosages of the vaccine led to a higher 
effectiveness. Typically one would assume that higher dosages are better.

Higher dosages, however, may lead to more side effects and also impact availability.  Imagine a study considering dosage level (of a different vaccine) impacts on systolic blood pressure (a potential side effect). Data can be found at

```{r}
dosage <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSEsIPp-49ERvzN2EAC37NrSr9GLsWFTVgIIjBxGwloYoxz5mpUTJ2KCH7kXVxBGLtg9eSLs6YIx32E/pub?gid=993951219&single=true&output=csv", stringsAsFactors = T)
```

How would you evaluate the results?

Make sure your answers include

* null hypothesis
* alternative hypothesis
* graph that clearly displays the data and represents these hypotheses
* explanation for test you will use
* results from statistical test (including post-hoc tests if needed and indication
of how much variation the model explains!)
* clear explanation of how results relate to your stated hypotheses

