---
title: "5. Contingency Analysis:  Extending the binomial to more categories"
author: "jsg answers"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
    toc_depth: 5
---

Before doing this, review the **Contingency Analysis** lecture set slides from 
https://sites.google.com/view/biostats/lessons/contingency-analysis and
the  **5_beyond_binomial.R** script in the lecture_files folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats).

Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  * **push** updates to github
  
## Examples

Issue is we often get data in spreadsheet format (expanded/long or wide/summarized,
each shown below), but we need to get a vector or matrix for chisq.test and
related functions.

### The data

Following the Everest example from class. Assume data is in a dataframe where 
each row is a group data point.  

```{r}
everest <- data.frame(Survived = c("Y","N","Y", "N"),
                      Oxygen = c("Used", "Used", "Not used", "Not used"),
                      Number = c(1045, 32, 88, 8))
```


Assume data is in a dataframe where 
each row is an individual data point.  

```{r}
library(mirt)
everest_expand <- expand.table(everest)
```

### tests

First, let's ask if the same amount of people used or did not use oxygen. WE can 
use the table command to summarize. Note the chisq.test, by default, assumes
each group is equally likely!

```{r}
table(everest_expand$Oxygen)
chisq.test(table(everest_expand$Oxygen)) 
```
Dong this with summarized data is actually harder

```{r}
aggregate(Number~Oxygen, everest, sum)$Number
chisq.test(aggregate(Number~Oxygen, everest, sum)$Number) 
```
But this is better!

```{r}
binom.test(table(everest_expand$Oxygen))
```

What if we wanted to compare to past years where 10% of climbers did not use oxygen? 
Note *table* function splits into alphabetical order.

```{r}
binom.test(table(everest_expand$Oxygen), p=.1)
```

What if we want to determine if using oxygen impacts surival?

```{r}
chisq.test(table(everest_expand$Oxygen, everest_expand$Survived))
```

Issue (which we'll address), but note same as

```{r}
chisq.test(table(everest_expand$Survived, everest_expand$Oxygen))
chisq.test(x = matrix(c(1045, 88, 32, 8), 2, 2, byrow = T))
chisq.test(x = matrix(c(1045, 32, 88,  8), 2, 2, byrow = T))
```

**Key is first argument  must be all the info**. This is different from (incorrect)
approach like

```{r}
chisq.test(everest$Survived,everest$Oxygen)
```
This is comparing split among Survived and not to split (expected) using Oxygen!


So order has minimal input with 2 groups. Other test options necessitated by 
the warning

```{r}
fisher.test(table(everest_expand$Oxygen, everest_expand$Survived))
library(DescTools)
GTest(table(everest_expand$Oxygen, everest_expand$Survived))
```

What if we added another group? Like Enriched, Regular, None for oxygen.

```{r}
everest_enriched <- data.frame(Survived = c("Y","N","Y", "N", "Y", "N"),
                      Oxygen = c("Regular", "Regular", "None", "None", rep("Enriched", 2)),
                      Number = c(1045, 32, 88, 8, 15, 2))
everest_enriched_expand <- expand.table(everest_enriched)
```

Now we compare

```{r}
table(everest_enriched_expand$Survived, everest_enriched_expand$Oxygen)
chisq.test(table(everest_enriched_expand$Survived, everest_enriched_expand$Oxygen))
```

Fisher again due to size

```{r}
fisher.test(table(everest_enriched_expand$Survived, everest_enriched_expand$Oxygen))
```

Now we follow-up, and rows/columns matter. Note default is row and fdr method. I 
order results for ease of view

```{r}
library(rcompanion)
everest_expand_correct_fdr <- pairwiseNominalIndependence(table(everest_enriched_expand$Survived, everest_enriched_expand$Oxygen))
everest_expand_correct_fdr[order(everest_expand_correct_fdr$p.adj.Fisher),]
```

Not quite what we wanted. How about


```{r}
everest_expand_correct_fdr <- pairwiseNominalIndependence(table(everest_enriched_expand$Survived, everest_enriched_expand$Oxygen),
                                                          compare = "col")
everest_expand_correct_fdr[order(everest_expand_correct_fdr$p.adj.Fisher),]
```

and you can change methods

```{r}
everest_expand_correct_fdr <- pairwiseNominalIndependence(table(everest_enriched_expand$Survived, everest_enriched_expand$Oxygen),
                                                          compare = "col",
                                                          method = "holm")
everest_expand_correct_fdr[order(everest_expand_correct_fdr$p.adj.Fisher),]
```
To put in manually, we need a few extra things

```{r}
everest_table <- as.table(matrix(c(2,8,32,15,88,1045), nrow = 2, byrow = T))
rownames(everest_table) = c("N", "Y")
colnames(everest_table) = c("Enriched", "None", "Regular")
everest_table
```


## Let's practice

### Heart attacks 

#### 1

1. Let’s look at some heart attack data.  Read in the data using
```{r}
heart_attacks <- read.table("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/heartatk4R.txt",header=T, stringsAsFactors = T)
```
Every entry is a person that has suffered a heart attack. More information on the dataset can be found at

http://statland.org/Software_Help/DataDesk/datafile.htm 

We want to again test if heart attacks occur equally across genders.
```{r}
table(heart_attacks$SEX)
binom.test(7779, 7779+5065)
```
*If I assume males compose 50% of the population, I can test the null hypothesis that 
50% of heart attacks occur in males using a binom.test to conduct a binomial test.
I used the table command 
to determine the number of heart attacks in males and
females and then used binom.test.  The alternative is less than or greater than 50% of
hear attacks occur in males. With a p-value of <.001, I reject the null hypothesis.
Data suggest that males are more likely to have heart attacks. Note this is better than* 
```{r}
chisq.test(table(heart_attacks$SEX), p=c(.50, .50))
```
*which is an approximate test*.

* What if we know that males actually make up 50.8% of the population?

```{r}
table(heart_attacks$SEX)
binom.test(7779, 7779+5065, .508)
```
*Note I can amend the test proportion as noted here.  Results do not change.*

#### 2

2. Still using the heart attack data, is survival independent of gender?
```{r}
#note what this does
table(heart_attacks$SEX, heart_attacks$DIED)
#then feed it to chisq.test (notice order here does not matter for 2x2 table)
chisq.test(table(heart_attacks$SEX, heart_attacks$DIED))
chisq.test(table(heart_attacks$DIED, heart_attacks$SEX))

```
*I used a chi^2^ to consider if survival was independent of sex. Our null 
hypothesis is that survival does not differ based on sex. the alternative is that
it does. I found a
chi^2^~1~=147.76, which corresponds to a p-value of <.001, so i reject the null
hypothesis.*

#### 3

3. For people that have a heart attack before they turn 30, is survival independent of gender?

```{r}
chisq.test(table(heart_attacks[heart_attacks$AGE < 30, "SEX"], 
                 heart_attacks[heart_attacks$AGE <30, "DIED"]))
#note warning on approximation, so check it
chisq.test(table(heart_attacks[heart_attacks$AGE < 30, "SEX"], 
                 heart_attacks[heart_attacks$AGE <30, "DIED"]))$expected
#several <1, so use fisher.test
fisher.test(table(heart_attacks[heart_attacks$AGE < 30, "SEX"], 
                  heart_attacks[heart_attacks$AGE <30, "DIED"]))
#so if you are young, no difference
```
*I amended the previous question's code to only focus on individuals who were under 
the age of 30 when they suffered a heart. Otherwise the hypotheses remain the same.
I attempted to use a chi^2^ test but was warned the approximation may be incorrect.
Remember that no cells can have expected values of <1 and <20% should have expected
values <5.  Upon checking 2 cells have expected values <1, so I instead used Fisher's Test.  
I found a p-value of 1, thus I fail to reject the null hypothesis.*


### Dolphins

#### 4


4. Data on dolphin behavior was collected off the coast of Iceland. Data is @

http://www.statsci.org/data/general/dolpacti.txt

Since this is a .txt file, not a .csv, you’ll need to use something like

```{r}
dolphin <- read.table("http://www.statsci.org/data/general/dolpacti.txt", sep="", header = T, stringsAsFactors = T)
```

More info on data @ 

http://www.statsci.org/data/general/dolpacti.html

Is traveling independent of time of day?  You’ll need to consider
traveling vs not traveling due to different number of groups observed
in each period.  Carry out post-hoc tests if needed.

*I looked at the data and then just made a table manually*

```{r}
dolphin
travel_table <- as.table(matrix(c(6, 28+ 38, 6, 9, 14, 9, 13, 66), nrow = 4, byrow = T))
#Adding in row and column names will make everything easier to read at end.
colnames(travel_table) = c("travel", "not_travel")
rownames(travel_table) = c("morning", "noon", "afternoon", "night")
#now look at it
travel_table
chisq.test(travel_table)
#check outcome given warning
chisq.test(travel_table)$expected
fisher.test(travel_table)
library(rcompanion)
pairwiseNominalIndependence(travel_table, compare = "row", method = "holm")
```
*I tested the null hypothesis that traveling is independent of time of day (compared
 to the alternative hypothesis that it is not, and thus differs aross time periods)
 using chi^2^ test. However, a warning and subsequent check indicated too many 
 cells had low expected values, so I instead used a Fisher's test. A resulting p-value 
 of <.001 led me reject the null hypothesis. Since I was comparing more than two 
 groups, I used a post-hoc test to see which periods were different and found that travel differed between morning and noon, morning and afternoon, and afternoon and night 
(using the p.adj.Fisher column).*
 
 

### Smoking

#### 5

5. Use data on smoking and exercise from

http://www.r-tutor.com/elementary-statistics/goodness-fit/chi-squared-test-independence

to determine if smoking is independent of exercise. You’ll need to input data manually.  Carry out post-hoc tests if needed.

*I created a table from the data and tested it using a chi^2^ test to determine 
if smoking was independent of exercise (null hypothesis) or differed based on exercise
levels (alternative). However, a warning led me to see the expected cell values were 
too small, so I instead used a Fisher's test.*

```{r}
smoke <- chisq.test(matrix(c(7, 1, 3, #spacing just for visual use
                             87,18,84,
                             12,3,4,
                             9,1,7), nrow = 4, byrow = T))
smoke$expected #too small!
fisher.test(matrix(c(7, 1, 3, #spacing just for visuals
                     87,18,84,
                     12,3,4,
                     9,1,7), nrow = 4, byrow = T))
```
*A p-value of .4138 meant I failed to reject the null hypothesis.*

