---
title: "6. Continuous tests 2 populations"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
    toc_depth: 6
---

Before doing this, review the **Extending continuous tests to 2 samples** lecture set slides from 
https://sites.google.com/view/biostats/lessons/2-sample-tests-for-continuous-data and
the  **6_2_sample_continuous tests.R** script in the lecture_files folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats).


Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  * **push** updates to github

## Examples

### Paired tests and rearranging data

Paired tests assume there is a one-to-one relationship between measurements in each
group (before/after, twin 1/twin 2, etc).
Note for paired tests the arrangement of the data matters. For example, if paired
data is stored so that each row has information from a single pair, we need a
different test format than if each row only has information from a single unit. 
In general, when a row only has information from one unit, we call it long-formatted
data. Wide-formatted data has information from multiple units on each row.  This
will get more complicated when we measure multiple traits/things on each unit, 
but its a start.

#### t-test

Let's first load a long-formatted dataset.

```{r}
cholesterol <- read.table("http://www.statsci.org/data/general/cholestg.txt", header = T,
                          stringsAsFactors = T)
head(cholesterol)
str(cholesterol)
```

Note this is long format data.  Day is also an integer here. For our purposes, lets consider it a factor (near/far
from heart attack).

```{r}
cholesterol$day <- as.factor(cholesterol$day)
library(plyr)
cholesterol$when <- revalue(cholesterol$day, c("2"="soon_after", "4" = "few_days",
                                               "14" = "weeks_later"))
```

To run a t-test on this format, we would use the formula notation and only focus 
on 2 periods (since its a t-test!). Note the *%in%* operator lets us search for 
a list of outcomes to match (instead of *==*, which matches one).

```{r, error=T}
t.test(cholest ~ when, cholesterol[cholesterol$when %in% c("soon_after", 
                                                           "weeks_later"),],
       paired = T)
```
This does not work (note I had to add an R chunk option, *error = T *, to make it
knit!).  Why?  The error mentions an issues with length.  Look at each piece

```{r}
 cholesterol[cholesterol$when %in% c("soon_after"), "cholest"]
 cholesterol[cholesterol$when %in% c("weeks_later"), "cholest"]
```

Note we have many NAs. Further inspection (look at the whole cholesterol dataset
by clicking on it in environment) shows we have 2 groups, which isn't helping. 
R often exclude's NAs by default. Try

```{r}
options("na.action")
```

It likely says "na.omit".  For paired tests in formula format this is an issue because
it looks for matched pairs and *then* excludes NAs. To make this work, let's 
make sure to send the function everything using the added na.action argument set to
"na.pass". 

```{r}
t.test(cholest~day, cholesterol[cholesterol$when %in% c("soon_after","weeks_later"),],
       paired = T, na.action = "na.pass")
```

Note this is same as 

```{r}
t.test(cholesterol[cholesterol$day == "2", "cholest"], 
       cholesterol[cholesterol$day == "14", "cholest"],
       paired = T)
```

which does not have the *NA* issue.  This setup is better for wide format data.

#### long to wide 

Let's make the cholesterol dataset wide usin the dcast function from the *reshape2* 
package. The formula notes row ~ columns.  

```{r}
library(reshape2)
#formula gives row ~ columns
#get daily cholesterol for every patient
cholesterol_wide <- dcast(data = cholesterol, formula = patient ~ day, 
                          value.var ="cholest" )
head(cholesterol_wide)
```

We could evaluate this using

```{r, error = T}
t.test(cholesterol_wide$2, cholesterol_wide$14, paired = T)
```
Another error!  R doesn't like columns to be named with numbers. Let's fix this.

```{r}
names(cholesterol_wide)[2:5] <- paste ("day", names(cholesterol_wide)[2:5], 
                                       sep="_")
head(cholesterol_wide)
```

Now try

```{r}
t.test(cholesterol_wide$day_2, cholesterol_wide$day_14, paired = T)
```
#### Wilcoxon test

Compare these to Wilcoxon test

```{r}
wilcox.test(cholest~day, cholesterol[cholesterol$when %in% c("soon_after","weeks_later"),],
       paired = T, na.action = "na.pass")
wilcox.test(cholesterol_wide$day_2, cholesterol_wide$day_14, paired = T)
```

#### Sign test

or Sign tests (which only take wide format and only work with paired data!)

```{r}
library(BSDA)
SIGN.test(cholesterol_wide$day_2, cholesterol_wide$day_14)
```

### Not paired

What if we had measured different people a few days before and several weeks after
a heart attack?  Then the data is not paired! Examples for all tests shown here.

#### t-tests

Note by default we get a Welsh test, which assumes variance is not the same among
the groups.

```{r}
t.test(cholest~day, cholesterol[cholesterol$when %in% c("soon_after","weeks_later"),])
t.test(cholesterol_wide$day_2, cholesterol_wide$day_14)
```
#### Wilcoxon test

```{r}
wilcox.test(cholest~day, cholesterol[cholesterol$when %in% c("soon_after","weeks_later"),])
wilcox.test(cholesterol_wide$day_2, cholesterol_wide$day_14)
```
#### Bootstrap

We also have options that don't rely on repeated sampling.  My bootstrap function
only works for wide format data (unless you manually subtract outcomes... and omit
NAs...)

```{r}
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/bootstrapjsg.R")
bootstrapjsg(cholesterol_wide$day_2, cholesterol_wide$day_14)

cholesterol_difference <- na.omit(cholesterol_wide$day_2 - cholesterol_wide$day_14)
bootstrapjsg(cholesterol_difference)
```
Note again how pairing helps!

#### Permutation

Note the permutation tests only works on unpaired data.

```{r}
library(coin)
independence_test(cholest~day, 
                  cholesterol[cholesterol$when %in% c("soon_after","weeks_later"),])
```
### More on rearranging data

If you need to re-arrange data **AND** summarize it in same step (not one-to-one
match when you rearrange), you'll need to tell R how to summarize it. For example,
we could find the average cholesterol for each patient (note there are 2 groups that
label each patient 1 ). Note I made a function to remove NAs, which otherwise cause
issues.

```{r}
#use fun.aggregate to get other option
#get average cholesterol per patient
#simple function to exclude na's (other option is to melt first and then drop them)
meannona <- function (x) mean(x, na.rm=T)
cholesterol_wide_summary <- dcast(data = cholesterol, formula = patient ~ ., #. means no variable, ... means all variables
                          value.var ="cholest", fun.aggregate = meannona )
head(cholesterol_wide_summary)
```

You can also name output column by putting in quotes in formula

```{r}
cholesterol_wide_summary <- dcast(data = cholesterol, formula = patient ~ "average_cholest", #. means no variable, ... means all variables
                          value.var ="cholest", fun.aggregate = meannona )
head(cholesterol_wide_summary)
```


#### Wide to long

If you have data in wide format

```{r}
cholesterol <- read.table("http://www.statsci.org/data/general/cholestr.txt", header = T)
head(cholesterol)
```

you can instead use the *melt* function (also in the *reshape2* package).  The
id.vars argument lists independent values to keep. The 
measure.vars variable is what you are measuring (not used here, and used rarely). 
The variable.name argument lists what to label things you group, and the value.name 
arguments gives name to value output

```{r}
cholesterol_long <- melt(cholesterol, id.vars =c())
head(cholesterol_long)

#name outcomes
cholesterol_long <- melt(cholesterol, id.vars =c(), variable.name = "day", 
                         value.name = "cholesterol")
head(cholesterol_long)
```

If you have more variables, you can melt them all to a single column (most useful
when faceting plots!).

```{r}
sport <- read.table("http://www.statsci.org/data/oz/ais.txt", header = T)
sport_melted <- melt(sport, id.vars = c("Sex", "Sport"),
                     variable.name = "measure", 
                     value.name = "value")
head(sport_melted)
```

## Practice

### For the following problems, use the data to conduct each noted test (note some sample sizes may be too small for these to all be good ideas!)


#### 1

1. The following data are human blood clotting times (in minutes) of individuals
given one of two different drugs. 

|Drug B | Drug G
|:-: |:-:
|8.8 | 9.9
|8.4 | 9.0
|7.9 | 11.1
|8.7 | 9.6
|9.1 | 8.7
|9.6 | 10.4 
|    | 9.5

Test the hypothesis that the mean clotting times
are equal for the two groups
*  Estimating the variance from the data 
```{r}
drug_b <- c( 8.8, 8.4, 7.9, 8.7, 9.1, 9.6)
drug_g <- c(9.9, 9.0, 11.1, 9.6, 8.7, 10.4, 9.5)
t.test(drug_b, drug_g)
```
*Using a un-paired t-test, since the experimental units were not matched and I 
assumed the means of each group would follow a normal distribution of unknown 
variance,  I found a test statistics of t~10.701~=-2.544.  This corresponds to a 
p-value of 0.02. This p-value is <.05, so I reject the null hypothesis that 
the mean clotting times are the same for the two drugs.*

* Using rank transform analysis 
```{r}
wilcox.test(drug_b, drug_g)
```
 
*Using a un-paired rank-based test, which is appropriate when normality assumptions 
can't be met and I assumed the means of each group would follow a similar distribution, 
I found a test statistics of W=7.  This corresponds to a 
p-value of 0.05.  This p-value is >.05, so I fail to reject the null hypothesis that 
the mean clotting times are the same for the two drugs.*

* Using a permutation test

```{r}
require(coin) #requires data_frame
clotting <- data.frame(drug = c(rep("drug_b", length(drug_b)), rep("drug_g", 
                                                                   length(drug_g))),
                       clotting = c(drug_b, drug_g))
clotting$drug <- factor(clotting$drug)
independence_test(clotting ~ drug, clotting)
```
*Using a permutation test, which is not fully appropriate here due to small sample 
sizes (and that also assumes similar distributions for each group), 
I found a test statistics of Z=-2.0726..  This corresponds to a 
p-value of 0.038.  This p-value is >.05, so I fail to reject the null hypothesis that 
the mean clotting times are the same for the two drugs.*

* Using a bootstrap test 

```{r}
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/bootstrapjsg.R")
bootstrapjsg(drug_b, drug_g)
```
*Using a bootstrap test with 10000 samples, which is not fully appropriate here 
due to small sample sizes, 
I found a p value of 0.0047.  This p-value is <.05, so I reject the null 
hypothesis that 
the mean clotting times are the same for the two drugs.*

#### 2


2. Data on plant heights (in cm) for plants grown with a new and old formulation 
of fertilizer can be found at
https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/fertilizer.csv . 
Use the data to test the hypothesis that there is no difference in mean plant
heights for the two groups 

*  Estimating the variance from the data 
```{r}
fertilizer <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/fertilizer.csv", stringsAsFactors = T)
t.test(height ~ fertilizer, fertilizer)
```
*Using a un-paired t-test, since the experimental units were not matched and I 
assumed the means of each group would follow a normal distribution of unknown 
variance,  I found a test statistics of t~15.559~=3.01.  This corresponds to a 
p-value of 0.008. This p-value is <.05, so I  to reject the null hypothesis that 
the impact on plant growth doesn not differ between the two fertilizer formulations.*

* Using rank transform analysis 
```{r}
wilcox.test(height ~ fertilizer, fertilizer)
```
*Using a un-paired rank-based test, which is appropriate when normality assumptions 
can't be met and I assumed the means of each group would follow a similar distribution, 
I found a test statistics of W=68.  This corresponds to a 
p-value of 0.012.  This p-value is <.05, so I  to reject the null hypothesis that 
the impact on plant growth doesn not differ between the two fertilizer formulations.*

* Using a permutation test 
```{r}
independence_test(height ~ factor(fertilizer), fertilizer)
```
*Using a permutation test, which is appropriate given the sample sizes and fact 
I assume similar distributions for each group, 
I found a test statistics of Z=2.47.  his corresponds to a 
p-value of 0.01.  This p-value is <.05, so I  to reject the null hypothesis that 
the impact on plant growth doesn not differ between the two fertilizer formulations.*

* Using a bootstrap test 

*This is where you may need the different (wide) format of data! As always, multiple
options exist in R*

```{r}
require(reshape2)
fertilizer$ID <- 1:nrow(fertilizer)
fertilizer_wide <- dcast(fertilizer, ID~fertilizer, value.var = "height")
#subsetting the long-form data
bootstrapjsg(fertilizer[fertilizer$fertilizer =="old", "height"],
              fertilizer[fertilizer$fertilizer =="new", "height"])
#is same as using the wide data
bootstrapjsg(fertilizer_wide$new, fertilizer_wide$old)
```
*Using a bootstrap test with 10000 samples, 
I found a p value of 0.013.  This p-value is <.05, so I  to reject the null hypothesis that 
the impact on plant growth doesn not differ between the two fertilizer formulations.*

Note you may need to cast this dataframe into a dataframe with a column for 
old and new outcomes.  You’ll have to add a unique identifier. Try something like

```{r, eval=FALSE}
fertilizer$ID <- 1:nrow(fertilizer)
fertilizer_wide <- dcast(fertilizer, ID~fertilizer, value.var = "height")
```


#### 3

3. (Example from  Handbook on Biological Statistics)  Odd (stunted, short, new)
feathers were compared in color to typical feathers
in Northern Flickers (*Colaptes auratus*). Data is at 

https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/wiebe_2002_example.csv

Test the hypothesis that odd and typical feathers did not differ using

*  a Student's t test
```{r}
feather <-  read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/wiebe_2002_example.csv
", stringsAsFactors = T)
t.test(Length ~ Feather, data=feather, paired=TRUE)
```
*I used a paired t-test because feathers were measured on the same bird.  
I also assumed the difference in means was normally
distributed given the trait and sample size.  The test resulted in a statistic of
t~15~ = -4.06. This corresponds to a p-value of .001.  Since the p-value is <.05,
I reject the null hypothesis that feather color is the same between odd and 
typical feathers.*

* a rank test
```{r}
wilcox.test(Length ~ Feather, data=feather, paired=TRUE)
```
*I used a paired rank-based test because feathers were measured on the same bird.
I did not assume the difference in means was normally
distributed but did assume it followed a symmetric distribution.  The test resulted in a statistic of
V = 10.   This corresponds to a p-value of .001.  Since the p-value is <.05,
I reject the null hypothesis that feather color is the same between odd and 
typical feathers.*

* a binary test
```{r}
library(BSDA)
SIGN.test(feather[feather$Feather == "Odd", "Length"], 
          feather[feather$Feather == "Typical", "Length"])
```

*I used a sign test (always paired!) because feathers were measured on the same bird.
I did not assume the difference in means was normally
distributed or that the differences followed a symmetric distribution.  The test resulted in a statistic of
s = 3.   This corresponds to a p-value of .02.  Since the p-value is <.05,
I reject the null hypothesis that feather color is the same between odd and 
typical feathers.*

* bootstrapping
```{r}
bootstrapjsg(feather[feather$Feather == "Odd", "Length"]- 
          feather[feather$Feather == "Typical", "Length"])
```
*Since feathers were measured on the same bird.
I used a bootstrap (10,000 samples) focused on the difference in color.  This 
resulted in a p-value of <.001. Since the p-value is <.05,
I reject the null hypothesis that feather color is the same between odd and 
typical feathers.*

#### 4

4. Endurance of the above animals was also measured. Data is available at

https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/endurance.csv

Test the hypothesis that the drug has no effect on using 
*  a Student's t test
```{r}
endurance <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/endurance.csv", stringsAsFactors = T)
head(endurance)
t.test(endurance ~ time, endurance, paired = T)
#or
t.test(endurance[endurance$time == "before", "endurance"],
       endurance[endurance$time == "after", "endurance"],
       paired = T)
```
*I used a paired t-test because the same animals were measured before and after 
the drug was administered.  I also assumed the difference in means was normally
distributed given the trait and sample size.  The test resulted in a statistic of
t~8~ = -3.39. This corresponds to a p-value of .29.  Since the p-value is >.05,
I fail to reject the null hypothesis that the drug has no effect on metabolic rate.*
* a rank test
```{r}
wilcox.test(endurance ~ time, endurance, paired = T)
```
*I used a paired rank-based test because the same animals were measured before and after 
the drug was administered.  I did not assume the difference in means was normally
distributed but did assume it followed a symmetic distribution.  The test resulted in a statistic of
V = 13.   This corresponds to a p-value of .30.  Since the p-value is >.05,
I fail to reject the null hypothesis that the drug has no effect on metabolic rate.*

* a binary test
```{r}
SIGN.test(endurance[endurance$time == "before", "endurance"],
          endurance[endurance$time == "after", "endurance"])
```
*I used a sign test (always paired!) because the same animals were measured before and after 
the drug was administered.  I did not assume the difference in means was normally
distributed or that the differences followed a symmetric distribution.  The test resulted in a statistic of
s = 7.   This corresponds to a p-value of .17.  Since the p-value is >.05,
I fail to reject the null hypothesis that the drug has no effect on metabolic rate.*

* bootstrapping
```{r}
bootstrapjsg(endurance[endurance$time == "before", "endurance"]-
             endurance[endurance$time == "after", "endurance"])
```
*Since the same animals were measured before and after the drug was administered,
I used a bootstrap (10,000 samples) focused on the difference in rates.  This 
resulted in a p-value of 0.23. Since the p-value is >.05,
I fail to reject the null hypothesis that the drug has no effect on metabolic rate.*


### For the following problems, make sure you can justify the test you choose to use and carry it out

#### 5

5. Using data from Australian athletes (http://www.statsci.org/data/oz/ais.html 
for details), determine if the average male training at the Australian Institute 
of Sport differs in from the average female training at the site in mercury
concentration (Hg). 

```{r}
sport <- sport <- read.table("http://www.statsci.org/data/oz/ais.txt", header = T, stringsAsFactors = T)
t.test(Hg ~ Sex, sport)
```
*I used a t-test (not paired) since mercury concentration can be assumed to follow
a normal distribution at this sample size.  This resulted in a test statistic of 
t~199.99~ = -15.25, with a corresponding p-value of <.001. Since this p-value is 
<.05, I reject the null hypothesis that there is no difference in average mercury 
concentration between males and females.  If I wanted to graph this, I might use*

```{r}
library(ggplot2)
ggplot(sport, aes_string("Hg"))+
  geom_histogram() +
  facet_wrap(~Sex) +
  ggtitle("	Plasma ferritin concentration \n of Australian athletes") +
  xlab("Ferritin concentration")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32),
        strip.text.x = element_text(size = 22))
```

*which suggests males tend to have a higher Hg concentration.*

#### 6

6. Laureysens et al. (2004) measured metal content in the wood of 13 poplar 
clones growing in a polluted area, once in August and once in November.
Concentrations of aluminum (in micrograms of Al per gram of wood) are shown 
below. (From Handbook of Biological Statistics). Has metal content changed over
time?

| Clone | August | November
|:-:| :-:|:-:
|Columbia River | 18.3 |12.7
|Fritiz Pauley | 13.3 |11.1
|Hazendas | 16.5 |15.3
|Primo | 12.6 | 12.7
|Raspalje | 9.5 |10.5
|Hoogvorst |13.6 |15.6
|Balsam Spire | 8.1 |11.2
|Gibecq   |8.9 |14.2
|Beaupre | 10.0 |16.3
|Unal |8.3 |15.5
|Trichobel |7.9 | 19.9
|Gaver | 8.1 | 20.4
|Wolterson | 13.4 |36.8

```{r}
poplar <- read.table("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/poplar.csv",
                   stringsAsFactors = T, header = T)
poplar$difference <- poplar$November - poplar$August

#plot to consider distribution
ggplot(poplar, aes_string("difference"))+
  geom_histogram() +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32),
        strip.text.x = element_text(size = 22))
#lots of holes, but fairly symmetric
wilcox.test(poplar$August, poplar$November, paired=TRUE)
#also could do
bootstrapjsg(poplar$difference)
```

*I plotted the paired data to consider its distribution.  Given its shape and a 
lack of knowledge regarding metal concentration distribution, I used a rank-based
test for paired data. This led to a statistic of V = 17 and p-value of 0.04. Since
this is < .05, I reject the null hypothesis that average metal concentration did not differ
between the two measurements.  Similar results were reached used a bootstrapping 
approach.*

#### 7

7. Data was collected on cholesterol levels following heart attacks.  Data
available at 
http://www.statsci.org/data/general/cholestg.txt 

Note to read a .txt file you use,
read.table("http://www.statsci.org/data/general/cholestg.txt", header = T, 
stringsAsFactors = T)
Check the defaults to understand the argument.  
More info on the data is @
http://www.statsci.org/data/general/cholest.html
Is there evidence cholesterol levels changed between day 2 and day 14?  
**THINK ABOUT RELABELLING DAY AS A FACTOR (WHY?)?**

```{r}
cholesterol <- read.table("http://www.statsci.org/data/general/cholestg.txt", header = T)
cholesterol$day <- as.factor(cholesterol$day)
head(cholesterol)

summary(cholesterol)

ggplot(cholesterol[cholesterol$day %in% c("2", "14"),], aes_string("cholest")) +
  geom_histogram() +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))+
  facet_wrap(~day)

t.test(cholesterol[cholesterol$day == "2", "cholest"], 
       cholesterol[cholesterol$day == "14", "cholest"],
       paired = T)
```

*I used a paired t-test since each individual was tested twice (experimental 
unit); this is an example of blocking!  I had to relabel days as a factor because
otherwise R will do regression (which we haven't covered yet). This is important 
because if we label something we want to be a factor with numbers R will assume
its a number!  After running the test I found a test statistic of t~18~=3.28,
which corresponds to a p-value of .004. Given this I reject the null hypothesis 
that there is no difference among days in average cholesterol levels. The 
alternative hypothesis is there is a difference among days.*