---
title: "8. ANOVA Extensions"
author: "jsg answers"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
---

Before doing this, review the **8. 2-way Anova** lecture set slides from 
https://sites.google.com/view/biostats/lessons/2-way-anovas-and-interactions and
the **8_Blocking_and_interactions.R** script in the lecture files folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats).
Make sure you are comfortable with null and alternative hypotheses and appropriate plots
for all examples.



Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  * **push** updates to github
  
## Examples

### If interaction is significant

Following the memory example from class, read in and check data

```{r}
memory <- read.table("http://www.statsci.org/data/general/eysenck.txt", header = T,
                     stringsAsFactors = T)
str(memory)
```

Let's put younger level first

```{r}
library(plyr)
memory$Age <- relevel(memory$Age, "Younger")
```

and graph

```{r}
library(Rmisc)
function_output <- summarySE(memory, measurevar="Words", groupvars =
                               c("Age", "Process"), na.rm = T)
library(ggplot2)
ggplot(function_output, aes(x=Age, y=Words,color=Process, 
                                   shape = Process)) +
  geom_line(aes(group=Process, linetype = Process), size=2) +
    geom_point(size = 5) +
  ylab("Words remembered")+ 
  xlab("Age") + 
  ggtitle("Process type interacts with age to impact memory")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```

There appears to be some interactions. Let' build a model

```{r}
memory_interactions <- lm(Words ~ Age * Process, memory)
```

and check assumptions.

```{r}
par(mfrow=c(2,2))
plot(memory_interactions)
```

These appear to be met, so look at output

```{r}
library(car)
Anova(memory_interactions, type = "III")
```

Since interaction is significant, analyze subsets. For example, 

```{r}
memory_interactions_young <- lm(Words ~ Process, memory[memory$Age == "Younger",])
plot(memory_interactions_young)
Anova(memory_interactions_young, type = "III")
```

There is a significant difference in words recalled based on process, but why? 
Investigate with post-hoc tests.

```{r}
library(multcomp)
comp_young <- glht(memory_interactions_young, linfct = mcp(Process = "Tukey"))
summary(comp_young)
```

### Blocking example


## Practice

### 1


1. A survey was conducted to see if athletes and non-athletes deal with anger in
the same way.  Data is @

angry <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSaawG37o1ZUEs1B4keIJpZAY2c5tuljf29dWnzqQ0tHNCzfbz85AlWobYzBQ3nPPXJBLP-FWe4BNZB/pub?gid=1784556512&single=true&output=csv", stringsAsFactors = T)

and more information is at 

http://onlinestatbook.com/case_studies/angry_moods.html.

Focus on the following variables:

Sports
1 = athletes, 2 = non-athletes
Gender
1 = males, 2 = females
Expression (AE)
index of general anger expression: 
(Anger-Out) + (Anger-In) - (Control-Out) - (Control-In) + 48

Is there any evidence that gender or athlete status impact how anger is expressed?

```{r}
angry <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSaawG37o1ZUEs1B4keIJpZAY2c5tuljf29dWnzqQ0tHNCzfbz85AlWobYzBQ3nPPXJBLP-FWe4BNZB/pub?gid=1784556512&single=true&output=csv", stringsAsFactors = T)
str(angry)
angry$Gender <- as.factor(angry$Gender)
library(plyr)
angry$Gender <- revalue(angry$Gender, c("1" = "male", 
                                        "2" = "female"))
angry$Sports <- as.factor(angry$Sports)
angry$Sports <- revalue(angry$Sports, c("1" = "athlete",
                                        "2" = "non-athlete"))
library(Rmisc)
anger_summary <- summarySE(angry, measurevar="Anger_Expression", groupvars =
                               c("Sports", "Gender"), na.rm = T)
library(ggplot2)
ggplot(anger_summary, aes(x=Gender, y=Anger_Expression, color=Sports, 
                                   shape = Sports)) +
  geom_point(size = 3) +
  geom_line(aes(group=Sports, linetype =Sports), size=2) +
  geom_errorbar(aes(ymin=Anger_Expression-ci, ymax=Anger_Expression+ci), size=1.5) +
  ylab("Anger level")+ 
  xlab("Gender") + 
  scale_shape_discrete(guide=FALSE)+
  scale_linetype_discrete(guide=FALSE)+
  ggtitle("Anger level among groups")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```
*I first read in and recoded some data for ease and plotting. I then produced a 
plot to consider the null hypotheses that* 

* *the sport an athlete plays does not influence anger level*
* *the gender of an athlete does not influence anger level*
* *the sport an athlete plays and their gender do not interact to influence anger 
level*

```{r}
angry_gender <- lm(Anger_Expression ~ Sports * Gender, angry)
plot(angry_gender)
library(car)
Anova(angry_gender, type = "III")

#remove interaction since not significant
angry_gender <- lm(Anger_Expression ~ Sports + Gender, angry)
plot(angry_gender)
Anova(angry_gender, type = "III") #only differs among those who play sports
```
*I then analyzed the data using factorial ANOVA.
The outcome is continuous 
and both explanatory variables are categorical. The design is also fully randomized. Resildual plots indicated all assumptions were met
(there is no pattern in the residuals and they are normally distributed).
Analysis shows an insignificant interaction (F~1,74~=.04, p=.855) between 
sport and gender, so I removed the interaction term. The reduced model
showed anger levels differed among athletes and non-athletes but not by 
gender. There was no need for post-hoc tests (only 2 levels/groups for
each categorial variable.)*

### 2

2. A professor carried out a long-term study to see how various factors impacted
pulse rate before and after exercise.  Data can be found at 
http://www.statsci.org/data/oz/ms212.txt
With more info at 
http://www.statsci.org/data/oz/ms212.html.
Is there evidence that frequency of exercise
(Exercise column) and gender impact change in pulse rate for students who ran 
(Ran column = 1)?

```{r}
pulse <- read.table("http://www.statsci.org/data/oz/ms212.txt", header = T, 
                    stringsAsFactors = T)
pulse$Exercise <- factor(pulse$Exercise)
library(plyr)
pulse$Exercise <- revalue(pulse$Exercise, c("1" = "high", 
                                            "2" = "moderate", 
                                            "3" = "low"))
pulse$Gender <- factor(pulse$Gender)
pulse$Gender <- revalue (pulse$Gender, c("1" = "male", "2" = "female"))
pulse$change <- pulse$Pulse2 - pulse$Pulse1
change_summary <- summarySE(pulse[pulse$Ran == 1, ], measurevar="change", groupvars =
                               c("Exercise", "Gender"), na.rm = T)

ggplot(change_summary, aes(x=Gender, shape = Exercise, color = Exercise,
                           y=change)) +
  geom_point(size = 3) +
  geom_line(aes(group=Exercise, linetype =Exercise), size=2) +
  geom_errorbar(aes(ymin=change-ci, ymax=change+ci), size=1.5) +
  ylab("Change in pulse \n (beats per minute)") +
  scale_color_discrete(name = "Exercise level")+
  scale_shape_discrete(guide=FALSE)+
  scale_linetype_discrete(guide=FALSE)+
  ggtitle("Change in pulse does \n not differ among groups") +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```
*I first read in and recoded some data for ease and plotting. I then produced a 
plot to consider the null hypotheses that* 

* *exercise level does not influence change in pulse rate*
* *gender does not influence change in pulse rate*
* *gender and exercise level do not interact to influence change in pulse rate*

```{r}
exercise <- lm(change ~ Gender * Exercise, pulse[pulse$Ran == 1, ])
summary(exercise)
Anova(exercise, type = "III")

#rerun without interaction
exercise <- lm(change ~ Gender + Exercise, pulse[pulse$Ran == 1, ])
summary(exercise)
Anova(exercise, type = "III") #no significance
```

*I then analyzed the data using factorial ANOVA.
The outcome is continuous 
and both explanatory variables are categorical. The design is also fully randomized. Residual plots indicated all assumptions were met
(there is no pattern in the residuals and they are normally distributed).
Analysis shows an insignificant interaction (F~2,40~=1.2, p=.21) between 
exercise level and gender, so I removed the interaction term. The reduced model
showed neither gender (F~1,42~=.16, p =.69) or exercise level (F~2,42~=.1, p=.90) 
influenced change in pulse rate, so I failed to reject the related null 
hypotheses.*

### 3


3. (Continuing from question 2) I’m curious if the treatment (running vs not) even had an impact, and if so did it depend on gender. How would you test this?  

```{r}
pulse$Ran <- factor(pulse$Ran)
pulse$Ran <- revalue(pulse$Ran, c("1" = "Ran", "2" = "Sat"))
run_summary <- summarySE(pulse, measurevar="change", groupvars =
                               c("Ran", "Gender"), na.rm = T)

ggplot(run_summary, aes(x=Gender, shape = Ran, color = Ran,
                           y=change)) +
  geom_point(size = 3) +
  geom_line(aes(group=Ran, linetype = Ran), size=2) +
  geom_errorbar(aes(ymin=change-ci, ymax=change+ci), size=1.5) +
  ylab("Change in pulse rate \n (beats per minute)") +
  xlab("Gender")+
  ggtitle("Change in pulse  \n  differed among groups") +
  scale_color_discrete(name = "Ran")+
  scale_shape_discrete(guide=FALSE)+
  scale_linetype_discrete(guide=FALSE)+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

pulse$Ran <- as.factor(pulse$Ran)
levels(pulse$Ran) <- c("ran", "sat")
running <- lm(change ~ Gender * Ran, pulse)
plot(running)
Anova(running, type = "III") 

#consider without interaction
running <- lm(change ~ Gender + Ran, pulse)
Anova(running, type = "III")  #running mattered
summary(running) #running increased pulse
```
*After plotting the data, I used a linear model to test the null hypotheses that*
* *change in pulse rate did not depend on gender*
* *change in pulse rate did not depend on running treatment*
* *there was no interaction between running treatment and gender on change in pulse
rate*

*Analyses indicated all model assumptions were met(
there is little pattern in the residuals and they are fairly normally distributed,
though you may need to rely more on robustness of test here!
) and there is not enough evidence for an interaction between gender and running
treatment on change in pulse rate (F~1,105~=1.3, p=.25). After removing the interaction,
analysis of a reduced model indicates only running treatment (F~1,106~=212, p<.01)
impacts change in pulse rate. Running did increase pulse rate.*