---
title: "9. Regression and correlation"
author: "jsg answers"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
---

Before doing this, review the **9. Linear Regression and Correlation** lecture set slides from 
https://sites.google.com/view/biostats/lessons/linear-regression-and-correlation and
the  **9_Regression_and_correlation.R** script in the lecture files folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats). 
Make sure you are comfortable with null and alternative hypotheses and appropriate plots
for all examples. Also make sure you understand the difference between
correlation and regression.


Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  * **push** updates to github
  
  
## Example

Following the iris dataset from class

```{r}
library(ggplot2)
ggplot(iris, aes(x=Petal.Length, y=Sepal.Length)) +
  geom_point(size = 3) +
  ylab("Sepal Length")+ggtitle("Sepal length increases with petal length")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))+
  xlab("Petal length (cm)") +
  ylab("Sepal length (cm)")


iris_regression <- lm(Sepal.Length ~ Petal.Length, iris)
par(mfrow = c(2,2))
plot(iris_regression)
library(car)
Anova(iris_regression, type = "III")
summary(iris_regression)
```

```{r}
cor.test(~ Sepal.Length + Petal.Length, data = iris)
cor.test(~ Sepal.Length + Petal.Length, data = iris,
         method="spearman")

```


## Practice

### 1
  
  

1.  A professor carried out a long-term study to see how various factors impacted
pulse rate before and after exercise.  Data can be found at 

http://www.statsci.org/data/oz/ms212.txt

With more info at 

http://www.statsci.org/data/oz/ms212.html.

Is there evidence that age, height, or weight impact change in pulse rate for students who ran (Ran column = 1)?  For each of these, how much variation in pulse rate do they explain?


```{r}
pulse <- read.table("http://www.statsci.org/data/oz/ms212.txt", header = T, stringsAsFactors = T)
pulse$change <- pulse$Pulse2 - pulse$Pulse1
#need to make columns entered as numeral change to factor, although it doesn't 
#really matter when only 2 groups (why?)
pulse$Exercise <-as.factor(pulse$Exercise)
pulse$Gender <- as.factor(pulse$Gender)

#age
exercise <- lm(change ~ Age, pulse[pulse$Ran == 1, ])
par(mfrow =c (2,2))
plot(exercise)
require(car)
Anova(exercise, type = "III")
summary(exercise)
```

*First we need to make a column that shows change in pulse rate. We also should 
change Exercise and gender to factors.*  

*For age we note the model meets assumptions (no patterns in residuals and 
residuals follow a normal distribution). We also find no evidence that 
age impacts change (F~1,44~ = .4950, p = 0.49). We fail to reject our null 
hypothesis that there is no relationship between age and change in pulse rate. We
also note that age only explains 1.1% of the variation in change in pulse rate 
(likely due to chance!).*  

```{r}
#weight
exercise <- lm(change ~ Weight, pulse[pulse$Ran == 1, ])
par(mfrow =c (2,2))
plot(exercise)
Anova(exercise, type = "III")
summary(exercise)
```
*For weight we note the model meets assumptions. We also find no evidence that 
weight impacts change (F~1,44~ = .4027, p = 0.53). We fail to reject our null 
hypothesis that there is no relationship between weight and change in pulse rate.
 We
also note that weight only explains 1% of the variation in change in pulse rate 
(likely due to chance!).*    

```{r}
#height
exercise <- lm(change ~ Height, pulse[pulse$Ran == 1, ])
par(mfrow =c (2,2))
plot(exercise)
Anova(exercise, type = "III")
summary(exercise)
```
*For height we note the model meets assumptions. We also find no evidence that 
weight impacts change (F~1,44~ = 1.15, p = 0.29). We fail to reject our null 
hypothesis that there is no relationship between height and change in pulse rate.
 We
also note that age only explains 2.5% of the variation in change in pulse rate 
(likely due to chance!).*   

### 2 

2. (from OZDASL repository, http://www.statsci.org/data/general/stature.html; reference for more information)

When anthropologists analyze human skeletal remains, an important piece of
information is living stature. Since skeletons are commonly based on
statistical methods that utilize measurements on small bones. The following
data was presented in a paper in the American Journal of Physical
Anthropology to validate one such method.
Data is available @

http://www.statsci.org/data/general/stature.txt

as a tab-delimted file (need to use read.table!)
Is there evidence that metacarpal bone length is a good predictor of
stature? If so, how much variation does it account for in the response
variable?

```{r}
height <- read.table("http://www.statsci.org/data/general/stature.txt", 
                     header = T, stringsAsFactors = T)
head(height)
metacarp_relationship <- lm(Stature ~ MetaCarp, height)
plot(metacarp_relationship)
Anova(metacarp_relationship, type = "III")
summary(metacarp_relationship)
```
*To consider the relationship among these continuous variables, we used linear regression. 
Analysis of model assumptions suggest assumptions are met, although the dataset is 
small.  Analysis suggests there is a significant positive relationship between 
metacarpal length and stature (F~1,7~ = 19.19, p = 0.003). The R^2^ value indicates
that metacarpal length explains 73% of the variation in stature. Coefficients 
indicate that stature increases with increasing metacarpal length.*

### 3


3. Data on medals won by various countries in the 1992 and 1994 Olympics is
available in a tab-delimited file at 

http://www.statsci.org/data/oz/medals.txt

More information on the data can be found at: 

http://www.statsci.org/data/oz/medals.html

Is there any relationship between a country’s population and the total
number of medals they win?

```{r}
medals <- read.table(header = T, "http://www.statsci.org/data/oz/medals.txt", 
                     stringsAsFactors = T)
head(medals)
medals$total <- medals$Summer + medals$Winter
population_medals <- lm(total ~ Population, medals)
plot(population_medals)
summary(population_medals)
Anova(population_medals, type = "III")
cor.test(~total + Population, medals, method = "spearman")
```
*There is a high leverage point in the dataset (row 4), but residuals appear to 
be fairly normally distributed and little structure exists in the graph of Residuals 
vs. Fitted Values.  Analysis using linear regression suggests a significant (
F~1,60~ = 10.45, p = 0.002) positive relationship between population size and 
medal count that explains ~15% of the variation in the response variable.  Rank-
correlation analysis also indicated this relationship exists.*

### 4


4. Continuing with the Olympic data, is there a relationship between the
latitude of a country and the number of medals won in summer or winter
Olympics?

```{r}
#still using medals
summer_medals <- lm(Summer ~ Latitude, medals)
plot(summer_medals)
Anova(summer_medals, type = "III")
summary(summer_medals)

winter_medals <- lm(Winter ~ Latitude, medals)
plot(winter_medals)
Anova(winter_medals, type = "III")
summary(winter_medals)
```

*Visual analysis of residuals from both models show some structure in the residual
and deviations from normality, but we continue on with linear regression given the 
small sample size.  Both summer and winter medal counts are positively (surpisingly)
and significantly (both p <.05)
related to latitude, with latitude explaining ~17% of the variation in winter medal
count and ~8% of the data in summer medal count.*

### 5


5. Data on FEV (forced expiratory volume), a measure of lung function, can
be found at 

http://www.statsci.org/data/general/fev.txt

More information on the dataset is available at 

http://www.statsci.org/data/general/fev.html.

Is there evidence that FEV depends on age or height? If so, how do these
factors impact FEV, and how much variance does each explain? 

```{r}
fev <- read.table("http://www.statsci.org/data/general/fev.txt", header = T, 
                  stringsAsFactors = T)
head(fev)
fev_height <- lm(FEV ~ Height, fev)
plot(fev_height)
Anova(fev_height, type = "III")
summary(fev_height)
```
Model assumptions appear to be met. Height appears to have a positive relationship
with FEV (F~1,652~ = 1995, p<.001).  

```{r}
fev_age <- lm(FEV ~ Age, fev)
plot(fev_age)
Anova(fev_age, type = "III")
summary(fev_age)
```

Model assumptions appear to be met. Age appears to have a positive relationship
with FEV (F~1,652~ = 872.2, p<.001). 

### 6 


6. Continuing with the FEV data, produce plots that illustrate how height, age, 
and gender each impact FEV.  

```{r}
library(ggplot2)
#age plot####
ggplot(fev, aes(x=Age, y=FEV)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm") +
  ylab("FEV (L)")+ggtitle("FEV increases with age")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

#height plot####
ggplot(fev, aes(x=Height, y=FEV)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm") +
  ylab("FEV (L)")+ggtitle("FEV increases with height")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))



#gender plot ####

#bar chart with error bars ####
library(Rmisc)
function_output <- summarySE(fev, measurevar="FEV", groupvars =
                               c("Sex"))

ggplot(function_output, aes(x=Sex, y=FEV)) +
  geom_col(size = 3) +
  ylab("FEV (L)") +
  ggtitle("FEV is higher in males ")+
  geom_errorbar(aes(ymin=FEV-ci, ymax=FEV+ci), size=1.5) +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

```
