---
title: "2. Estimates and ggplot2"
author: "jsg"
date: "9/17/2020"
output: github_document
---
Before doing this, review the **Estimation and Probability** lecture set slides from 
https://sites.google.com/view/biostats/bioenv-2100/estimation-and-probablity and
the  **2_estimation_lecture.R** script in the lecture_files folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats).


Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  * **push** updates to github

## Using ggplot2
Let’s return to the mammal sleep dataset that we left off with last week. 
Load the dataset
```{r}
sleep <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/sleep.csv", stringsAsFactors = T)
#need to use stringsAsFactors to make characters read in as factors
```
Last time you used the built-in plot functions to do some plots. Let’s replace 
those with ggplot2 and do some more.

1. First plot how TotalSleep is explained by BrainWt (remember the issues with 
the data).  Use ggplot2 to plot the relationship.


```{r}
library(ggplot2)
ggplot(sleep[sleep$BrainWt <1000, ], aes(x=BrainWt, y = TotalSleep)) +
  geom_point(size = 4) +
  #below here is ylabel, xlabel, and main title
  ylab("Average hours spent \n sleeping daily") +
  xlab("Brain weight (g)") +
  ggtitle("Time spent sleeping \n decreases with brain \n weight") +
  #theme sets sizes, text, etc
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32),
        # change plot background, grid lines, etc (just examples so you can see)
        panel.background = element_rect(fill="white"),
        panel.grid.minor.y = element_line(size=3),
        panel.grid.major = element_line(colour = "black"),
        plot.background = element_rect(fill="gray"),
        legend.background = element_rect(fill="gray"))
```

2. Next color code each plot point by whether or not its a primate.  In order 
to do this you can use the Primate column or (following class code) make a new 
column called Taxa to represent the information (hint:search for “ revalue”). 
Make sure axes are well-labeled.

```{r}
library(plyr)
sleep$Taxa <- revalue(sleep$Primate, c(Y = "Primate", N = "Non-primate"))
sleep$Taxa <- relevel(sleep$Taxa, "Primate")

ggplot(sleep[sleep$BrainWt <1000, ], aes(x=BrainWt, y = TotalSleep)) +
  geom_point(aes(colour=Taxa), size = 4) +
  #below here is ylabel, xlabel, and main title
  ylab("Average hours spent \n sleeping daily") +
  xlab("Brain weight (g)") +
  ggtitle("Time spent sleeping \n decreases with brain \n weight") +
  #scale commands help with legends
  scale_colour_manual(name="Type of mammal",values = c("#FFA373","#50486D")) +
  #theme sets sizes, text, etc
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32),
        # change plot background, grid lines, etc (just examples so you can see)
        panel.background = element_rect(fill="white"),
        panel.grid.minor.y = element_line(size=3),
        panel.grid.major = element_line(colour = "black"),
        plot.background = element_rect(fill="gray"),
        legend.background = element_rect(fill="gray"))
```


3. Let’s work with histograms.
* What type of variation do we see in total time spent sleeping? Create a 
histogram to explore this issue.


```{r}
ggplot(sleep
       , aes(x=TotalSleep)) +
  geom_histogram() +
  xlab("Total sleep (hours per day")+ggtitle("Variation in sleep levels")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```
* Facet the graph you created based on whether or not the animal is a primate 
(Primate column).

```{r}
ggplot(sleep
       , aes(x=TotalSleep)) +
  geom_histogram() +
  xlab("Total sleep (hours per day")+ggtitle("Variation in sleep levels")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))+ 
  facet_wrap(~Taxa)
```

* Now only graph the data for primates.

```{r}
ggplot(sleep[sleep$Taxa == "Primate",]
       , aes(x=TotalSleep)) +
  geom_histogram() +
  xlab("Total sleep (hours per day")+ggtitle("Variation in sleep levels")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```

4. Develop a properly-labeled bar graph with error bars to explore how total 
sleep changes with 
* Primate (relabeled as yes/no as Primate/Non-Primate; note 
there are multiple ways to do this!) – use a 95% confidence interval for the bar

```{r}
#use summarySE function from Rmisc package
sleep$Primate <- revalue(sleep$Primate, c(Y = "Yes", N = "No"))
sleep$Primate <- relevel(sleep$Primate, "No")
library(Rmisc)
sleep_by_primate <- summarySE(sleep, measurevar = "TotalSleep", groupvars = "Primate", na.rm = T)
#look at it
sleep_by_primate
library(ggplot2)
ggplot(sleep_by_primate
       , aes(x=Primate, y=TotalSleep)) +
  geom_col(size = 3) +
  geom_errorbar(aes(ymin=TotalSleep-ci, ymax=TotalSleep+ci), size=1.5) +
  ylab("Total sleep \n (hours per day")+ 
  xlab("Primate?")+ 
  ggtitle("Sleep across different taxa")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32)) 

```

* Predation risk (as a factor!) – use 1 standard error for the bar. Note the difference!

```{r}
sleep$Predation <- as.factor(sleep$Predation)
sleep_by_predation <- summarySE(sleep, measurevar = "TotalSleep", 
                                groupvars = "Predation", na.rm = T)
#look at it
sleep_by_predation
require(ggplot2)
ggplot(sleep_by_predation
       , aes(x=Predation, y=TotalSleep)) +
  geom_col(size = 3) +
  geom_errorbar(aes(ymin=TotalSleep-se, ymax=TotalSleep+se), size=1.5) +
  ylab("Total sleep \n (hours per day)") + 
  ggtitle("Sleep across different \n predation levels")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

```

## Estimates and Certainty Concepts

5. What does a 95% confidence interval mean?

*A 95% confidence interval means the true population parameter value will be in 
the created interval 95% of the time we create it.*

6. To make sure you understand the ideas of sampling, confidence intervals, and the central limit theorem, review the visualizations produced by UBC @ http://www.zoology.ubc.ca/~whitlock/Kingfisher/CLT.htm.  

*Key outcomes here are understanding that, regardless of the distribution of the 
data, the distribution of the means of the data (what we typically consider), will
follow a normal distribution if the sample size is large enough.*

7. For this question you'll need the 
central_limit_theorem.R script from the 
[code_examples folder](https://github.com/jsgosnell/CUNY-BioStats/tree/master/code_examples).

To show you how well the central limit theorem (and functions) works, you can also examine 6
distributions that display traits we described at the beginning of lab using the
central_limit_theorem.R script from the Examples folder on github (or download here);
you may need to download the VGAM package.  Source it and and then use Rstudio to review the plots
it produces. Decide what characteristics (truncation, skewness, kurtosis, sample size) speed vs.
slow the convergence of the sample mean upon normality.  The reviewed distributions are:

* Normal(Z) (0,1)  {no Kurtosis / no skewness / no truncation}
* Double exponential (0,2) {high Kurtosis / no skewness / no truncation}
* Uniform(0,1) {moderate Kurtosis / no skewness / double truncation}
* Exponential(1,1)  {high asymmetric Kurtosis / high skewness / single truncation} 
* Chi-square(df=4) {low Kurtosis / moderate skewness / single truncation} 
* Binomial distribution (p=.7) {discrete distribution]

For this question, its easiest to just source the main file and see what happens.
When you source a script, it is run in R without showing any console output
(but graphs and objects are still produced!).  Try
*source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/central_limit_theorem.R")*.  You'll also need to have the VGAM library installed.

```{r}
#make sure you have VGAM library installed
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/central_limit_theorem.R")
```

*This allows you to visualize what we noted in question 6 and begin to develop a
sense for what "large sample size" and "odd distribution" really mean.*