---
title: "7. ANOVAs"
author: "jsg"
date: "9/22/2020"
output: github_document
---
Before doing this, review the **7. ANOVA and multiple comparisons** lecture set slides from 
https://sites.google.com/view/biostats/bioenv-2100/anova-and-multiple-comparisons and
the  **7_ANOVAs.R** script in the lecture files folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats).
Make sure you are comfortable with null and alternative hypotheses for all examples.


Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  * **push** updates to github
  
## Just for practice 
*Do this problems using ANOVA, Kruskal-Wallis, and 
bootstrapping methods.  Make sure you can plot the data and carry out multiple 
comparison methods as needed. Also be sure to understand the use of coefficients 
and adjusted R^2^ values and where to find them.* 

1. Use the iris dataset in R to determine if petal length differs among species.  

```{r}
#plot
library(Rmisc)

function_output <- summarySE(iris, measurevar="Petal.Length", groupvars =
                               c("Species"))
library(ggplot2)
ggplot(function_output, aes(x=Species, y=Petal.Length)) +
  geom_col(aes(fill=Species), size = 3) +
  geom_errorbar(aes(ymin=Petal.Length-ci, ymax=Petal.Length+ci), size=1.5) +
  ylab("Petal Length (cm)")+ggtitle("Petal Length of \n various iris species")+
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

petal <- lm(Petal.Length ~ Species, iris)
plot(petal)
library(car)
Anova(petal, type = "III")
#compare to
summary(petal)
library(multcomp)
comp_cholest <- glht(petal, linfct = mcp(Species = "Tukey"))
summary(comp_cholest)

#kw approach
petal <- kruskal.test(Petal.Length ~ Species, iris)
pairwise.wilcox.test(iris$Sepal.Length, 
                          iris$Species, 
                          p.adjust.method="holm")

#bootstrap
library(WRS2)
t1waybt(Petal.Length~Species, iris)
bootstrap_post_hoc <- mcppb20(Petal.Length~Species, iris)
#use p.adjust to correct for FWER
p.adjust(as.numeric(bootstrap_post_hoc$comp[,6]), "holm")
```

*Answer: We used an ANOVA (a special case of linear models) to investigate how a numerical response variable 
differed among 3 groups.  This was appropriate as evidenced by the residual plots
(there is no pattern in the residuals and they are normally distributed),
but other methods are demonstrated as well.
Using an ANOVA, we found F~2,147~= 1180.2, which led to a p-value of <.001.  Given
this, I reject the null hypothesis there is no difference among mean measurements
for each species.  
Post-hoc testing indicated all species significantly differed from all others 
(all p <.05)
using a Tukey approach to control for family-wise error rate. Kruskal-Wallis and
bootstrapping approaches led to similar conclusions.*

## t-tests are special cases of ANOVAs, which are special cases of linear models!

2. Data on plant heights (in cm) for plants grown with a new and old formulation 
of fertilizer can be found at

https://docs.google.com/spreadsheets/d/e/2PACX-1vSUVowOKlmTic4ekL7LSbwDcqrsDSXv5K_c4Qyfcvz1lLE1_iINmGzy0zMGxY7z5DImlUErK4S2wY7Y/pub?gid=0&single=true&output=csv.

Analyze this data using the t.test function and the lm function to convince yourself the outcome is the same.

```{r}
fertilizer <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSUVowOKlmTic4ekL7LSbwDcqrsDSXv5K_c4Qyfcvz1lLE1_iINmGzy0zMGxY7z5DImlUErK4S2wY7Y/pub?gid=0&single=true&output=csv",
                       stringsAsFactors = T)
#note use of var.equal!  assumption of ANOVAs
t.test(height ~ fertilizer, fertilizer, var.equal = T)
fert_lm <- lm(height ~ fertilizer, fertilizer)
plot(fert_lm)
summary(fert_lm)
require(car)
Anova(fert_lm, type = "III")
```

*Answer: t-tests and ANOVA (lm) approaches yield the same results. Note for the 
tests to match exactly we have to assume equal variances among groups for the 
t-tests. In both we reject the null hypothesis of no difference among mean height
of plants based on fertilizer.  Notice the t statistic (2.9884) is the square root
of the F statistic (8.931).  The t distribution corresponds to the F with only 1
df in the numerator (so its not listed!).*

## For the following questions, pick the appropriate method for analyzing the question.  Use a plot of the data and/or model analysis to justify your decision.  Make sure you can carry out multiple comparison methods as needed. Also be sure to understand the use of coefficients and adjusted R^2^ values and where to find them.

3. Data on sugar cane yield for multiple fields is available using 

read.table("https://docs.google.com/spreadsheets/d/e/2PACX-1vRjstKreIM6UknyKFQCtw2_Q6itY9iOAVWO1hUNZkBFL8mwVssvTevqgzV22YDKCUeJq0HBDrsBrf5O/pub?gid=971470377&single=true&output=tsv", 
header = T, stringsAsFactors = T)

More info on the data can be found at 
http://www.statsci.org/data/oz/cane.html.
Is there evidence that location (DistrictPosition column) impacts yield 
(Tonn.Hect column)?  If so, which areas are driving this distance?

```{r}
cane <- read.table("https://docs.google.com/spreadsheets/d/e/2PACX-1vRjstKreIM6UknyKFQCtw2_Q6itY9iOAVWO1hUNZkBFL8mwVssvTevqgzV22YDKCUeJq0HBDrsBrf5O/pub?gid=971470377&single=true&output=tsv", header = T, stringsAsFactors = T)
summary(cane)
cane_summary <- summarySE(cane, measurevar="Tonn.Hect", groupvars =
                               c("DistrictPosition"))

ggplot(cane_summary, aes(x=DistrictPosition, y=Tonn.Hect)) +
  geom_col(size = 3) +
  geom_errorbar(aes(ymin=Tonn.Hect-ci, ymax=Tonn.Hect+ci), size=1.5) +
  ylab("Production (tonnes per hectare)") +
  xlab("District Position") +
  ggtitle("Production differs \n among locations") +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

impact_district <- lm(Tonn.Hect ~ DistrictPosition, cane)
summary(impact_district)
plot(impact_district)#not really normal...lets bootstrap
require(WRS2)
t1waybt(Tonn.Hect ~ DistrictPosition, cane)
mcppb20(Tonn.Hect ~ DistrictPosition, cane)
p <- mcppb20(Tonn.Hect ~ DistrictPosition, cane)
p.adjust(as.numeric(p$comp[,6]), "holm")


#compare to lm apporach
require(car)
Anova(impact_district, type = "III")
require(multcomp)
comp_district <- glht(impact_district, linfct = mcp(DistrictPosition = "Tukey"))
summary(comp_district)

```

*Answer: For this analysis I used a bootstrap approach as the residual plots
suggested a non-normal distribution.  Analysis revealed a test statistics of 
16.52 and p-value of 0, so I reject the null hypothesis of no difference among
Districts.  since I rejected the null hypothesis, I have to use post-hoc tsts
to determine which groups are different than the others.  
Post-hoc tests reveal all district areas differ from each other 
except for
south and east, south and central, and east and central (using sequential FDR 
to control for family-wise error rate.)*

*Note that a linear model does lead to slightly different findings regarding which
districts differ from which others.*

4. Data on FEV (forced expiratory volume), a measure of lung function, can be 
found at 

http://www.statsci.org/data/general/fev.txt

More information on the dataset is available at 

http://www.statsci.org/data/general/fev.html.

Is there evidence that FEV depends on gender? If so, which gender has the higher
FEV score?  How much variance does gender explain? 

```{r}
fev <- read.table("http://www.statsci.org/data/general/fev.txt", header = T,
                  stringsAsFactors = T)
fev_summary <- summarySE(fev, measurevar="FEV", groupvars =
                               c("Sex"))

ggplot(fev_summary, aes(x=Sex, y=FEV)) +
  geom_col(size = 3) +
  geom_errorbar(aes(ymin=FEV-ci, ymax=FEV+ci), size=1.5) +
  ylab("FEV (liters)") +
  xlab("Sex") +
  ggtitle("FEV differs \n among males and females") +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))


fev_gender <- lm(FEV ~ Sex, fev)
plot(fev_gender) #anova is fine
Anova(fev_gender, type = "III")
summary(fev_gender)
```

*I used an ANOVA (or linear model, or t-test, here, all the same since 2 groups!)
to consider the impact of sex on FEV.  This was appropriate as evidenced by the
residual plots
(there is no pattern in the residuals and they are normally distributed).  Results indicate there is a difference among sexes (F~1,652~ = 29.607, p<.001). There is 
no need for post-hoc tests here since there are only 2 groups being considered.  
Coefficients related to the groups (note female is replaced by intercept here, 
and the SexMale coefficient is relative to that) indicates that males have a 
higher FEV on average.  Graphs also show this relationship.*

5. A professor carried out a long-term study to see how various factors impacted
pulse rate before and after exercise.  Data can be found at 

http://www.statsci.org/data/oz/ms212.txt

With more info at 

http://www.statsci.org/data/oz/ms212.html.

Is there evidence that frequency of exercise (Exercise column) impacts resting
pulse rate (Pulse1 column)?

```{r}
pulse <- read.table("http://www.statsci.org/data/oz/ms212.txt", header = T, 
                    stringsAsFactors = T)
pulse$Exercise <- factor(pulse$Exercise)
library(plyr)
pulse$Exercise <- revalue(pulse$Exercise, c("1" = "high", 
                                            "2" = "moderate", 
                                            "3" = "low"))

exercise_summary <- summarySE(pulse, measurevar="Pulse1", groupvars =
                               c("Exercise"), na.rm = T)

ggplot(exercise_summary, aes(x=Exercise, y=Pulse1)) +
  geom_col(size = 3) +
  geom_errorbar(aes(ymin=Pulse1-ci, ymax=Pulse1+ci), size=1.5) +
  ylab("Resting pulse (beats per minute)") +
  xlab("Exercise level")+
  ggtitle("Resting pulse decreases \n with exercise") +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
initial <- lm(Pulse1 ~ Exercise, pulse)
plot(initial)
summary(initial)
Anova(initial, type = "III") 
```

*I used a linear model to test for difference in mean resting pulse rates based 
on group. 
This was appropriate as evidenced by the
residual plots
(there is no pattern in the residuals and they are normally distributed). 
Analyses suggest there is no difference in resting pulse rate among groups
(F~2,106~ = 2.8, p = .07), so I fail to reject the null hypothesis that a 
difference does not exist in mean resting pulse rate based on groups. The 
alternative hypothesis is that a 
difference does  exist in mean resting pulse rate based on groups.*

6. Continuing with the pulse dataset, is there evidence frequency of exercise
impacts change in pulse rate?  Note we should only consider change for students 
who ran between the first and second pulse measures (Ran column = 1). 

```{r}
pulse$change <- pulse$Pulse2 - pulse$Pulse1

change_summary <- summarySE(pulse[pulse$Ran == 1, ], measurevar="change", groupvars =
                               c("Exercise"), na.rm = T)

ggplot(change_summary, aes(x=Exercise, y=change)) +
  geom_col(size = 3) +
  geom_errorbar(aes(ymin=change-ci, ymax=change+ci), size=1.5) +
  ylab("Resting pulse \n (beats per minute)") +
  xlab("Exercise level")+
  ggtitle("Change in pulse does \n not differ among groups") +
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))

change <- lm(change ~ Exercise, pulse[pulse$Ran == 1, ])
plot(change)
Anova(change, type = "III")
```

*I used a linear model to test for difference in change in pulse rates (before 
and after running) based 
on group. 
This was appropriate as evidenced by the
residual plots
(there is no pattern in the residuals and they are normally distributed). 
Analyses suggest there is no difference in resting pulse rate among groups
(F~2,43~ = .09, p = .90), so I fail to reject the null hypothesis that a 
difference does not exist in change in pulse rates (before 
and after running) based 
on group.  The 
alternative hypothesis is that a 
difference does  exist in change in pulse rates (before 
and after running) based 
on group.*

