---
title: "Binomial"
author: "jsg"
date: "10/21/2020"
output: github_document
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ZOMBIES!

You wake up one morning and find that zombies have taken over your neighborhood (bummer).
One idea is to use biocontrol to reduce zombie attacks. A zombie parasite is 
identified.  Test plots (neighborhoods!) are subjected to one or two releases of 
the parasite at low, medium, or high densities. Past evidence suggests that the 
parasite should infect 70% of the zombies, but some citizens are concerned it won't 
work in New York due to lower temperatures.  In a random sample, 7 of 12 zombies 
are infected with the parasite.  

* What is the observed rate of infection in the sample?
```{r}
7/12
```
*The observed rate of infection is .58 (58%).*

* Is there evidence the parasite infects zombies 
at a different rate in New York?
** If so, why
** If not, how do you explain any discrepancy between the observed rate of infection
and your results?

Make sure your answer includes

* null hypothesis
* alternative hypothesis
* explanation for test you will use
* results from statistical test
* clear explanation of how results relate to your stated hypotheses
* confidence interval for your estimate 

```{r}
binom.test(7,12, .7)
library(binom)
binom.confint(7,12)
```
*I analyzed the data using a binomial test looking for any deviations from 70% 
infection rates.  Using this method I found a p-value of .3614, so I fail to reject
the null hypothesis that the infection rate is 70%.  The alternative hypothesis is
that the infection rate is not equal to 70%.  Although the rate I observed
was 58%, the small sample size means this isn't that unlikely to occur by chance if the
true infection rate is 70%. The small sample size means the 95% confidence interval
for our estimate of infection rate is .43-.72 (using Agresti-Coul estimates). 
We need more zombies to truly assess the population for
any differences in infection rate.*





