---
title: "Quiz 3"
author: "jsg"
date: "10/21/2020"
output: github_document
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ZOMBIES!

You wake up one morning and find that zombies have taken over your neighborhood (bummer).

Evidence also suggests  that temperature may influence propensity for zombie 
attacks, with zombies preferring warmer bodies.  Long-term studies suggest that 
average human temperature is 36.57 C.  
Unfortunately, measuring temperatures before/after attacks is hard, so data on 
5 survivors of attacks have been recorded. They were: 36.12, 36.37, 35.7, 36.01, 36.9.  

1. Is there any evidence that temperature influences zombie attacks? 

```{r}
temp <- c(36.12, 36.37, 35.7, 36.01, 36.9)
t.test(temp, mu = 36.57, alternative = "greater")
```

*Although this is a small sample size, we know temperature typically follows a 
normal distribution.  For this reason, and given that bootstrapping was not an option,
I used a t-test to test the null hypothesis that the temperature of zombie attack 
victims is less than or equal to 36.57. I used a sided-test because we wanted to 
know if zombies prefer warmer bodies.  The alternative hypothesis, therefore, is
that the temperature of zombie attack victims is greater than 36.57. Using this 
approach I found t~4~=-1.74, leading to a p-value of .9216. Since this is greater 
than the alpha level we typically use (.05), I fail reject the null hypothesis.*

2. How would knowing the standard deviation of human temperature is 0.42 change
your methods and results?

```{r}
library(BSDA)
z.test(temp, mu = 36.57, sigma.x = .42, alternative = "greater")
```

*If I knew the standard deviation of human temperature I could use a z-test 
instead to test the same hypothesis.  I would still fail to reject the null
hypothesis, which makes sense.*


Make sure your answers include

* null hypothesis
* alternative hypothesis
* explanation for test you will use
* results from statistical test
* clear explanation of how results relate to your stated hypotheses
* confidence interval for your estimate


