---
title: "Quiz 4"
author: "jsg"
date: "11/16/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Vaccine effectiveness!

A drug company wants to make sure the vaccine they are developing has similar 
effectiveness for multiple groups.  Out of 10,000 patients, they find the following 
effectivess among groups (labelled with letters to prevent bias!)

 *Outcome*  | Success | Failure
  -| ---     | --
 A |  4600   | 400 
 B |  2750   | 250
 C |  800    | 200
 D |  495    | 505
 
 
1. What is the overall effectiveness of the vaccine?

```{r}
(4600+2750+800+495)/10000
```

*The overall effectiveness of the vaccine (among all users) could be measured as
`r (4600+2750+800+495)/10000 * 100`%. However, we could also consider it as 
`r (4600/5000+2750/3000+800/1000+495/1000)/4 ` given the unequal sample sizes.*

2. Is there any evidence for differences among groups?

```{r}
effectiveness <-matrix(c(4600, 400, 2750, 250, 800, 200, 495, 505), ncol = 2, byrow = T)
colnames(effectiveness) <- c("Success", "Failure")
rownames(effectiveness) <- c("A","B", "C", "D")
#check it
effectiveness
chisq.test(effectiveness)
library(rcompanion)
pairwiseNominalIndependence(effectiveness, compare = "row", method = "holm")
```

*I tested for a difference among groups using a chi^2 test.  This led to a test 
statistics of chi^2~3~ = 1402.2, which corresponds to a p-value of <.001. I thus 
reject the null hypothesis of no difference in effectiveness among groups 
(the alternative is that there is a difference in effectiveness among groups). Given
this result, I used a post-hoc test to compare groups.  Data (using the
p.adj.Chisq column) indicate that groups A and B do not differ from each other, but all others do.*

```{r}
effectiveness_t <-matrix(c(4600, 400, 2750, 250, 800, 200, 495, 505), nrow = 2, byrow = F)
rownames(effectiveness_t) <- c("Success", "Failure")
colnames(effectiveness_t) <- c("A","B", "C", "D")
#check it
effectiveness_t
chisq.test(effectiveness_t)
library(rcompanion)
pairwiseNominalIndependence(effectiveness_t, compare = "col", method = "holm")
```
3. How would you related your findings from question 2 to your answer to question 1? 
Why does it matter (what you would you recommend or do to follow-up)?

*Even though the overall vaccine was 86% effective, this was heavily influenced
by data from groups a and b. Further analysis may be needed to consider effects on
other groups.*

Make sure your answers include

* null hypothesis
* alternative hypothesis
* explanation for test you will use
* results from statistical test (including post-hoc tests if needed!)
* clear explanation of how results relate to your stated hypotheses


