---
title: "Quiz 9"
author: "jsg"
date: "11/18/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Vaccine effectiveness

An earlier study (Quiz 8) focused on a relationship between dosage and systolic blood pressure).  A follow-up wanted to consider if ethnicity impacted this 
relationship.  Data were collected from 4 different groups (labeled A-D here) on 
vaccine dosage level and systolic blood pressure. Data can be found at

```{r eval=F, include=F}
set.seed(26)
dosage_group <- data.frame(dose_micrograms = c(rep(0,40),rep(25,40), rep(50,40), 
                                         rep (75, 40), rep (100,40)),
                     group = rep(c(rep("A", 10), rep("B", 10), 
                                   rep("C",10), rep("D",10)), 5),
                     systolic = NA)
dosage_group[dosage_group$group %in% c("A","B"), "systolic"] =
  120+.35*dosage_group[dosage_group$group %in% c("A","B"), "dose_micrograms"] +
  rnorm(100,0,3)
dosage_group[dosage_group$group %in% c("C"), "systolic"] =
  110+.35*dosage_group[dosage_group$group %in% c("C"), "dose_micrograms"] + 
  rnorm(50,0,3)
dosage_group[dosage_group$group %in% c("D"), "systolic"] = 
  100+.35*dosage_group[dosage_group$group %in% c("D"), "dose_micrograms"] + 
  rnorm(50,0,3)
write.csv(dosage_group, "dosage_group.csv", row.names = F)
```

```{r}
dosage_group <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSXfNA4nl1SRK6BDxfJqo7izW254hxP8N4tjK9snXZngfgQFJ3aeKlg5EAAWPfrStv_61e3FKsAOjjx/pub?gid=73925069&single=true&output=csv", stringsAsFactors = T)
```

How would you evaluate the results?

Make sure your answers include

* null hypothesis

*We have three null hypotheses here*:

* *There is no difference in systolic blood pressure based on dosage*

* *There is no difference in systolic blood pressure based on group*

* *There is no interaction between group and dosage agent on systolic blood 
pressure* 

* alternative hypothesis

*We have three alternative hypotheses here*:

* *There is a difference in systolic blood pressure based on dosage*

* *There is a difference in systolic blood pressure based on group*

* *There is an interaction between group and dosage agent on systolic blood 
pressure* 

* graph that clearly displays the data and represents these hypotheses

```{r}
library(ggplot2)
ggplot(dosage_group, aes(x=dose_micrograms, y=systolic, shape = group, 
                         color = group)) +
         geom_point()+
         geom_smooth(method="lm")+
         xlab("Vaccine dose (micrograms)")+
         ylab("Systolic blood pressure")+
         ggtitle("Systolic blood pressure increases as vaccine dosage increases,
                 but impacts differ among groups")
```

* explanation for test you will use

*I will use an ANCOVA here. The explanatory variables combine a continuous and 
categorial variable, the outcome is continuous and should follow a normal 
distribution (so should meet model assumptions).*


* results from statistical test (including post-hoc tests if needed and indication
of how much variation the model explains!)

```{r}
impacts_group <- lm(systolic~dose_micrograms*group, dosage_group)
plot(impacts_group)
library(car)
Anova(impacts_group, type = "III")
```

*I first ensured model assumptions were met. Plots show no patterns in residuals
and a relatively normal distribution, which means they are. Analysis indicates no significant interaction between dose and group on  systolic blood pressure
(F~3,192~=.87,p = .46), so I fail to reject the null hypothesis that an interaction
does note exist. Given this, I remove the interaction from the model.*

```{r}
impacts_group <- lm(systolic~dose_micrograms+group, dosage_group)
plot(impacts_group)
Anova(impacts_group, type = "III")
library(multcomp)
group_compare <-glht(impacts_group, linfct = mcp(group = "Tukey"))
summary(group_compare)
summary(impacts_group)
```
*Analysis of the reduced model suggests that both dosage (F~1,195~=3232, p <.01)
and group (F~3,195~=493, p <.01) have a 
significant impact on systolic blood pressure, so I reject both related null 
hypotheses. Since there are more than 2 groups, I used post-hoc tests to determine
which groups were different. Analysis using Tukey tests shows that groups A and B
do not differ, but all other possible pairs are significantly different.  Full
model analysis indicates dosage level and group
explains 96% of the variation in blood pressure (this an example!).*

* clear explanation of how results relate to your stated hypotheses

*noted above.*