---
title: "1b. Intro to Rmd files and literate programming"
author: "stephen's example "
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true

---

Rmd files differ from R files in that they combine regular text with code chunks.
This is a code chunk

```{r}
print("this is a chunk")
```

Code chunks combine code with output. When combined with regular text/prose, 
this makes it easier to produce a range of documents.  You set the output in 
the YAML header (the stuff between the 3 dashes you see at top of this document).

After you write the file, you **Knit** it to turn the Rmd file into the selected 
output.  Try it now.  Note the first time you do this in a project you may be
prompted to install a number of packages!  If you are using a webservice you may
also need to allow pop-ups in your browser. Don't be surprised if a new window 
pops up (it should). 

![The knit button turns your .rmd file into other products](https://lh3.googleusercontent.com/pw/ACtC-3dlSoGJDHtdGqEBr8L2X-yqZ-08Z95RHUMvaxHqF9EOFcBnqtamYMAWOr75mohUSL_KvWtBTt-u4KrdoHgceHc-sZiViw6l9ZqEQToLIsy6AwvQIQMrJgLbtXfV6gNLDgQvgT3N7aq9pk9-x5ugpegjYA=w378-h109-no)


The **Knit** button saves the .Rmd file and renders a new version whose output 
depends on what you selected in the header. Here we have html_document,
so if everything works a preview of a webpage like document should appear. The file
also produces a github friendly .md file. **This means you should only edit the 
Rmd file (leave the md and output files alone! They are automatically produced
any changes you make there will be overwritten by your next knit)**.  

When you **Knit** a file, it runs in a totally new R instance. this means anything
you only added in your instance (like working in the console) won't be available.
In other words, its the best way to see what a "new" user gets when they use your
code.

however, you don't have to knit the file every time. if you just want to see output,
note you can press the green button next to an R chunk.

![The green arrows just runs the chunk in the console and shows the output](https://lh3.googleusercontent.com/pw/AM-JKLUYgHbhk7YzhXdAZwV-fvLFlnOc4IcCMwt6U21qsHP7sXcjQ5xDL86NewZo2THSGAveP0Y1cL2PP4yysUTLn4N6iXoO6B1h_8RtAlqmNONY2W5V_j_4hqtQ8d3GhroTNJewT3oEqSVA-Vjh4IkDRqE-pw=w784-h73-no?authuser=0)


```{r}
print("this is a chunk")
```

Now we'll start changing the file to show you how rmarkdown works.
First, amend the file by replacing the **NAME** and **DATE** spots in the header
(top of the
file between the --- markers) with your name and the real date.  Then **Knit** the
file again. You should see your name in the new preview. 

Rstudio has a **Markdown Quick Reference** guide (look under the help tab), but 
some general notes. 

* Pound/Hashtag signs denote headers
* you can surround something double asterisks for bold or single asterisks for italics
* lists are denoted by numbers or asterisks at beginning of line (followed by space!)
  * and can be indented for sublevels
* R code can be done inline, but is generally placed in stand-alone chunks
  * these will, by default, show the code and output
* lots of other options exist!

The main idea is Rmd files allow you to combine code, text, graphs, etc into 
multiple outputs that you can share (including with coding illiterate colleagues
who just want output).